---
title: "Forest Point Cloud Alignment – Stage 6"
subtitle: "Compose the solved transforms and export the aligned point cloud"
author: "Point Cloud Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: flatly
    highlight-style: github
    fig-width: 9
    fig-height: 6
    self-contained: true
execute:
  warning: false
  message: false
  eval: false
params:
  make_plots: true
---


# Overview – Stage 6 of 6

Every earlier stage solved a transform and wrote it down. **Stage 6 is the only
stage that applies the whole thing to the point cloud**, in a single pass, and
writes the result to disk.

The composed transform

$$p' = T_z\big(R_{\text{align+xyz}}\, p + t_{\text{align+xyz}}\big) + t$$

where

- $R_{\text{align+xyz}},\ t_{\text{align+xyz}}$ – the **composed** raw -> final
  map recorded at the end of Stage 2 (`R_align_xyz_1` / `t_align_xyz_1`): the
  Stage 1A rough vertical alignment followed by the fine 3D rotation `R_xyz`,
  both about `pivot1`,
- $T_z$ – the Z-only rotation solved in Stage 4, about `pivot1` in XY, rebuilt
  here from the stored `z_rotation_deg` using the identical sign convention
  Stage 5 used,
- $t$ – the XYZ translation: `t_trans_xy` from Stage 5 in X and Y,
  `t_trans_z_applied` from Stage 4 in Z.

Why the ROUGH alignment has to be in there

`las_1` sits on disk in the original scan frame. Stage 1A verticalised it **in
memory only** – there is no `writeLAS` anywhere in Stage 1 – and every
measurement made after that point, `pivot1` and `R_xyz` included, lives in the
aligned frame. So `R_xyz` on its own is a rotation solved for a cloud this stage
never reads. Stage 2 already composed the two legs into `R_align_xyz_1`, and
that is what Step 6.2 applies.

One pivot, so no rotation produces a shift

Stage 1A fixes `pivot1` **before** it rotates and does not recompute it
afterwards, so the rough alignment and the fine `R_xyz` turn about the same
point and their composition is a pure rotation about it. Stage 4 then shifted
the frame vertically and moved the pivot to `pivot2`, and Stage 5 rotated about
that – but `pivot2` differs from `pivot1` only in Z, and a rotation about the
vertical axis is unaffected by a vertical offset of its pivot. `pivot2` sits in
the same place relative to the cloud after the Z-shift as `pivot1` did before
it, so the Z-only leg can be built about `pivot1` and reproduce the staged
result exactly.

The consequence, and the reason it is worth stating: **there is one set of shift
figures, `t_exp`, and nothing else in this stage is a shift.** The translation
columns inside `M_export` are pivot-recentring terms – in UTM coordinates they
run to hundreds of metres for a fraction of a degree – and quoting them as a
"net translation" would be reporting the pivot's coordinates back as though they
were a measurement. Steps 6.2 and the summary each assert that re-expressing the
transform about `pivot1` returns `t_exp` unchanged; if either ever fails, some
leg has acquired a pivot of its own.

Diagnostic toggles

Each transform component can be switched off independently
(`apply_xyz_rotation`, `apply_z_rotation`, `apply_xy_translation`,
`apply_z_translation`). Exporting with a single component enabled is the
quickest way to attribute a residual misalignment to a particular stage.

Export completeness

The export is deliberately **not** restricted to paired or kept trees. Every
point of the source cloud is written, including points whose `TreeID` is `0` or
`NA` (ground, understorey, unsegmented returns). The alignment was *derived*
from a subset of stems; it is *applied* to everything.

# Stage 0 - Setup


## 0.1 Load libraries
```{r}
#| label: setup
#| code-summary: "Load libraries and set plot theme"

library(lidR)       # LAS I/O and point cloud processing
library(ggplot2)    # Plotting
library(dplyr)      # Data wrangling
library(plotly)     # Interactive 3D overlay figure
library(base64enc)  # Embed rendered plots into HTML galleries
library(htmltools)  # Assemble the HTML gallery output

# ── Driver-supplied parameter overrides ──────────────────────────────────────
# Every params-* chunk in this file ends with .par_apply("<its label>"). Run the
# file by hand and that call does nothing: the values the chunk just assigned
# stand. Run it under the batch driver, which puts a .batch_params list in the
# environment before the first chunk, and any entry keyed to that chunk label is
# assigned here - AFTER the defaults, so the driver's value is the one that
# survives into the functional chunk that follows.
#
# Why this exists. Parameters used to sit in the same chunk as the code that
# consumed them, so the driver's overrides - which land after a chunk finishes -
# arrived too late to matter, and every pair silently used the .qmd's hardcoded
# value. Splitting each chunk in two fixes the ordering; this helper is what the
# driver reaches through.
#
# The defaults stay in this file rather than moving into the driver. They are
# the single definition, the driver names only what a batch changes, and there
# is no second copy of ~670 parameters to drift out of step with this one.
#
# envir = parent.frame() so a call at chunk top level lands in the environment
# the chunk is being evaluated in - globalenv by hand, the pair environment
# under the driver - without needing .stage_env(), which is defined further down.
.par_apply <- function(label, envir = parent.frame()) {
  bp <- mget(".batch_params", envir = envir, ifnotfound = list(NULL),
             inherits = TRUE)[[1]]
  if (is.null(bp) || is.null(bp[[label]])) return(invisible(character(0)))
  vals <- bp[[label]]
  if (!length(vals)) return(invisible(character(0)))
  if (is.null(names(vals)) || any(!nzchar(names(vals))))
    stop(sprintf("[.par_apply] .batch_params[['%s']] must be a fully NAMED list.", label))
  # A name the chunk did not define is almost always a typo or a rename in the
  # .qmd that the driver has not caught up with. Silence there would mean the
  # batch runs on the default while looking like it was overridden, so it stops.
  unknown <- names(vals)[!vapply(names(vals), exists, logical(1),
                                 envir = envir, inherits = FALSE)]
  if (length(unknown))
    stop(sprintf(paste0("[.par_apply] %s: %d override(s) name nothing this chunk assigns:\n  %s\n",
                        "  Fix the name in the driver's .build_params(), or remove it."),
                 label, length(unknown), paste(unknown, collapse = "\n  ")))
  for (nm in names(vals)) assign(nm, vals[[nm]], envir = envir)
  cat(sprintf("  [params] %s <- driver (%s)\n", label, paste(names(vals), collapse = ", ")))
  invisible(names(vals))
}
```

## 0.2 Global Helpers
```{r}
#| label: params-helpers
#| code-summary: "Parameters - Colours, ggplot theme, stage save/load utilities, and the rotation/shift description helpers"

# ============================================================
# Colours
# ============================================================
COL_1     <- "#2196F3"
COL_2     <- "#FF5722"
COL_MATCH <- "#D9F9A5"

.par_apply("params-helpers")
```

```{r}
#| label: helpers
#| code-summary: "Colours, ggplot theme, stage save/load utilities, and the rotation/shift description helpers"

# ============================================================
# ggplot theme
# ============================================================
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      legend.position  = "bottom"
    )
)

# ============================================================
# Stage save/load utilities – shared by every stage file.
#
# Contract (identical in all six stage files):
# SAVE → hard ERROR if any *required* object is missing from the
# environment, so a checkpoint can never be written half-empty.
# A *warning* is raised for any missing *optional* object
# (diagnostic products that may legitimately not have been run);
# those are simply omitted from the file.
# LOAD → never errors on content. A *warning* is raised for each
# requested object that is absent from the RDS. Objects present in
# the RDS but not requested are COUNTED in the summary line and
# never listed. Everything found in the file is loaded either way,
# so nothing is silently discarded.
# PATHS → a save never overwrites: if the target path exists, a numeric
# suffix (_001, _002, ...) is appended until the name is free.
# ============================================================

#' Return `path` if free, otherwise path_001 / path_002 / ... (first free one).
.unique_path <- function(path) {
  if (!file.exists(path)) return(path)
  base_noext <- tools::file_path_sans_ext(path)
  ext        <- tools::file_ext(path)
  counter    <- 1L
  repeat {
    candidate <- sprintf("%s_%03d.%s", base_noext, counter, ext)
    if (!file.exists(candidate)) {
      cat(sprintf("Target already exists - using unique name: %s\n", candidate))
      return(candidate)
    }
    counter <- counter + 1L
  }
}

#' Which environment do the stage helpers read and write?
#'
#' A .qmd rendered or run by hand evaluates its chunks in the global
#' environment. The batch driver does not: it gives every LAS pair a fresh
#' environment of its own, so that a leftover object from the previous pair
#' cannot be picked up by one of the exists() guards this pipeline is full of.
#' Under the driver, everything the stages assign therefore lives THERE, and
#' not in globalenv.
#'
#' parent.env(environment()) is the environment this chunk was evaluated in:
#' globalenv() by hand, the pair environment under the driver. It is resolved
#' from where .stage_env was DEFINED rather than from where it is called, so it
#' returns the same environment inside a function as it does at chunk top
#' level, and nothing has to be told which mode it is in.
#'
#' This replaced a hardcoded .stage_env() throughout. Under the driver that
#' pointed at an environment the pipeline had never written to, so every object
#' in a save manifest read as missing and the checkpoint refused to write - the
#' Stage 1 failure that reported 30 missing required objects after a 64-minute
#' fit. (`run_counts` was the one it did find, and only because .tally()'s `<<-`
#' walked past the pair environment into globalenv. That is fixed below too.)
.stage_env <- function() parent.env(environment())

#' Is a batch driver running this file?
#'
#' The driver assigns .batch_mode <- TRUE into the pair environment before the
#' first chunk runs. It changes one thing, and only for the checkpoints: in a
#' batch, every stage runs in ONE session, so Stage N+1 already holds in memory
#' everything Stage N computed and never has to read it back off disk. The
#' checkpoints stop being load-bearing and become a record of the run.
#'
#' So under a batch: nothing in a manifest is required (a missing object is
#' reported and left out, and the pair carries on), a failed write is reported
#' rather than thrown, and .stage_load is a no-op. Run the same .qmd by hand and
#' every one of those is an error again, which is what it should be when the
#' checkpoint is the only thing carrying the stage forward.
.in_batch <- function() {
  isTRUE(mget(".batch_mode", envir = .stage_env(),
              ifnotfound = list(FALSE), inherits = FALSE)[[1]])
}

#' Read one of the driver's batch settings out of the pair environment.
#' Absent - i.e. this file is being run by hand - gives `default`.
.batch_opt <- function(name, default = NULL) {
  v <- mget(name, envir = .stage_env(), ifnotfound = list(NULL), inherits = FALSE)[[1]]
  if (is.null(v)) default else v
}

#' Where a checkpoint goes when a batch is running.
#'
#' Every save chunk in this pipeline assigns its .rds path and then USES it a
#' few lines later, in the same chunk. The driver applies its overrides AFTER a
#' chunk has run, so an override of stage1_rds_path arrives after the file has
#' already been written to the path hardcoded here - which means, without this,
#' every sample in a batch writes over the same hardcoded name.
#'
#' So in batch mode the destination is derived instead, from two values the
#' driver puts in the environment before anything runs: .batch_rds_dir and
#' .batch_sample. The label the save chunk already passes supplies the rest, so
#' "Stage 3.2" becomes <dir>/<sample>_Stage3_2.rds with nothing to keep in sync.
#'
#' NULL means: not in a batch, or the driver did not set the directory - use the
#' path the chunk asked for.
.batch_rds_path <- function(label) {
  dir <- .batch_opt(".batch_rds_dir")
  smp <- .batch_opt(".batch_sample")
  if (is.null(dir) || is.null(smp)) return(NULL)
  file.path(dir, sprintf("%s_%s.rds", smp,
                         gsub("\\.", "_", gsub("\\s+", "", label))))
}

#' Save a stage checkpoint. Errors on missing required objects - unless the
#' batch driver is running this file, in which case nothing is required, the
#' destination is the per-sample one derived above, and a failure to write is
#' reported rather than thrown. See .in_batch() above.
`%||%` <- function(a, b) if (is.null(a)) b else a

.stage_save <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  required <- unique(required)
  optional <- setdiff(unique(optional), required)

  # A batch may switch an individual checkpoint off by label - the driver's
  # keep_stage3_fit_checkpoint does exactly this for "Stage 3.1", whose chunk
  # must still RUN (the final Stage 3 manifest inherits from it) but need not
  # write a second large file per sample.
  if (!strict && label %in% .batch_opt(".batch_skip_saves", character(0))) {
    cat(sprintf("[%s save] Skipped - switched off for this batch.\n", label))
    return(invisible(NA_character_))
  }

  # inherits = FALSE, matching the mget() below, which has always been strict.
  # With inheritance on, an object could be FOUND in a parent environment and
  # then fail to be FETCHED, and under the driver the parent is globalenv -
  # i.e. the previous pair's leftovers.
  .have <- function(x)
    if (!length(x)) logical(0)
    else vapply(x, exists, logical(1), envir = envir, inherits = FALSE)

  missing_req <- required[!.have(required)]
  if (length(missing_req) > 0L) {
    if (strict)
      stop(sprintf(
        "[%s save] Cannot save - %d required object(s) missing from the environment:\n  %s",
        label, length(missing_req), paste(missing_req, collapse = "\n  ")))
    cat(sprintf(
      "[%s save] %d required object(s) missing - treated as optional and omitted (batch mode):\n  %s\n",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))
  }

  missing_opt <- optional[!.have(optional)]
  if (length(missing_opt) > 0L) {
    msg <- sprintf(
      "[%s save] %d optional object(s) missing - omitted from the checkpoint:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  "))
    if (strict) warning(msg) else cat(msg, "\n", sep = "")
  }

  present  <- c(setdiff(required, missing_req), setdiff(optional, missing_opt))
  n_opt_in <- length(setdiff(optional, missing_opt))
  dat      <- mget(present, envir = envir, inherits = FALSE)

  # ── The checkpoint describes itself ────────────────────────────────────────
  # The manifest travels INSIDE the file. Without it, loading a checkpoint means
  # first reconstructing the exact `required`/`optional` vectors it was written
  # with - and those are built by chunks, so a later checkpoint could only be
  # loaded after running the chunks that define an earlier one's manifest. That
  # made resuming from an arbitrary point in a fresh session impossible: the
  # Stage 3.2 chunk, for instance, stops outright unless .stage3_1_required is
  # already in scope.
  #
  # Stored under a name no manifest can collide with, and stripped again on
  # load, so it never reaches the session as an object.
  dat[[".__manifest__"]] <- list(
    required = setdiff(required, missing_req),
    optional = setdiff(optional, missing_opt),
    label    = label,
    saved_utc = format(Sys.time(), tz = "UTC", usetz = TRUE))

  .write <- function() {
    p <- path
    if (!strict) {
      bp <- .batch_rds_path(label)
      if (!is.null(bp)) p <- bp
    }
    dir.create(dirname(p), recursive = TRUE, showWarnings = FALSE)
    # .unique_path only interactively. A batch path already carries the sample
    # name, so it cannot collide with another sample, and a re-run of the same
    # sample should REPLACE its checkpoint rather than leave _001, _002 behind
    # for the next reader to guess between.
    if (strict) p <- .unique_path(p)
    saveRDS(dat, p)
    cat(sprintf("[%s save] Saved %d object(s) (%d required, %d optional) -> %s\n",
                label, length(dat), length(present) - n_opt_in, n_opt_in, p))
    p
  }

  if (strict) return(invisible(.write()))

  out <- tryCatch(.write(), error = function(e) {
    cat(sprintf("[%s save] Checkpoint NOT written: %s\n", label, conditionMessage(e)))
    cat(sprintf("[%s save] Batch mode - the objects are still in this session, so the run continues.\n",
                label))
    NA_character_
  })
  invisible(out)
}

#' Load a stage checkpoint. Warns (never errors) on content mismatches, and
#' under the batch driver does not read at all - see .in_batch() above.
.stage_load <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  # In a batch the objects this would read are already live in `envir`, put
  # there by the stage that just ran. Reading them back would at best be a
  # no-op. At worst it would be wrong: checkpoint paths are per-sample and
  # .unique_path() means a re-run writes _001 rather than replacing the
  # original, so the path named here can easily be an EARLIER run's file. And
  # if the write failed or was partial, there would be nothing to read and the
  # pair would die at the start of the next stage having done all the work.
  if (!strict) {
    cat(sprintf("[%s load] Skipped - batch mode: this stage's objects are already in the session.\n",
                label))
    return(invisible(character(0)))
  }

  if (!file.exists(path))
    stop(sprintf("[%s load] File not found:\n  %s", label, path))

  dat <- readRDS(path)

  # ── Prefer the manifest the file carries ───────────────────────────────────
  # A checkpoint written by .stage_save records what it holds, so a load needs
  # nothing from the session - no earlier chunk, no earlier checkpoint, no
  # reconstructed name vectors. `required` and `optional` become a CHECK against
  # what the caller expected rather than the definition of what to read, which
  # is what lets any stage be resumed from any checkpoint in a fresh session.
  #
  # Files written before manifests existed simply have none; those fall back to
  # the caller's vectors, exactly as before.
  mf <- dat[[".__manifest__"]]
  dat[[".__manifest__"]] <- NULL
  caller_req <- unique(required)
  caller_opt <- setdiff(unique(optional), caller_req)

  if (!is.null(mf)) {
    required <- unique(mf$required)
    optional <- setdiff(unique(mf$optional), required)
    src <- sprintf("its own manifest (written %s)", mf$saved_utc %||% "?")
  } else {
    required <- caller_req; optional <- caller_opt
    src <- "the caller's manifest - this file predates self-describing checkpoints"
  }
  expected <- c(required, optional)

  cat(sprintf("[%s load] Reading %d object(s) from: %s\n",
              label, length(dat), path))
  cat(sprintf("[%s load] Manifest: %s\n", label, src))

  # Named by the caller but absent from the file: the real incompatibility, and
  # the only one worth a warning when the file describes itself.
  if (!is.null(mf)) {
    unmet <- setdiff(caller_req, names(dat))
    if (length(unmet))
      warning(sprintf(
        "[%s load] %d object(s) this stage expects are not in the checkpoint:\n  %s",
        label, length(unmet), paste(unmet, collapse = "\n  ")))
  }

  missing_req <- required[!required %in% names(dat)]
  if (length(missing_req) > 0L)
    warning(sprintf(
      "[%s load] %d REQUIRED object(s) absent from the RDS - downstream chunks that use them will fail:\n  %s",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))

  missing_opt <- optional[!optional %in% names(dat)]
  if (length(missing_opt) > 0L)
    warning(sprintf(
      "[%s load] %d optional object(s) absent from the RDS:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  ")))

  # Unrequested objects are COUNTED, never listed. Every checkpoint from Stage 2
  # onward is cumulative, so this set is routinely hundreds of names long and
  # printing it buried the two messages that actually matter - an absent
  # required object, and an absent optional one - under a wall of text. The
  # count still appears in the summary line below, which is the only part of it
  # anyone reads.
  extra <- setdiff(names(dat), expected)

  # ── Publish what was loaded, for the next stage's save manifest ────────────
  # Every stage's SAVE manifest is built as "the previous stage's manifest, plus
  # my own objects", and the previous manifest is a vector defined by that
  # stage's LOAD chunk. Load a LATER checkpoint directly and that chunk never
  # runs, so the save dies on a missing .stageN_required - which is exactly the
  # resume-from-anywhere case these checkpoints now exist to support.
  #
  # Recording the manifest here breaks the dependency: .manifest_from() below
  # falls back to whatever was actually loaded, whichever chunk did it.
  assign(".stage_last_manifest",
         list(required = required, optional = optional,
              names = names(dat), label = label, path = path),
         envir = envir)

  list2env(dat, envir = envir)

  cat(sprintf("[%s load] Loaded %d object(s) | required present: %d/%d | optional present: %d/%d | unrequested: %d\n",
              label, length(dat),
              length(required) - length(missing_req), length(required),
              length(optional) - length(missing_opt), length(optional),
              length(extra)))
  invisible(names(dat))
}

#' The manifest a save chunk inherits from the stage before it.
#'
#' Looks for the named vector first - that is the normal path, when this
#' session's own load chunk built it. Falls back to the manifest carried by
#' whichever checkpoint was last loaded, which is what makes it possible to load
#' any checkpoint and still save the next one.
#'
#' `which` is "required" or "optional". Stops with something actionable rather
#' than returning character(0): a save that silently inherits nothing writes a
#' checkpoint missing every object from every earlier stage, and nothing would
#' notice until a later stage failed to find them.
.manifest_from <- function(name, which = c("required", "optional"),
                           envir = .stage_env()) {
  which <- match.arg(which)
  if (exists(name, envir = envir, inherits = FALSE))
    return(get(name, envir = envir, inherits = FALSE))

  lm <- if (exists(".stage_last_manifest", envir = envir, inherits = FALSE))
          get(".stage_last_manifest", envir = envir, inherits = FALSE) else NULL

  if (!is.null(lm)) {
    cat(sprintf("[manifest] %s not in the session - inheriting from the %s checkpoint that was loaded.\n",
                name, lm$label))
    # For "required", everything the loaded file actually held is the honest
    # inheritance: it is what this session has, regardless of how the file that
    # produced it split required from optional.
    return(if (identical(which, "required")) lm$names else character(0))
  }

  # ── Last resort: the session itself ────────────────────────────────────────
  # No named vector and no checkpoint loaded. That is the ordinary case for a
  # session built by RUNNING chunks rather than by loading - Stage 1 through
  # Stage 3 straight down the document, or an interactive session where the
  # objects were produced rather than read. There is nothing wrong with it, and
  # a save must not refuse.
  #
  # So fall back to what Stages 4 and 5 have always done: take the manifest from
  # the environment. Every non-function, non-dotted object in scope is what this
  # stage inherited, by definition - it is the same rule, just applied later.
  #
  # Only "required" gets this. Optional means "save it if it happens to exist",
  # which is meaningless once the list IS everything that exists.
  if (identical(which, "required")) {
    nm <- setdiff(ls(envir = envir), lsf.str(envir = envir))
    nm <- nm[!startsWith(nm, ".")]
    if (length(nm)) {
      cat(sprintf(paste0("[manifest] %s not found and no checkpoint was loaded - taking the manifest\n",
                         "           from the session instead (%d object(s)). This is the normal path\n",
                         "           when the stage was RUN rather than loaded.\n"), name, length(nm)))
      return(nm)
    }
  } else {
    return(character(0))
  }

  stop(sprintf(paste0("[manifest] %s not found, no checkpoint has been loaded, and the session is\n",
                      "  empty - there is nothing to save. Run this stage's chunks, or load a\n",
                      "  checkpoint, before the save chunk."), name))
}

#' Print the Stage 1 pivot point.
.report_pivot1 <- function() {
  if (exists("pivot1")) {
    cat(sprintf("pivot1 (Stage 1 footprint centre): (%.4f, %.4f, %.4f)\n",
                pivot1[1], pivot1[2], pivot1[3]))
  } else {
    cat("pivot1 NOT available.\n")
  }
}

# ============================================================
# Rotation description: one factorisation, used everywhere
#
# A 3x3 rotation is not readable. Every summary and every error term in this
# stage is quoted as four numbers instead:
#
#   XY rotation magnitude - tilt away from vertical, degrees
#   XY rotation direction - azimuth the +Z AXIS tips toward, CCW from +X, [0,360)
#   Z rotation            - yaw about vertical, CCW seen from above, degrees
#   full XYZ rotation     - the total turn about the matrix's own axis
#
# The first three are an EXACT factorisation, not a description:
#     R = tilt(mag, dir) %*% yaw(z)
# so the three numbers rebuild R to machine precision and .decompose_rotation()
# returns that residual as rebuild_err. That distinction matters for an error
# report - a set of angles that merely characterises a residual rotation is not
# something anyone can act on or re-apply.
#
# The tilt builder is byte-identical to Stage 3's .tilt_rotation_matrix and to
# the one in the test-sample generator. All three must agree: if this stage
# described a rotation in a different convention from the one that generated the
# truth values, the errors below would be measuring the gap between two
# conventions rather than the pipeline's accuracy. Edit one, edit all three.
#
# Convention, verified against the matrix rather than asserted:
#     tilt(m, d) %*% c(0, 0, 1) = (sin m cos d, sin m sin d, cos m)
# so m is exactly the tilt and d is exactly the azimuth the +Z axis leans toward.
# (Points on the ground at azimuth d move DOWN, not up - the axis tips that way.)
# ============================================================
.tilt_rotation_matrix <- function(magnitude_deg, direction_deg) {
  m <- magnitude_deg * pi / 180
  d <- direction_deg * pi / 180
  axis <- c(-sin(d), cos(d), 0)          # horizontal axis, perpendicular to direction
  ax <- axis[1]; ay <- axis[2]; az <- axis[3]
  K <- matrix(c(  0, -az,  ay,
                 az,   0, -ax,
                -ay,  ax,   0), nrow = 3, byrow = TRUE)
  diag(3) + sin(m) * K + (1 - cos(m)) * (K %*% K)
}

.yaw_rotation_matrix <- function(deg) {
  a <- deg * pi / 180
  R <- diag(3)
  R[1:2, 1:2] <- matrix(c(cos(a), -sin(a),
                          sin(a),  cos(a)), nrow = 2, byrow = TRUE)
  R
}

.compose_rotation <- function(xy_mag_deg, xy_dir_deg, z_deg) {
  .tilt_rotation_matrix(xy_mag_deg, xy_dir_deg) %*% .yaw_rotation_matrix(z_deg)
}

#' Exact inverse of .compose_rotation. Returns tilt magnitude, tilt direction,
#' yaw and total angle, plus the residual of rebuilding R from the three.
#' At zero tilt the direction is undefined and returned as NA rather than as the
#' 0 that atan2(0, 0) happens to give - "no tilt" and "tilt pointing due east"
#' are different claims and only one of them is true.
.decompose_rotation <- function(R, dir_tol_deg = 1e-9) {
  m <- acos(max(-1, min(1, R[3, 3]))) * 180 / pi
  d <- if (m < dir_tol_deg) NA_real_ else (atan2(R[2, 3], R[1, 3]) * 180 / pi) %% 360
  M <- t(.tilt_rotation_matrix(m, if (is.na(d)) 0 else d)) %*% R
  a <- atan2(M[2, 1], M[1, 1]) * 180 / pi
  list(xy_mag_deg  = m,
       xy_dir_deg  = d,
       z_deg       = a,
       total_deg   = acos(max(-1, min(1, (sum(diag(R)) - 1) / 2))) * 180 / pi,
       rebuild_err = max(abs(.compose_rotation(m, if (is.na(d)) 0 else d, a) - R)))
}

#' Break a translation into the five numbers the summaries and the error report
#' both quote: the three components, the horizontal magnitude, and the full 3-D
#' magnitude.
.shift_specs <- function(t) {
  t <- as.numeric(t)
  list(x = t[1], y = t[2], z = t[3],
       xy = sqrt(sum(t[1:2]^2)), xyz = sqrt(sum(t^2)))
}

#' The translation a full affine map applies AT a chosen reference point - i.e.
#' rewriting "R x + t_col" as "rotate about P, then shift by s".
#'
#' Needed because a shift is only meaningful relative to a pivot once a rotation
#' is involved. The raw translation column of a world-coordinate 4x4 is dominated
#' by the pivot-recentring terms (these are UTM coordinates, so it runs to
#' hundreds of metres for a fraction of a degree) and comparing two of those
#' tells you nothing. Referred to a pivot, the same matrix reports the metres of
#' shift a reader would recognise.
.shift_about <- function(M, P) {
  P <- as.numeric(P)
  as.numeric(M[1:3, 4] - P + M[1:3, 1:3] %*% P)
}
```

## 0.3 Load the Stage 5 Checkpoint

Stage 6 asks for only the handful of objects it actually needs: the file paths,
the pivots, and the solved transform
components. Everything else in the checkpoint is loaded too (nothing is
discarded) but is only counted, not listed.

```{r}
#| label: params-load_Stage6
#| code-summary: "Parameters - Load the Stage 5 checkpoint and report the pivots and the solved transform components"


stage5_rds_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/RunR/exports/SaveLoad/I_SNBG_F1_ai_134p77_2020_2024_TLS_TLS_Stage5.rds"

.par_apply("params-load_Stage6")
```

```{r}
#| label: load_Stage6
#| code-summary: "Load the Stage 5 checkpoint and report the pivots and the solved transform components"
#| code-fold: false

# Stage 6 uses only a handful of Stage 5 objects. .stage_load pulls EVERYTHING in
# the RDS via list2env regardless of this list (the manifest only drives the
# present/absent warnings), so Stage 6 requests just what the compose-and-apply
# needs and lets the loader bring the rest in silently; the loader counts the
# unrequested objects rather than listing them.
.stage6_required <- c(
  # file paths (to re-load the cloud)
  "las_1_path", "las_2_path",
  # rotation origin(s)
  "pivot1", "pivot2",
  # Stage 2: the COMPOSED raw->final rotation (Stage 1A rough align + R_xyz).
  # This, not R_xyz alone, is what a raw on-disk las_1 has to be put through -
  # see the note in Step 6.2.
  "R_align_xyz_1", "t_align_xyz_1", "angle_align_xyz_1",
  # the two components, kept so the composition can be rebuilt or audited
  "R_xyz", "angle_xyz", "R_align_1", "align_pivot_1",
  # Stage 3/4 Z translation
  "t_trans_z_applied",
  # Stage 4/5 Z-only rotation (applied as +z_rotation_deg, matching Stage 5's 5.1)
  "z_rotation_deg", "z_rotation_rad",
  # Stage 5 XY translation - the residual horizontal offset
  "t_trans_xy",
  # final pair set that produced the transforms (reporting)
  "pairs_stage5"
)
.stage6_optional <- c(
  # reporting-only chain + context; absent on some builds
  "pairs_stage1", "pairs_stage2", "pairs_ax_fitted", "pairs_stage4",
  "R_rot_xyz", "angle_rot_xyz", "t_trans_z"
)

.stage_load(stage5_rds_path, .stage6_required, .stage6_optional, label = "Stage 5")

cat("\n══ Settings carried in from the earlier stages ════════════════════════════\n")
if (exists(".report_pivot1")) .report_pivot1()       # <- the rotation origin for the composed transform
cat(sprintf("pivot2 (post Z-shift):            (%.4f, %.4f, %.4f)\n",
            pivot2[1], pivot2[2], pivot2[3]))
cat("\n── Solved transform components ───────────────────────────────────────────\n")
cat(sprintf("  rough align     (Stage 1A): %.4f deg\n",
            acos(pmax(-1, pmin(1, (sum(diag(R_align_1)) - 1) / 2))) * 180 / pi))
cat(sprintf("  XYZ rotation    (Stage 2): %.4f deg\n", angle_xyz))
cat(sprintf("  composed rough+XYZ        : %.4f deg  <- this is what Step 6.2 applies\n",
            angle_align_xyz_1))
cat(sprintf("  Z translation   (Stage 4): %+.4f m (%+.2f cm)\n",
            t_trans_z_applied, t_trans_z_applied * 100))
cat(sprintf("  Z-only rotation (Stage 5): %+.4f deg\n", z_rotation_deg))
cat(sprintf("  XY translation  (Stage 5): dx = %+.5f m, dy = %+.5f m\n",
            t_trans_xy[1], t_trans_xy[2]))
cat(sprintf("\nPairs that produced these values: %d%s\n",
            nrow(pairs_stage5),
            if (exists("pairs_stage1")) sprintf(" (of %d originally matched)", nrow(pairs_stage1)) else ""))
```



## 0.4 Load Point Clouds

```{r}
#| label: params-load-stage6
#| code-summary: "Parameters - Read las_1 (the cloud to transform) and report point + TreeID counts"


# Only las_1 is needed here – las_2 is the fixed reference epoch and is never
# transformed, so there is nothing to export for it. It is loaded anyway (behind
# a toggle) purely so the optional overlay figure at the end can be drawn.
load_reference_cloud <- FALSE

.par_apply("params-load-stage6")
```

```{r}
#| label: load-stage6
#| code-summary: "Read las_1 (the cloud to transform) and report point + TreeID counts"

las_1 <- readLAS(las_1_path)

las_1 <- lidR::add_lasattribute(
  las_1,
  x    = seq_len(nrow(las_1@data)),
  name = "point_id",
  desc = "Unique ID for las_1 points"
)

if (load_reference_cloud) {
  las_2 <- readLAS(las_2_path)
}

# Keep an untouched copy of the raw cloud. The default export writes every
# original point of las_1, regardless of segmentation.
las_1_raw <- las_1

.tid_1 <- las_1@data$TreeID
cat(sprintf("las_1: %s points | %d unique TreeID values (NA and 0 not counted)\n",
            format(nrow(las_1@data), big.mark = ","),
            length(unique(.tid_1[!is.na(.tid_1) & .tid_1 != 0L]))))
cat(sprintf("  points with TreeID = 0 : %s\n",
            format(sum(.tid_1 == 0L, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  points with TreeID = NA: %s\n",
            format(sum(is.na(.tid_1)), big.mark = ",")))
cat("  All of these are included in the export.\n")

if (load_reference_cloud) {
  cat(sprintf("las_2 (reference, never transformed): %s points\n",
              format(nrow(las_2@data), big.mark = ",")))
}
```

# Stage 6 - Compose, Apply Export

## 6.1 Select the Transformations to Apply

```{r}
#| label: params-select-transformations
#| code-summary: "Parameters - Which solved components to apply, and which source cloud to export"


# ── Transformation toggles ───────────────────────────────────────────────────
# Each component can be disabled independently as a diagnostic. Exporting with
# exactly one enabled isolates that component's effect; exporting with all
# disabled writes an untransformed copy, which is a useful null case for
# checking the export path itself.
apply_xyz_rotation   <- TRUE    # Stage 1A rough align + Stage 2 R_xyz, composed

apply_z_rotation     <- TRUE    # Stage 4/5 Z-only yaw, about pivot1 (XY)
apply_xy_translation <- TRUE    # Stage 5 t_trans_xy
apply_z_translation  <- TRUE    # Stage 4 t_trans_z_applied

# ── Which cloud to export ──────────────────────────────────────────
# "raw"      -> the point cloud exactly as read from las_1_path, every point
#               retained. This is the default and, in the current pipeline, the
#               only meaningful setting: no stage filters the cloud any more, so
#               las_1 and las_1_raw hold the same points.
# "filtered" -> the in-memory las_1. Kept as a hook for a run that deliberately
#               subsets the cloud before this stage; identical to "raw" otherwise.
#
# Declared HERE rather than in the export step: Step 6.4 reads it when it picks
# the source cloud, which runs BEFORE the export chunk that used to define it.
export_source <- "raw"          # "raw" or "filtered"

.par_apply("params-select-transformations")
```

```{r}
#| label: select-transformations
#| code-summary: "Which solved components to apply, and which source cloud to export"
#| code-fold: false


```

## 6.2 Compose the Rotation

`R_xyz` alone is **not** the rotation that carries a raw on-disk `las_1` into the
final frame. Stage 1A verticalised `las_1` in memory (`R_align_1`) and never
wrote it back, and `R_xyz` was then solved *in that aligned frame*. Stage 2
composed the two into `R_align_xyz_1` / `t_align_xyz_1`, and that pair - not
`R_xyz` - is the raw -> final map. Applying `R_xyz` on its own here would rotate
the disk cloud by a matrix solved for a cloud it is not, leaving the export in
neither frame.

Both legs turn about the **same** point: Stage 1A fixes `pivot1` before it
rotates and does not recompute it, so `align_pivot_1` and `pivot1` are the same
coordinates and the composition is a pure rotation about that point. The
translation `t_align_xyz_1` carries is the pivot-recentring term and nothing
else - no rotation here produces a shift.

Everything is still assembled as 4x4 homogeneous matrices and multiplied once.
Not because the pivots differ, but because one multiplication is harder to get
wrong than three sequential recentrings, and because the finished matrix is what
gets exported for CloudCompare anyway.

```{r}
#| label: compose-rotation
#| code-summary: "Compose the Z-only rotation on top of the composed rough + fine XYZ rotation, honouring the toggles"

# ── Leg 1: the composed rough vertical alignment + fine XYZ rotation ─────────
# R_align_xyz_1 / t_align_xyz_1 come from the Stage 2 composition step:
#     x_final = R_align_xyz_1 %*% x_raw + t_align_xyz_1
# which reproduces, in one application to the RAW cloud,
#     x_aligned = R_align_1 (x_raw     - align_pivot_1) + align_pivot_1
#     x_final   = R_xyz     (x_aligned - pivot1)        + pivot1
# with align_pivot_1 == pivot1, so t_align_xyz_1 is purely (p - R p) and this is
# one rotation about one point, not a rotation plus a shift.
#
# If an older checkpoint predates the composition, it is rebuilt here from the
# components rather than silently falling back to R_xyz alone - the fallback IS
# the bug this step exists to fix.
if (exists("R_align_xyz_1") && exists("t_align_xyz_1")) {
  R_fine <- R_align_xyz_1
  t_fine <- as.numeric(t_align_xyz_1)
} else if (exists("R_align_1") && exists("align_pivot_1")) {
  .pa    <- as.numeric(align_pivot_1)
  .p1    <- as.numeric(pivot1)
  if (sqrt(sum((.pa - .p1)^2)) > 1e-3)
    stop(sprintf(paste0("[6.2] align_pivot_1 and pivot1 differ by %.4f m. Both rotation ",
                        "legs must turn about the same point; re-run Stage 1A (it must not ",
                        "recompute pivot1 after rotating) and Stage 2."),
                 sqrt(sum((.pa - .p1)^2))))
  R_fine <- R_xyz %*% R_align_1
  t_fine <- as.numeric(.p1 - R_fine %*% .p1)
  cat("[6.2] R_align_xyz_1 absent - rebuilt the composition from R_align_1 + R_xyz.\n")
  rm(.pa, .p1)
} else {
  stop("[6.2] Neither R_align_xyz_1 nor R_align_1 / align_pivot_1 is available. ",
       "This checkpoint predates the rough alignment; re-run Stages 1 and 2 and reload. ",
       "Applying R_xyz on its own would skip the Stage 1A alignment entirely.")
}

M_fine <- diag(4)
if (apply_xyz_rotation) {
  M_fine[1:3, 1:3] <- R_fine
  M_fine[1:3, 4]   <- t_fine
}

# ── Leg 2: the Z-only yaw, with the SAME sign Stage 5 applied ────────────────
# ang_diff = angle_2 - angle_1 is the rotation carrying las_1 TO las_2, so
# aligning las_1 onto las_2 applies +z_rotation_deg. Stage 5's 5.1 rotates the
# centres by exactly this (a = +z_rotation_deg), and Stage 6 MUST rotate the
# exported cloud the same way - otherwise the cloud and the measured t_trans_xy
# would live in opposite-yaw frames and the export would be mis-rotated by ~2x
# the angle.
#
# Build the 2x2 block with the EXACT same layout Stage 5's 5.1 uses
# (nrow = 2, byrow = TRUE) so the two stages apply the identical rotation, not
# its transpose. NOTE: matrix(c(cos, -sin, sin, cos), 2, 2) with the default
# column-major fill would produce the TRANSPOSE (the opposite rotation); the
# byrow = TRUE here is load-bearing and must match Stage 5.
z_rotation_rad_exp <- z_rotation_deg * pi / 180        # +z_rotation_deg (matches Stage 5 5.1)

Rz_exp <- diag(3)
Rz_exp[1:2, 1:2] <- matrix(c(cos(z_rotation_rad_exp), -sin(z_rotation_rad_exp),
                              sin(z_rotation_rad_exp),  cos(z_rotation_rad_exp)),
                            nrow = 2, byrow = TRUE)

# The yaw acts about pivot1 in XY. Stage 5 rotated about pivot2, but a Z-only
# translation cannot move a horizontal centre, so pivot2[1:2] == pivot1[1:2] and
# the two are the same rotation.
.pz  <- as.numeric(pivot1)
M_z  <- diag(4)
if (apply_z_rotation) {
  M_z[1:3, 1:3] <- Rz_exp
  M_z[1:3, 4]   <- .pz - Rz_exp %*% .pz
}

# ── The composed rotation leg: fine first, then the yaw on top ───────────────
# Order matters: the composed rough+XYZ rotation is what the cloud was carrying
# when the Z-only rotation was solved, so it goes first.
M_rot_exp      <- M_z %*% M_fine
R_combined_exp <- M_rot_exp[1:3, 1:3]
t_combined_exp <- M_rot_exp[1:3, 4]

angle_combined <- acos(pmax(-1, pmin(1, (sum(diag(R_combined_exp)) - 1) / 2))) * 180 / pi

cat("── Composed rotation ─────────────────────────────────────────────────────\n")
cat(sprintf("  rough align + XYZ rotation: %s (%.4f deg, composed)\n",
            if (apply_xyz_rotation) "ON " else "OFF",
            acos(pmax(-1, pmin(1, (sum(diag(R_fine)) - 1) / 2))) * 180 / pi))
cat(sprintf("    of which fine XYZ only  : %.4f deg\n", angle_xyz))
cat(sprintf("  Z-only rotation           : %s (%+.4f deg)\n",
            if (apply_z_rotation) "ON " else "OFF", z_rotation_deg))
cat(sprintf("  Combined magnitude        : %.4f deg\n", angle_combined))
# The translation column here is the pivot-recentring term for a rotation about
# pivot1 - in UTM coordinates it runs to hundreds of metres for a fraction of a
# degree. It is NOT a shift and is deliberately not reported as one; the check
# below is what matters.
.rot_shift <- max(abs(.shift_about(M_rot_exp, pivot1)))
cat(sprintf("  Translation induced by the rotation: %.2e m (must be ~0)\n", .rot_shift))
if (.rot_shift > 1e-6)
  warning(sprintf(paste0("The composed rotation displaces pivot1 by %.5f m. Some leg is ",
                         "turning about a pivot the others do not share."), .rot_shift),
          call. = FALSE)
cat("\nR_combined_exp:\n")
print(round(R_combined_exp, 8))

rm(.pz, .rot_shift)
```

## 6.3 Compose the Translation

```{r}
#| label: compose-translation
#| code-summary: "Assemble the XYZ translation vector, honouring the toggles"

t_exp <- c(
  if (apply_xy_translation) t_trans_xy[1]      else 0,
  if (apply_xy_translation) t_trans_xy[2]      else 0,
  if (apply_z_translation)  t_trans_z_applied  else 0
)
names(t_exp) <- c("dx", "dy", "dz")

cat("── Composed translation ──────────────────────────────────────────────────\n")
cat(sprintf("  XY translation: %s (dx = %+.5f, dy = %+.5f m)\n",
            if (apply_xy_translation) "ON " else "OFF", t_trans_xy[1], t_trans_xy[2]))
cat(sprintf("  Z  translation: %s (dz = %+.5f m)\n",
            if (apply_z_translation) "ON " else "OFF", t_trans_z_applied))
cat(sprintf("  Applied vector: (%+.5f, %+.5f, %+.5f) m\n", t_exp[1], t_exp[2], t_exp[3]))
```

## 6.4 Apply the Transform

```{r}
#| label: apply-transform-stage6
#| code-summary: "Apply the single composed 4x4 transform – every point of the source cloud"
#| code-fold: false

las_export <- if (identical(export_source, "raw")) las_1_raw else las_1

xyz_orig_exp <- as.matrix(las_export@data[, c("X", "Y", "Z")])

# ── One matrix, applied once ─────────────────────────────────────────────────
# M_export folds every enabled component into a single homogeneous transform:
#   M_export = [translation] %*% [Z-only yaw about pivot1] %*% [rough align + XYZ]
# so no pivot is re-applied here by hand. The pivot IS still in the matrix -
# every rotation leg was built about pivot1 - which is what keeps this consistent
# with the geometry the fitting was done on. Rotating a UTM cloud about the
# coordinate origin would turn a fraction of a degree into a translation of
# hundreds of metres; none of these legs does that, and the only true
# translation in M_export is t_exp.
M_export <- diag(4)
M_export[1:3, 4] <- t_exp
M_export <- M_export %*% M_rot_exp

xyz_new_exp <- t(M_export[1:3, 1:3] %*% t(xyz_orig_exp) + M_export[1:3, 4])

cat(sprintf("Transformed %s points (source: %s cloud).\n",
            format(nrow(xyz_new_exp), big.mark = ","), export_source))
cat(sprintf("Every rotation leg is about pivot1 = (%.4f, %.4f, %.4f)\n",
            pivot1[1], pivot1[2], pivot1[3]))

# ── Net displacement summary ─────────────────────────────────────────────────
.disp <- sqrt(rowSums((xyz_new_exp - xyz_orig_exp)^2))
cat(sprintf("Net point displacement: min %.4f m | median %.4f m | max %.4f m\n",
            min(.disp), median(.disp), max(.disp)))
```

## 6.5 Export

```{r}
#| label: params-export-aligned-cloud
#| code-summary: "Parameters - Write the transformed cloud – all points, all trees, including TreeID 0/NA"




# export_source is set in Step 6.1, alongside the transformation toggles, so
# that Step 6.4 (which reads it) is not depending on a chunk further down.

# ── Output ───────────────────────────────────────────────────────────────────
output_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/Exports/BOSF1_aiiman.las"

# ── The test-sample marker box ───────────────────────────────────────────────
# A cloud from the test-sample generator carries seventeen extra points: the
# rotation pivot, eight TargetBox = 1 corners at their truth positions, and
# eight TargetBox = 2 corners that were moved by the forward transform. Step 6.6
# scores the run on how close the transformed box 2 lands to box 1. On real data
# these fields are simply absent and everything below is a no-op.
box_field_name    <- "TargetBox"
corner_field_name <- "BoxCorner"

# The box-1 corners are the TRUTH. Transforming them along with everything else
# would move the reference the run is being judged against, and the exported
# file would then show two boxes that are both wrong in the same way. With this
# TRUE they are written back where the generator put them, so the export can be
# opened on its own and read directly: box 1 = where the corners belong, box 2 =
# where this run put them, and the gap between the two is the error.
#
# It changes nothing in the arithmetic - Step 6.6 works from M_export and the
# unmoved coordinates either way, never from what was written to disk.
restore_truth_box <- TRUE

# Add a scalar field so CloudCompare labels the exported points on load.
add_source_scalar_field <- TRUE

.par_apply("params-export-aligned-cloud")
```

```{r}
#| label: export-aligned-cloud
#| code-summary: "Write the transformed cloud – all points, all trees, including TreeID 0/NA"
#| code-fold: false

# The batch names this per sample. It has to be read here rather than pushed in:
# this chunk assigns the path and writes the file, so a driver override applied
# after the chunk would arrive after the write - which is how every pair of a
# batch ended up queued behind the one hardcoded name above.
if (.in_batch())
  output_path <- .batch_opt(".batch_aligned_las", output_path)


las_out        <- las_export
las_out@data$X <- xyz_new_exp[, 1]
las_out@data$Y <- xyz_new_exp[, 2]
las_out@data$Z <- xyz_new_exp[, 3]

# ── The truth corners stay put ───────────────────────────────────────────────
# Taken from xyz_orig_exp, the snapshot Step 6.4 made BEFORE the transform, not
# from las_export@data - which shares its storage with las_out and has just been
# overwritten. Every other point, including the box-2 corners, is transformed.
.n_box1 <- 0L
if (isTRUE(restore_truth_box) && box_field_name %in% names(las_out@data)) {
  .keep <- which(las_out@data[[box_field_name]] == 1L)
  .n_box1 <- length(.keep)
  if (.n_box1) {
    las_out@data$X[.keep] <- xyz_orig_exp[.keep, 1]
    las_out@data$Y[.keep] <- xyz_orig_exp[.keep, 2]
    las_out@data$Z[.keep] <- xyz_orig_exp[.keep, 3]
  }
  rm(.keep)
}

# No TreeID-based subsetting happens anywhere in this stage: las_out holds every
# point of the source cloud, whatever its TreeID, whether or not its tree ever
# appeared in a kept pair.
.tid_out <- las_out@data$TreeID
cat("── Export contents ───────────────────────────────────────────────────────\n")
cat(sprintf("  Total points          : %s\n", format(nrow(las_out@data), big.mark = ",")))
cat(sprintf("  Distinct TreeIDs (>0) : %d\n",
            length(unique(.tid_out[!is.na(.tid_out) & .tid_out != 0L]))))
cat(sprintf("  Points with TreeID = 0: %s\n",
            format(sum(.tid_out == 0L, na.rm = TRUE), big.mark = ",")))
cat(sprintf("  Points with TreeID =NA: %s\n",
            format(sum(is.na(.tid_out)), big.mark = ",")))
cat(sprintf("  Trees in kept pairs   : %d (a subset – NOT a filter on the export)\n",
            nrow(pairs_stage5)))
if (box_field_name %in% names(las_out@data)) {
  .bx <- las_out@data[[box_field_name]]
  cat(sprintf("  Marker box            : %d truth corner(s)%s | %d transformed corner(s)\n",
              sum(.bx == 1L, na.rm = TRUE),
              if (isTRUE(restore_truth_box)) " left at truth" else " ALSO transformed",
              sum(.bx == 2L, na.rm = TRUE)))
  rm(.bx)
}

if (add_source_scalar_field) {
  # A constant extra LAS attribute; CloudCompare surfaces extra attributes as
  # named scalar fields, which makes the two epochs easy to tell apart once
  # both are loaded in the same viewer.
  las_out <- lidR::add_lasattribute(
    las_out,
    x    = rep(1L, nrow(las_out@data)),
    name = "Aligned",
    desc = "Aligned"
  )
}

# Interactively, never overwrite - a rerun parks itself at _001. In a batch the
# sample name already makes the path unique, and the driver records THIS path in
# the report, so a silent _001 would make the report point at the previous run's
# file. The pair's own file is overwritten instead.
if (!.in_batch()) output_path <- .unique_path(output_path)
writeLAS(las_out, output_path)

cat(sprintf("\nSaved %s points -> %s\n",
            format(nrow(xyz_new_exp), big.mark = ","), output_path))
```

### Data - Applied-Transformation Summary

```{r}
#| label: transform-summary-stage6
#| code-summary: "Final summary: the nine quantities that describe what was applied"

# ── Where the numbers come from ──────────────────────────────────────────────
# Every quantity below is read off M_export - the single matrix Step 6.4
# actually applied - rather than off the individual stage outputs. Those are
# reported too, in the provenance block, but they are not the summary: a stage
# value that never made it into the composition (because its toggle was off, or
# because it was superseded) would otherwise appear here as though it had been
# applied.
#
# ONE SET OF SHIFT FIGURES, and it is t_exp. No rotation in this pipeline
# induces a translation:
#
#   - the Stage 1A rough alignment and the Stage 2 fine XYZ rotation turn about
#     the SAME point (Stage 1A fixes pivot1 before rotating and never recomputes
#     it; Step 2.5 asserts the two pivots agree), so their composition is a pure
#     rotation about that point;
#   - the Stage 4/5 Z-only rotation turns about pivot2, which differs from pivot1
#     only in Z - and a vertical offset of the pivot cannot affect a rotation
#     about the vertical axis. pivot2 is the same place relative to the cloud
#     after the Z-shift as pivot1 was before it.
#
# So the translation columns that appear in M_export are pivot-recentring terms
# and nothing else. Reporting them as a second, "net" set of shifts would be
# quoting the pivot's coordinates back as though they were a measurement.
.rot_applied  <- .decompose_rotation(R_combined_exp)
.shift_applied <- .shift_specs(t_exp)

cat("\n─── Applied Transformations Summary ─────────────────────────────────────\n")
cat(sprintf("  toggles: xyz-rot %s | z-rot %s | xy-trans %s | z-trans %s\n",
            if (apply_xyz_rotation)   "ON " else "OFF",
            if (apply_z_rotation)     "ON " else "OFF",
            if (apply_xy_translation) "ON " else "OFF",
            if (apply_z_translation)  "ON " else "OFF"))

cat("\n  ── Rotation, as applied (from R_combined_exp) ─────────────────────────\n")
cat(sprintf("    XY rotation magnitude : %10.4f deg   (tilt away from vertical)\n",
            .rot_applied$xy_mag_deg))
cat(sprintf("    XY rotation direction : %10s deg   (azimuth the +Z axis leans toward, CCW from +X)\n",
            if (is.na(.rot_applied$xy_dir_deg)) "n/a" else sprintf("%.4f", .rot_applied$xy_dir_deg)))
cat(sprintf("    Z rotation (yaw)      : %+10.4f deg   (about vertical, CCW from above)\n",
            .rot_applied$z_deg))
cat(sprintf("    full XYZ rotation     : %10.4f deg   (total turn about the matrix's own axis)\n",
            .rot_applied$total_deg))
cat(sprintf("    factorisation residual: %.2e (the three angles rebuild the matrix exactly)\n",
            .rot_applied$rebuild_err))

cat("\n  ── Translation, as applied (t_exp) ────────────────────────────────────\n")
cat(sprintf("    X-only shift          : %+10.5f m\n",  .shift_applied$x))
cat(sprintf("    Y-only shift          : %+10.5f m\n",  .shift_applied$y))
cat(sprintf("    Z-only shift          : %+10.5f m   (%+.2f cm)\n",
            .shift_applied$z, .shift_applied$z * 100))
cat(sprintf("    XY shift              : %10.5f m\n",  .shift_applied$xy))
cat(sprintf("    XYZ shift             : %10.5f m\n",  .shift_applied$xyz))

# Cross-check, not a second report: re-expressing M_export as "rotate about
# pivot1, then shift" must return t_exp. If it does not, a rotation somewhere in
# the composition has picked up a pivot of its own and is smuggling in a
# translation - exactly the failure the one-pivot rule exists to prevent.
.shift_check <- max(abs(.shift_about(M_export, pivot1) - as.numeric(t_exp)))
cat(sprintf("    (M_export re-expressed about pivot1 returns t_exp to %.2e m -\n",
            .shift_check))
cat("     confirming no rotation in the composition induced a shift)\n")
if (.shift_check > 1e-6)
  warning(sprintf(paste0("The composed transform carries %.5f m of translation beyond t_exp. ",
                         "A rotation is turning about a pivot the others do not share - check ",
                         "Stage 1A.4 (pivot1 must not be recomputed) and Step 2.5."),
                  .shift_check), call. = FALSE)

cat("\n  ── Provenance: what each stage reported ───────────────────────────────\n")
cat(sprintf("    Stage 1A rough align  : %10.4f deg about pivot1\n",
            acos(pmax(-1, pmin(1, (sum(diag(R_align_1)) - 1) / 2))) * 180 / pi))
cat(sprintf("    Stage 2 fine XYZ      : %10.4f deg about pivot1 (IRLS-SVD)\n", angle_xyz))
cat(sprintf("    Stage 1A + 2 composed : %10.4f deg (R_align_xyz_1)\n", angle_align_xyz_1))
if (exists("tilt_dir_deg") && !is.na(tilt_dir_deg))
  cat(sprintf("      Stage 2 data-derived tilt direction: %.2f deg (consistency %.4f)\n",
              tilt_dir_deg, tilt_consistency))
cat(sprintf("    Stage 4 Z translation : %+10.5f m\n", t_trans_z_applied))
cat(sprintf("    Stage 4/5 Z rotation  : %+10.4f deg (line-network solve)\n", z_rotation_deg))
cat(sprintf("    Stage 5 XY translation: dx %+.5f m, dy %+.5f m\n",
            t_trans_xy[1], t_trans_xy[2]))

cat("\n  ── Frame and output ───────────────────────────────────────────────────\n")
cat(sprintf("    rotation pivot        : (%.4f, %.4f, %.4f)  [pivot1; every leg]\n",
            pivot1[1], pivot1[2], pivot1[3]))
cat(sprintf("    export source         : %s cloud, %s points, all TreeIDs incl. 0/NA\n",
            export_source, format(nrow(xyz_new_exp), big.mark = ",")))
cat(sprintf("    output                : %s\n", output_path))
cat("─────────────────────────────────────────────────────────────────────────\n")

# ── 4x4 homogeneous form, for CloudCompare / external tools ──────────────────
# This is the matrix Step 6.4 actually applied - not a second, independently
# assembled copy of it. Its translation column already folds in every pivot
# recentring term for pivot1, so it is expressed in world coordinates and can be
# pasted straight into an external tool. Its translation column is NOT the shift -
# the shift is t_exp, reported above.

cat("\nEquivalent 4x4 homogeneous transform (world coordinates):\n")
print(format(M_export, digits = 12))
```

### Vis - final alignment check

```{r}
#| label: fig-final-alignment-3D
#| fig-cap: "Paired stems after the full composed transform, overlaid on the reference epoch."

if (params$make_plots && exists("las_2") && nrow(pairs_stage5) > 0L) {

  inspect_z_min_final   <- 1
  inspect_z_max_final   <- 5
  points_per_tree_final <- 1000

  las1_final <- data.frame(
    X = xyz_new_exp[, 1], Y = xyz_new_exp[, 2], Z = xyz_new_exp[, 3],
    TreeID = las_export@data$TreeID
  )
  las1_final <- las1_final[!is.na(las1_final$TreeID) & las1_final$TreeID != 0L, ]

  las2_final <- data.frame(
    X = las_2@data$X, Y = las_2@data$Y, Z = las_2@data$Z,
    TreeID = las_2@data$TreeID
  )
  las2_final <- las2_final[!is.na(las2_final$TreeID) & las2_final$TreeID != 0L, ]

  extract_slice_final <- function(df, tree_ids) {
    df <- df[df$TreeID %in% tree_ids, ]
    df |>
      dplyr::group_by(TreeID) |>
      dplyr::mutate(Z_rel = Z - min(Z)) |>
      dplyr::filter(Z_rel >= inspect_z_min_final, Z_rel <= inspect_z_max_final) |>
      dplyr::ungroup()
  }

  inspect_1_final <- extract_slice_final(las1_final, pairs_stage5$TreeID_1) |>
    dplyr::mutate(cloud = "las_1 (all transforms)")
  inspect_2_final <- extract_slice_final(las2_final, pairs_stage5$TreeID_2) |>
    dplyr::mutate(cloud = "las_2")

  inspect_df_final <- dplyr::bind_rows(inspect_1_final, inspect_2_final) |>
    dplyr::group_by(cloud, TreeID) |>
    dplyr::group_modify(~ {
      n_pts  <- nrow(.x)
      n_take <- min(points_per_tree_final, n_pts)
      .x[sample.int(n_pts, n_take, replace = FALSE), ]
    }) |>
    dplyr::ungroup()

  cat("Inspection points (final):", nrow(inspect_df_final), "\n")

  cx_f <- mean(inspect_df_final$X)
  cy_f <- mean(inspect_df_final$Y)
  inspect_df_final <- inspect_df_final |>
    dplyr::mutate(X_local = X - cx_f, Y_local = Y - cy_f)

  plot_ly(
    data   = inspect_df_final,
    x = ~X_local, y = ~Y_local, z = ~Z,
    color  = ~cloud,
    colors = c("las_1 (all transforms)" = COL_1, "las_2" = COL_2),
    type   = "scatter3d", mode = "markers",
    marker = list(size = 2, opacity = 0.6),
    text   = ~paste0("TreeID: ", TreeID, "<br>Cloud: ", cloud),
    hoverinfo = "text"
  ) |>
    layout(
      title = sprintf("Temporal pairs after all transforms (1-5 m slice, %d pairs)",
                      nrow(pairs_stage5)),
      scene = list(
        xaxis = list(title = "X local (m)"),
        yaxis = list(title = "Y local (m)"),
        zaxis = list(title = "Z (m)"),
        aspectmode = "data"
      ),
      legend = list(title = list(text = "Cloud"))
    )
}
```

## 6.6 Alignment Error Against a Known Truth

Only meaningful on a **generated test sample** - one built by the test-sample
generator from a perfectly aligned baseline, where the correct answer was
recorded before the pipeline ever saw the cloud. On real data there is no truth
to compare against and this step should stay switched off.

The targets are read straight out of the generator's CSV. Nothing is retyped:
the exact target rotation matrix (`target_R_*`) and the truth pivot
(`pivot_shifted_*`) come from the file, so the comparison never depends on
angles that were rounded for display, and a mismatch between the sample on disk
and the truth being scored against it is impossible as long as the right CSV is
named.

Two things this does *not* do, both deliberate.

**It does not subtract reported angles.** Comparing a recovered tilt of 0.58 deg
against a target of 0.59 deg and calling the difference 0.01 deg is wrong
whenever more than one axis is involved, because rotations about different axes
do not add or subtract. The error is the *residual rotation*
$R_{\text{err}} = R_{\text{recovered}} R_{\text{target}}^{\top}$ - literally what
is left unrotated once the pipeline's answer has been applied. A perfect run
makes it the identity, and its tilt, yaw and total angle are all zero together
exactly when the run was right, which a difference of angles is not.

**It does not compare raw translation columns.** In UTM coordinates the
translation column of a world-frame 4x4 is dominated by pivot-recentring terms
running to hundreds of metres, and it changes completely if the pivot moves even
when the transform is identical. Both transforms are therefore re-expressed as
"rotate about the truth pivot, then shift", and those shifts are what get
compared. The truth pivot is the one the generator turned the sample about - not
`pivot1`, unless Step 0.7 was set to read the same CSV, in which case they
coincide and the intermediate stage printouts become directly readable against
the truth as well.

**The primary shift measurement is the marker box, not the matrix.** The
generator plants eight points at the vertices of the reference cloud's bounding
box twice over: once at their truth positions (`TargetBox = 1`) and once moved
by the forward transform (`TargetBox = 2`), with `BoxCorner` 1-8 identifying
which vertex is which on both copies. This step puts the box-2 corners through
`M_export` - the exact matrix Step 6.4 applied to the cloud - and differences
each against its box-1 twin. That is a displacement in metres measured on
geometry the pipeline never saw, and no pivot convention, no factorisation and
no choice of frame enters it.

The residual is defined as the user specified it: the per-corner differences are
averaged **as vectors**, one average per offset type (X, Y, Z, XY, XYZ), and the
*magnitude of the average vector* is the residual. Averaging before taking the
magnitude is deliberate - it measures the net displacement of the box as a body,
which is what a shift error is - but it has one consequence worth stating
plainly, because it is not a defect to be fixed so much as a property to know:

> A rotation error about the centre of the box moves every corner and averages
> to **exactly zero**. Simulated on a 60 x 45 x 35 m box, a 0.05 deg yaw error
> displaces all eight corners by 3.26 cm and returns an XYZ residual of 0.00000 m.

So the averaged residuals answer "did the box end up in the right place", and
the rotation error terms answer "did it end up in the right attitude" - they are
complementary and neither substitutes for the other. The per-corner distances
are recorded alongside (`box_percorner_mean_m`, `box_percorner_max_m`, and one
column per corner), and those do *not* cancel: on that same simulated yaw error
they read 3.26 cm. If the averaged residuals are near zero and the per-corner
distances are not, the box is in the right place and turned the wrong way.

Corners are read from the generator's CSV when it carries them, because those
coordinates are exact; the copies inside the `.las` went through the file's
scale factor and are quantised. The cloud's own `TargetBox` field is the
fallback.

Alongside the error terms, the exported row carries the run's own provenance -
every rotation matrix in the composition, and the pair counts recorded by
`run_counts` at each filter. That is what turns a folder of these CSVs into
something answerable: *did the runs that lost most of their pairs at the
straightness gate also come out worst?*

```{r}
#| label: params-alignment-error-vs-truth
#| code-summary: "Parameters - Score this run against the generator's truth CSV and write a full result row"


score_against_truth <- FALSE

# The generator's truth CSV for the sample this run processed.
truth_csv_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/TestSamples/Generated/SNBG_BOS_F1_Feb2020_March2026_ULS_TLS.csv"

# Result rows are written one per run and never overwritten (.unique_path).
result_csv_dir <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/TestSamples/Results"

.par_apply("params-alignment-error-vs-truth")
```

```{r}
#| label: alignment-error-vs-truth
#| code-summary: "Score this run against the generator's truth CSV and write a full result row"
#| eval: false

# Scoring the run against the truth is the POINT of a batch, so all three are
# taken from the driver here. The chunk carries eval: false and FALSE defaults
# for interactive use, and the driver force-evaluates it - but the switch and
# the paths have to be read inside the chunk, because it uses them in the same
# chunk that sets them. Pushed in as an after-override they landed too late,
# leaving score_against_truth FALSE: the chunk ran, did nothing, built no
# alignment_result, and every pair was reported as no_result.
if (.in_batch()) {
  score_against_truth <- isTRUE(.batch_opt(".batch_score_truth", TRUE))
  truth_csv_path      <- .batch_opt(".batch_truth_csv",  truth_csv_path)
  result_csv_dir      <- .batch_opt(".batch_results_dir", result_csv_dir)
}

if (isTRUE(score_against_truth)) {

  # ── 1. Read the truth ──────────────────────────────────────────────────────
  if (!file.exists(truth_csv_path))
    stop("[6.6] Generator truth CSV not found:\n  ", truth_csv_path)
  truth <- utils::read.csv(truth_csv_path, stringsAsFactors = FALSE)
  if (nrow(truth) != 1L)
    stop(sprintf("[6.6] Truth CSV has %d rows; expected exactly 1. Subset it to the sample this run used.",
                 nrow(truth)))

  .nm <- sprintf("target_R_%d%d", rep(1:3, each = 3), rep(1:3, 3))
  .need <- c(.nm, "target_shift_x_m", "target_shift_y_m", "target_shift_z_m",
             "pivot_shifted_x", "pivot_shifted_y", "pivot_shifted_z")
  if (!all(.need %in% names(truth)))
    stop("[6.6] The CSV is missing columns this step needs (target_R_*, target_shift_*, ",
         "pivot_shifted_*). It is not a generator truth file, or it predates the current generator.")

  R_target    <- matrix(as.numeric(unlist(truth[1, .nm])), nrow = 3, byrow = TRUE)
  t_target    <- c(truth$target_shift_x_m, truth$target_shift_y_m, truth$target_shift_z_m)
  truth_pivot <- c(truth$pivot_shifted_x,  truth$pivot_shifted_y,  truth$pivot_shifted_z)

  if (!isTRUE(all.equal(t(R_target) %*% R_target, diag(3),
                        tolerance = 1e-6, check.attributes = FALSE)))
    stop("[6.6] The target rotation read from the CSV is not a rotation matrix - the file is corrupt.")

  # Reported, not enforced: the pipeline may legitimately have chosen its own
  # pivot. But if Step 0.7 was set to "csv" these coincide, and a large offset
  # when they were meant to coincide means the wrong CSV is named here.
  .pivot_gap <- sqrt(sum((as.numeric(pivot1) - truth_pivot)^2))

  # ── 2. Rotation error, from the residual rotation ──────────────────────────
  R_recovered <- R_combined_exp
  R_err       <- R_recovered %*% t(R_target)
  .e_rot <- .decompose_rotation(R_err)
  .t_rot <- .decompose_rotation(R_target)
  .r_rot <- .decompose_rotation(R_recovered)

  # ── 3. Shift error, both transforms referred to the truth pivot ────────────
  s_recovered <- .shift_about(M_export, truth_pivot)
  .d          <- s_recovered - t_target
  .rs <- .shift_specs(s_recovered); .ts <- .shift_specs(t_target)

  # XY and XYZ errors are the length of the error VECTOR, not the difference of
  # the two lengths. Two shifts of equal length pointing different ways would
  # otherwise score zero error, which is the opposite of the truth.
  .es <- list(x = .d[1], y = .d[2], z = .d[3],
              xy = sqrt(sum(.d[1:2]^2)), xyz = sqrt(sum(.d^2)))

  # ── 4. End-to-end residual ─────────────────────────────────────────────────
  # Rotation error and shift error folded together: where a point actually lands
  # versus where it should have. This is what a reader sees on opening the two
  # clouds, and it is the number to quote if only one is quoted.
  T_target <- diag(4)
  T_target[1:3, 1:3] <- R_target
  T_target[1:3, 4]   <- as.numeric(truth_pivot - R_target %*% truth_pivot + t_target)

  .at <- function(M, p) as.numeric(M[1:3, 1:3] %*% p + M[1:3, 4])
  .resid_pivot <- sqrt(sum((.at(M_export, truth_pivot) - .at(T_target, truth_pivot))^2))

  .resid_rms <- NA_real_; .resid_max <- NA_real_
  if (exists("xyz_orig_exp")) {
    .a  <- t(M_export[1:3, 1:3] %*% t(xyz_orig_exp) + M_export[1:3, 4])
    .b  <- t(T_target[1:3, 1:3] %*% t(xyz_orig_exp) + T_target[1:3, 4])
    .dd <- sqrt(rowSums((.a - .b)^2))
    .resid_rms <- sqrt(mean(.dd^2)); .resid_max <- max(.dd)
    rm(.a, .b, .dd)
  }

  # ── 4b. Marker-box residuals - the primary shift measurement ───────────────
  # Two sources, in order of preference:
  #   csv -> box1_c1_x ... box2_c8_z, the exact coordinates the generator used
  #   las -> the TargetBox / BoxCorner fields in the cloud, quantised by the
  #          file's scale factor (sub-millimetre, but it is a real difference and
  #          the report says which source it used)
  # Absent from both -> every box_* column comes out NA and the transform-based
  # terms above stand alone. That is the case on any real, non-generated sample.
  .bfld <- if (!is.null(truth$box_field_name)) as.character(truth$box_field_name) else
           if (exists("box_field_name")) box_field_name else "TargetBox"
  .cfld <- if (!is.null(truth$corner_field_name)) as.character(truth$corner_field_name) else
           if (exists("corner_field_name")) corner_field_name else "BoxCorner"

  box_corners_truth <- NULL; box_corners_moved <- NULL
  box_corner_ids    <- integer(0)
  .box_source       <- "none"

  .b1nm <- sprintf("box1_c%d_%s", rep(1:8, each = 3), rep(c("x", "y", "z"), 8))
  .b2nm <- sprintf("box2_c%d_%s", rep(1:8, each = 3), rep(c("x", "y", "z"), 8))

  if (all(c(.b1nm, .b2nm) %in% names(truth))) {
    box_corners_truth <- matrix(as.numeric(unlist(truth[1, .b1nm])), ncol = 3, byrow = TRUE)
    box_corners_moved <- matrix(as.numeric(unlist(truth[1, .b2nm])), ncol = 3, byrow = TRUE)
    box_corner_ids    <- 1:8
    .box_source       <- "csv"

  } else if (exists("las_1_raw") &&
             all(c(.bfld, .cfld) %in% names(las_1_raw@data))) {
    .bt <- as.data.frame(las_1_raw@data)[, c("X", "Y", "Z", .bfld, .cfld)]
    names(.bt) <- c("X", "Y", "Z", "box", "corner")
    .m1 <- .bt[!is.na(.bt$box) & .bt$box == 1L, ]
    .m2 <- .bt[!is.na(.bt$box) & .bt$box == 2L, ]

    # Matched on the corner ID, never on proximity - a badly wrong run is exactly
    # the case where the nearest box-1 corner to a moved box-2 corner is not its
    # twin, so pairing by distance would flatter the very runs it should punish.
    if (nrow(.m1) > 0L && nrow(.m1) == nrow(.m2) &&
        !anyDuplicated(.m1$corner) && setequal(.m1$corner, .m2$corner)) {
      .m1 <- .m1[order(.m1$corner), ]; .m2 <- .m2[order(.m2$corner), ]
      box_corners_truth <- as.matrix(.m1[, c("X", "Y", "Z")])
      box_corners_moved <- as.matrix(.m2[, c("X", "Y", "Z")])
      dimnames(box_corners_truth) <- NULL; dimnames(box_corners_moved) <- NULL
      box_corner_ids    <- as.integer(.m1$corner)
      .box_source       <- "las"
    } else if (nrow(.m1) > 0L || nrow(.m2) > 0L) {
      warning(sprintf(paste0("[6.6] The cloud carries %d %s = 1 and %d %s = 2 point(s) but they do ",
                             "not pair up one-to-one on %s. Marker residuals skipped."),
                      nrow(.m1), .bfld, nrow(.m2), .bfld, .cfld), call. = FALSE)
    }
    rm(.bt, .m1, .m2)
  }

  .box_ok    <- !is.null(box_corners_truth)
  box_res    <- list(x = NA_real_, y = NA_real_, z = NA_real_,
                     xy = NA_real_, xyz = NA_real_)
  box_mean_vec  <- c(NA_real_, NA_real_, NA_real_)
  box_percorner <- numeric(0)
  box_delta     <- NULL
  box_recovered <- NULL

  if (.box_ok) {
    # The same matrix Step 6.4 applied to every point of the cloud, applied to
    # the exact corner coordinates rather than to their quantised copies.
    box_recovered <- t(M_export[1:3, 1:3] %*% t(box_corners_moved) + M_export[1:3, 4])

    # Per-corner difference vectors: where this run put the corner, minus where
    # it belongs. Sign convention matches the shift errors above (recovered
    # minus reference), so a positive dx means the run overshot in +X.
    box_delta    <- box_recovered - box_corners_truth
    box_mean_vec <- colMeans(box_delta)

    # Average the VECTORS, then take the magnitude - not the other way round.
    # X, Y and Z are one-component vectors, so their magnitude is an absolute
    # value; XY and XYZ take the length of the averaged 2- and 3-vectors.
    box_res <- list(x   = abs(box_mean_vec[1]),
                    y   = abs(box_mean_vec[2]),
                    z   = abs(box_mean_vec[3]),
                    xy  = sqrt(sum(box_mean_vec[1:2]^2)),
                    xyz = sqrt(sum(box_mean_vec^2)))

    # Per-corner distances, which do NOT cancel. A rotation error about the box
    # centre averages to zero above and shows up here, so the two together say
    # both where the box landed and how it is turned.
    box_percorner <- sqrt(rowSums(box_delta^2))
  }

  # ── 5. Report ──────────────────────────────────────────────────────────────
  .fd <- function(v) if (is.null(v) || is.na(v)) "n/a" else sprintf("%.4f", v)

  cat("\n═══ Alignment error against the known truth ══════════════════════════════\n")
  cat(sprintf("  sample     : %s\n", truth$sample_name))
  cat(sprintf("  truth pivot: (%.4f, %.4f, %.4f)   pivot1 offset from it: %.4f m\n\n",
              truth_pivot[1], truth_pivot[2], truth_pivot[3], .pivot_gap))
  cat("  ── Marker-box residuals (the primary shift measurement) ───────────────\n")
  if (!.box_ok) {
    cat("    no marker box in this sample - the transform-based terms below stand alone\n")
  } else {
    cat(sprintf("    %d corner(s), coordinates from the %s\n", nrow(box_delta),
                if (identical(.box_source, "csv")) "generator CSV (exact)"
                else "cloud's marker fields (LAS-quantised)"))
    cat(sprintf("\n    %-8s %12s %12s %12s %14s\n",
                "corner", "dx (m)", "dy (m)", "dz (m)", "distance (m)"))
    for (.bi in seq_len(nrow(box_delta)))
      cat(sprintf("    %-8d %12.5f %12.5f %12.5f %14.5f\n",
                  box_corner_ids[.bi], box_delta[.bi, 1], box_delta[.bi, 2],
                  box_delta[.bi, 3], box_percorner[.bi]))
    cat(sprintf("    %-8s %12.5f %12.5f %12.5f %14s\n", "mean",
                box_mean_vec[1], box_mean_vec[2], box_mean_vec[3], "-"))

    cat(sprintf("\n    X residual  : %.5f m        (|mean dx|)\n",   box_res$x))
    cat(sprintf("    Y residual  : %.5f m        (|mean dy|)\n",     box_res$y))
    cat(sprintf("    Z residual  : %.5f m        (|mean dz|)\n",     box_res$z))
    cat(sprintf("    XY residual : %.5f m        (length of the mean XY vector)\n",  box_res$xy))
    cat(sprintf("    XYZ residual: %.5f m        (length of the mean XYZ vector)\n", box_res$xyz))
    cat(sprintf("\n    per-corner distance: mean %.5f m | max %.5f m\n",
                mean(box_percorner), max(box_percorner)))
    # The floor keeps this quiet on a perfect run, where both numbers are
    # floating-point noise and the ratio between them means nothing.
    if (mean(box_percorner) > 1e-4 && box_res$xyz < 0.5 * mean(box_percorner))
      cat("    NOTE: the averaged residual is well below the per-corner distances, so the\n",
          "         corners are moving in opposing directions - the box is close to the\n",
          "         right place but turned the wrong way. Read the rotation errors below.\n", sep = "")
  }

  cat("\n  ── Transform-based error terms (recorded alongside) ───────────────────\n")
  cat(sprintf("  %-28s %13s %13s %13s\n", "", "recovered", "target", "error"))
  cat(sprintf("  %-28s %13.4f %13.4f %13.4f\n", "XY rotation magnitude (deg)",
              .r_rot$xy_mag_deg, .t_rot$xy_mag_deg, .e_rot$xy_mag_deg))
  cat(sprintf("  %-28s %13s %13s %13s\n", "XY rotation direction (deg)",
              .fd(.r_rot$xy_dir_deg), .fd(.t_rot$xy_dir_deg), .fd(.e_rot$xy_dir_deg)))
  cat(sprintf("  %-28s %13.4f %13.4f %13.4f\n", "Z rotation (deg)",
              .r_rot$z_deg, .t_rot$z_deg, .e_rot$z_deg))
  cat(sprintf("  %-28s %13.4f %13.4f %13.4f\n", "full XYZ rotation (deg)",
              .r_rot$total_deg, .t_rot$total_deg, .e_rot$total_deg))
  cat(sprintf("  %-28s %13.5f %13.5f %13.5f\n", "X-only shift (m)", .rs$x,  .ts$x,  .es$x))
  cat(sprintf("  %-28s %13.5f %13.5f %13.5f\n", "Y-only shift (m)", .rs$y,  .ts$y,  .es$y))
  cat(sprintf("  %-28s %13.5f %13.5f %13.5f\n", "Z-only shift (m)", .rs$z,  .ts$z,  .es$z))
  cat(sprintf("  %-28s %13.5f %13.5f %13.5f\n", "XY shift (m)",     .rs$xy, .ts$xy, .es$xy))
  cat(sprintf("  %-28s %13.5f %13.5f %13.5f\n", "XYZ shift (m)",    .rs$xyz, .ts$xyz, .es$xyz))
  cat("\n  (the XY and XYZ error entries are the length of the error vector, not the\n")
  cat("   difference of the two lengths - equal-length shifts pointing different ways\n")
  cat("   are not a zero error)\n")

  cat("\n  ── End-to-end residual ────────────────────────────────────────────────\n")
  cat(sprintf("    at the truth pivot     : %.5f m\n", .resid_pivot))
  if (!is.na(.resid_rms))
    cat(sprintf("    over the exported cloud: RMS %.5f m | max %.5f m\n", .resid_rms, .resid_max))
  cat("    (the rotation error acts on a lever arm, so the plot edge moves further than\n")
  cat("     the pivot does - the max is the worst case a reader will see)\n")

  # ── 6. Provenance: the matrices and the pair counts ────────────────────────
  # Every rotation in the composition, kept separately. A run can be wrong
  # because the rough alignment was wrong, because the fine solve was wrong, or
  # because they composed badly, and only the separate matrices distinguish
  # those three.
  .flat <- function(M, prefix) {
    v  <- as.vector(t(M))
    nm <- as.vector(t(outer(seq_len(nrow(M)), seq_len(ncol(M)),
                            function(i, j) sprintf("%s_%d%d", prefix, i, j))))
    stats::setNames(as.list(v), nm)
  }

  # Same, but for a 3x3 that a checkpoint may predate: emits the identical nine
  # column names filled with NA rather than dropping them, so a folder of result
  # rows from different pipeline versions still rbinds.
  .flat_opt <- function(nm_obj, prefix) {
    if (exists(nm_obj, envir = .stage_env(), inherits = FALSE) &&
        is.matrix(get(nm_obj, envir = .stage_env())))
      .flat(get(nm_obj, envir = .stage_env()), prefix)
    else .flat(matrix(NA_real_, 3, 3), prefix)
  }
  .val_opt <- function(nm_obj, default = NA) {
    if (exists(nm_obj, envir = .stage_env(), inherits = FALSE))
      get(nm_obj, envir = .stage_env()) else default
  }

  # run_counts is built up by .tally() across Stages 1-5 and travels in the
  # checkpoints. Read through a helper so a checkpoint that predates the tally
  # yields NA columns rather than stopping the report.
  .rc <- if (exists("run_counts")) run_counts else list()
  .cnt <- function(k) { v <- .rc[[k]]; if (is.null(v)) NA else v }

  result <- c(
    list(
      # ── identity ────────────────────────────────────────────────────────
      sample_name         = truth$sample_name,
      scored_utc          = format(Sys.time(), tz = "UTC", usetz = TRUE),
      truth_csv           = truth_csv_path,
      aligned_las         = if (exists("output_path")) output_path else NA_character_,
      toggles             = sprintf("xyzrot=%s;zrot=%s;xytrans=%s;ztrans=%s",
                                    apply_xyz_rotation, apply_z_rotation,
                                    apply_xy_translation, apply_z_translation),

      # ── frames ──────────────────────────────────────────────────────────
      truth_pivot_x       = truth_pivot[1],
      truth_pivot_y       = truth_pivot[2],
      truth_pivot_z       = truth_pivot[3],
      pivot1_x            = as.numeric(pivot1)[1],
      pivot1_y            = as.numeric(pivot1)[2],
      pivot1_z            = as.numeric(pivot1)[3],
      pivot1_mode         = if (exists("pivot1_mode")) pivot1_mode else NA_character_,
      pivot1_offset_m     = .pivot_gap,
      # Recorded so a run where the two pivots drifted apart is identifiable
      # after the fact; it should be 0 on every correct run.
      pivot_leg_gap_m     = sqrt(sum((as.numeric(align_pivot_1) - as.numeric(pivot1))^2)),

      # ── (1) APPLIED: what this run actually did ─────────────────────────
      applied_rot_xy_mag_deg = .r_rot$xy_mag_deg,
      applied_rot_xy_dir_deg = .r_rot$xy_dir_deg,
      applied_rot_z_deg      = .r_rot$z_deg,
      applied_rot_total_deg  = .r_rot$total_deg,
      applied_shift_x_m      = .rs$x,
      applied_shift_y_m      = .rs$y,
      applied_shift_z_m      = .rs$z,
      applied_shift_xy_m     = .rs$xy,
      applied_shift_xyz_m    = .rs$xyz,

      # ── (2) REFERENCE: what a perfect run would have done ───────────────
      reference_rot_xy_mag_deg = .t_rot$xy_mag_deg,
      reference_rot_xy_dir_deg = .t_rot$xy_dir_deg,
      reference_rot_z_deg      = .t_rot$z_deg,
      reference_rot_total_deg  = .t_rot$total_deg,
      reference_shift_x_m      = .ts$x,
      reference_shift_y_m      = .ts$y,
      reference_shift_z_m      = .ts$z,
      reference_shift_xy_m     = .ts$xy,
      reference_shift_xyz_m    = .ts$xyz,

      # ── (3) ERROR ───────────────────────────────────────────────────────
      err_rot_xy_mag_deg  = .e_rot$xy_mag_deg,
      err_rot_xy_dir_deg  = .e_rot$xy_dir_deg,
      err_rot_z_deg       = .e_rot$z_deg,
      err_rot_total_deg   = .e_rot$total_deg,
      err_shift_x_m       = .es$x,
      err_shift_y_m       = .es$y,
      err_shift_z_m       = .es$z,
      err_shift_xy_m      = .es$xy,
      err_shift_xyz_m     = .es$xyz,
      resid_at_pivot_m    = .resid_pivot,
      resid_cloud_rms_m   = .resid_rms,
      resid_cloud_max_m   = .resid_max,

      # ── (3b) MARKER-BOX RESIDUALS - the primary shift measurement ────────
      # box_res_* are the magnitudes of the corner-averaged difference vectors.
      # box_mean_d* are the same averages kept signed, so a systematic offset
      # keeps its direction when a folder of these rows is stacked.
      box_source          = .box_source,
      box_n_corners       = if (.box_ok) nrow(box_delta) else NA_integer_,
      box_mean_dx_m       = box_mean_vec[1],
      box_mean_dy_m       = box_mean_vec[2],
      box_mean_dz_m       = box_mean_vec[3],
      box_res_x_m         = box_res$x,
      box_res_y_m         = box_res$y,
      box_res_z_m         = box_res$z,
      box_res_xy_m        = box_res$xy,
      box_res_xyz_m       = box_res$xyz,
      # These do not cancel, and are what distinguishes a displaced box from a
      # rotated one.
      box_percorner_mean_m = if (.box_ok) mean(box_percorner) else NA_real_,
      box_percorner_max_m  = if (.box_ok) max(box_percorner)  else NA_real_,

      # ── (8-17) how many things survived each filter ─────────────────────
      n_treeids_1           = .cnt("n_treeids_1"),          # (8)  unique TreeIDs, las_1
      n_treeids_2           = .cnt("n_treeids_2"),          # (8)  unique TreeIDs, las_2
      n_pairs_original      = .cnt("n_pairs_original"),     # (9)  Stage 1 matched pairs
      irls_stem_group       = .cnt("irls_stem_group"),      # (10) "consensus" or "all"
      n_pairs_gate_passed   = .cnt("n_pairs_gate_passed"),
      n_pairs_irls_in       = .cnt("n_pairs_irls_in"),      # (11) pairs entering the IRLS-SVD
      n_pairs_irls_kept     = .cnt("n_pairs_irls_kept"),    # (11) survivors of the residual filter
      ztrans_mode           = .cnt("ztrans_mode"),
      n_pairs_ztrans_set    = .cnt("n_pairs_ztrans_set"),
      n_pairs_ztrans_topup  = .cnt("n_pairs_ztrans_topup"),
      n_pairs_stage3_in     = .cnt("n_pairs_stage3_in"),    # (12) entering 3.1
      n_pairs_pair_avg      = .cnt("n_pairs_pair_avg"),     # (13) given a per-pair average (3.15)
      n_pairs_stage4_in     = .cnt("n_pairs_stage4_in"),    # (14) entering Stage 4
      n_pairs_zoverlap_kept = .cnt("n_pairs_zoverlap_kept"),
      n_stems_lines_start   = .cnt("n_stems_lines_start"),
      n_stems_lines_kept    = .cnt("n_stems_lines_kept"),
      n_pairs_lines_kept    = .cnt("n_pairs_lines_kept"),   # (15) after inter-stem line filtering
      n_pairs_stage5_in     = .cnt("n_pairs_stage5_in"),    # (16) entering Stage 5
      n_pairs_xy_kept       = .cnt("n_pairs_xy_kept")       # (17) after XY displacement filtering
    ),
    # ── (4-7) every rotation in the composition, flattened row-major ───────
    # Which rotations Stage 1A actually applied, and the round-trip error proving
    # R_rough is their composition. NA-filled on a checkpoint that predates them.
    list(
      rough_sideways          = .val_opt("align_sideways", NA),
      rough_sideways_rotated  = .val_opt("align_sideways_rotated", NA),
      rough_sideways_probed   = .val_opt("align_sw_probed", NA),
      rough_orientation       = .val_opt("align_orientation", NA_character_),
      rough_ud_confident      = .val_opt("align_ud_confident", NA),
      rough_check_max_err_m   = .val_opt("align_check_max_err", NA_real_)
    ),
    .flat_opt("R_sideways_yaw", "R_rough_yaw"),   # 45 deg sideways probe, if kept
    .flat_opt("R_sideways_90",  "R_rough_tip90"), # 90 deg sideways tip, if applied
    .flat_opt("R_residual",     "R_rough_resid"), # verticalisation of the forest axis
    .flat(R_align_1,      "R_rough"),      # (4) Stage 1A rough vertical alignment
    .flat(R_xyz,          "R_fine"),       # (5) Stage 2 fine XYZ solve
    .flat(R_align_xyz_1,  "R_rough_fine"), # (6) rough + fine, composed (Stage 2 §2.5)
    .flat(R_combined_exp, "R_full"),       # (7) rough + fine + Z-only, as exported
    .flat(R_target,       "R_reference"),
    .flat(R_err,          "R_error"),
    .flat(M_export,       "M_applied"),
    .flat(T_target,       "M_reference"),
    # One column per corner: the distance that corner ended up from its truth
    # position. NA-filled when the sample carries no box, so the column set is
    # the same width on every run and a folder of these still rbinds.
    stats::setNames(
      as.list(if (.box_ok) box_percorner else rep(NA_real_, 8L)),
      sprintf("box_c%d_dist_m",
              if (.box_ok) box_corner_ids else 1:8))
  )

  alignment_result <- as.data.frame(result, stringsAsFactors = FALSE)

  cat("\n  ── Pairs surviving each filter ────────────────────────────────────────\n")
  cat(sprintf("    unique TreeIDs      : las_1 %s | las_2 %s\n",
              .cnt("n_treeids_1"), .cnt("n_treeids_2")))
  cat(sprintf("    original pairs      : %s\n", .cnt("n_pairs_original")))
  cat(sprintf("    IRLS-SVD (%s group): %s in -> %s kept\n",
              .cnt("irls_stem_group"), .cnt("n_pairs_irls_in"), .cnt("n_pairs_irls_kept")))
  cat(sprintf("    Stage 3 Z-translation: %s in -> %s averaged (3.15)\n",
              .cnt("n_pairs_stage3_in"), .cnt("n_pairs_pair_avg")))
  cat(sprintf("    Stage 4             : %s in -> %s after line filtering\n",
              .cnt("n_pairs_stage4_in"), .cnt("n_pairs_lines_kept")))
  cat(sprintf("    Stage 5             : %s in -> %s after XY filtering\n",
              .cnt("n_pairs_stage5_in"), .cnt("n_pairs_xy_kept")))
  if (length(.rc) == 0L)
    cat("    (run_counts is empty - this checkpoint predates the tally; re-run from Stage 1)\n")

  # ── 7. Export ──────────────────────────────────────────────────────────────
  # One wide row per run, so a folder of these stacks with rbind and joins onto
  # the generator's truth table on sample_name.
  dir.create(result_csv_dir, recursive = TRUE, showWarnings = FALSE)
  result_csv_path <- file.path(result_csv_dir, paste0(truth$sample_name, "_result.csv"))
  # As with the aligned cloud: unique names by hand, one file per sample in a
  # batch. A rerun of a pair should replace that pair's row, not leave a _001
  # beside it for the reader to date-sort.
  if (!.in_batch()) result_csv_path <- .unique_path(result_csv_path)
  utils::write.csv(alignment_result, result_csv_path, row.names = FALSE)
  cat(sprintf("\nSaved %d result column(s) -> %s\n", ncol(alignment_result), result_csv_path))
  cat("═════════════════════════════════════════════════════════════════════════\n")

  rm(.nm, .need, .e_rot, .t_rot, .r_rot, .d, .rs, .ts, .es, .at,
     .resid_pivot, .resid_rms, .resid_max, .fd, .flat, .rc, .cnt, .pivot_gap,
     .bfld, .cfld, .b1nm, .b2nm, .box_ok, .box_source)
}
```

Workflow complete

No checkpoint is written by Stage 6 – its output is the exported `.las` file
itself. To re-export with a different combination of components, change the
toggles in Step 6.1 and re-run from Step 6.2; nothing upstream needs to be
recomputed.
