---
title: "Forest Point Cloud Alignment – Stage 5"
subtitle: "Vertical profile window, Z-only rotation, and residual XY translation"
author: "Point Cloud Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: flatly
    highlight-style: github
    fig-width: 9
    fig-height: 6
    self-contained: true
execute:
  warning: false
  message: false
  eval: false
params:
  make_plots: true
---


# Overview – Stage 5 of 6

Stage 5 **solves the residual XY (horizontal) translation** `t_trans_xy` – the
final piece of the rigid alignment. By this point Stage 3 has rotated `las_1`
in full 3D and Stage 4 has shifted it vertically and solved the Z-only yaw, so
the only motion still to determine is how far the two epochs slide apart in the
horizontal plane.

The solve works entirely on the windowed average ellipse centres – one XY point
per stem per epoch – that Stage 4 produced:

1. **apply the Stage 4 yaw** (`z_rotation_deg`, solved but left unapplied) to the
   `las_1` average centres, rotating them in XY about `pivot2` with the same sign
   convention the cloud builders use, and restrict to the pairs that survived
   every upstream filter (`pairs_zrot_valid`, minus the stems the Stage 4
   angular-difference MAD filter rejected). This produces the fully transformed
   `las_1` centres `avg_centres_xyt_1`, measured against the fixed `las_2`
   reference `avg_centres_xyt_2` (5.1);
2. **measure the per-pair XY displacement** `(dx, dy)` between each transformed
   `las_1` centre and its `las_2` partner, and **remove outlier stems** whose
   displacement disagrees with the consensus, by MAD on `dx` / `dy` (5.2);
3. **average the surviving displacements** into a single
   `t_trans_xy = (mean dx, mean dy)` (5.3).

What this stage does not do

Stage 5 **does not apply** `t_trans_xy`, and it loads no point clouds for the
solve (a single optional 3D figure loads its own on demand). Following the same
solve-then-apply split as Stages 3 and 4, the translation is recorded here and
first **applied** in Stage 6, where the full rotation-and-translation transform
is composed and written out. `t_trans_xy` is the stage's headline output.


# Stage 0 - Setup


## 0.1 Load libraries
```{r}
#| label: setup
#| code-summary: "Load libraries and set plot theme"

library(lidR)       # LAS I/O and point cloud processing
library(ggplot2)    # Plotting
library(dplyr)      # Data wrangling
library(plotly)     # Interactive 3D centre figure
library(base64enc)  # Embed rendered plots into HTML galleries
library(htmltools)  # Assemble the HTML gallery output

# ── Driver-supplied parameter overrides ──────────────────────────────────────
# Every params-* chunk in this file ends with .par_apply("<its label>"). Run the
# file by hand and that call does nothing: the values the chunk just assigned
# stand. Run it under the batch driver, which puts a .batch_params list in the
# environment before the first chunk, and any entry keyed to that chunk label is
# assigned here - AFTER the defaults, so the driver's value is the one that
# survives into the functional chunk that follows.
#
# Why this exists. Parameters used to sit in the same chunk as the code that
# consumed them, so the driver's overrides - which land after a chunk finishes -
# arrived too late to matter, and every pair silently used the .qmd's hardcoded
# value. Splitting each chunk in two fixes the ordering; this helper is what the
# driver reaches through.
#
# The defaults stay in this file rather than moving into the driver. They are
# the single definition, the driver names only what a batch changes, and there
# is no second copy of ~670 parameters to drift out of step with this one.
#
# envir = parent.frame() so a call at chunk top level lands in the environment
# the chunk is being evaluated in - globalenv by hand, the pair environment
# under the driver - without needing .stage_env(), which is defined further down.
.par_apply <- function(label, envir = parent.frame()) {
  bp <- mget(".batch_params", envir = envir, ifnotfound = list(NULL),
             inherits = TRUE)[[1]]
  if (is.null(bp) || is.null(bp[[label]])) return(invisible(character(0)))
  vals <- bp[[label]]
  if (!length(vals)) return(invisible(character(0)))
  if (is.null(names(vals)) || any(!nzchar(names(vals))))
    stop(sprintf("[.par_apply] .batch_params[['%s']] must be a fully NAMED list.", label))
  # A name the chunk did not define is almost always a typo or a rename in the
  # .qmd that the driver has not caught up with. Silence there would mean the
  # batch runs on the default while looking like it was overridden, so it stops.
  unknown <- names(vals)[!vapply(names(vals), exists, logical(1),
                                 envir = envir, inherits = FALSE)]
  if (length(unknown))
    stop(sprintf(paste0("[.par_apply] %s: %d override(s) name nothing this chunk assigns:\n  %s\n",
                        "  Fix the name in the driver's .build_params(), or remove it."),
                 label, length(unknown), paste(unknown, collapse = "\n  ")))
  for (nm in names(vals)) assign(nm, vals[[nm]], envir = envir)
  cat(sprintf("  [params] %s <- driver (%s)\n", label, paste(names(vals), collapse = ", ")))
  invisible(names(vals))
}
```

## 0.2 Global Helpers
```{r}
#| label: params-helpers
#| code-summary: "Parameters - Colours, ggplot theme, stage save/load utilities, and the on-demand point-cloud loader"

# ============================================================
# Colours
# ============================================================
COL_1     <- "#2196F3"
COL_2     <- "#FF5722"
COL_MATCH <- "#D9F9A5"

# ============================================================
# Progress bars
#
# Thin wrappers around cli, using EXACTLY the idiom Stage 3's axis-RANSAC fit
# has always used, because that one works when these documents are run on their
# own. Their only job is to save repeating the same six lines at every long
# loop.
#
# THE ONE THING THAT MUST NOT BE CHANGED HERE is `.envir = parent.frame()`.
# cli ties a progress bar's lifetime to a frame and tears the bar down when
# that frame exits. .pb_new() is a function, so a bar created with cli's
# default `.envir` would be created and immediately destroyed the moment
# .pb_new returned - the loop then ticks a bar that no longer exists and
# nothing is ever drawn. parent.frame() hands the bar to the CALLER's frame,
# which is the loop, and is the difference between this working and silently
# doing nothing. (An earlier revision passed globalenv() here, which is why the
# bars added through this helper never appeared while the hand-written ones in
# 3.3, 3.6 and 3.9 did.)
#
# Under the batch driver these will NOT be visible: cli draws on stderr, and
# the driver holds a message sink, which cannot be split. That is expected and
# is not worked around here - the driver prints its own progress instead.
# ============================================================
# Progress bars
# ============================================================
# One set of helpers, two renderers, chosen at the moment a bar is created.
#
#   cli    a redrawing bar, when there is a terminal able to show one
#   lines  one ordinary printed line per update, when there is not
#
# The choice is not cosmetic. cli draws on STDERR, and the batch driver holds a
# sink(type = "message"), which cannot split - so under the driver a cli bar
# goes to the log file and nowhere else, arriving as carriage returns and
# escape codes wrapped around each other. That is what the last run's log shows
# for every fitting step except the library's own: minutes of work with nothing
# on screen, and an unreadable smear in the log afterwards.
#
# The printed-line renderer is what the RANSAC library's cat handler already
# does, and it is the one visible line in that log. Same format here, so a fit
# reports identically whether the progress comes from the library or from a
# loop in this file.
show_progress_bars <- TRUE

# Print at most one line per 5% of the total or per 15 seconds, whichever comes
# first, plus always the first and last. A 20-stem fit gives ~20 lines; a
# 5000-slice loop does not give 5000.
progress_step_frac    <- 0.05
progress_min_seconds  <- 15

.par_apply("params-helpers")
```

```{r}
#| label: helpers
#| code-summary: "Colours, ggplot theme, stage save/load utilities, and the on-demand point-cloud loader"

# ============================================================
# ggplot theme
# ============================================================
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      legend.position  = "bottom"
    )
)

# ============================================================
# Stage save/load utilities – shared by every stage file.
#
# Contract (identical in all six stage files):
# SAVE → hard ERROR if any *required* object is missing from the
# environment, so a checkpoint can never be written half-empty.
# A *warning* is raised for any missing *optional* object
# (diagnostic products that may legitimately not have been run);
# those are simply omitted from the file.
# LOAD → never errors on content. A *warning* is raised for each
# requested object that is absent from the RDS. Objects present in
# the RDS but not requested are COUNTED in the summary line and
# never listed. Everything found in the file is loaded either way,
# so nothing is silently discarded.
# PATHS → a save never overwrites: if the target path exists, a numeric
# suffix (_001, _002, ...) is appended until the name is free.
# ============================================================

#' Return `path` if free, otherwise path_001 / path_002 / ... (first free one).
.unique_path <- function(path) {
  if (!file.exists(path)) return(path)
  base_noext <- tools::file_path_sans_ext(path)
  ext        <- tools::file_ext(path)
  counter    <- 1L
  repeat {
    candidate <- sprintf("%s_%03d.%s", base_noext, counter, ext)
    if (!file.exists(candidate)) {
      cat(sprintf("Target already exists - using unique name: %s\n", candidate))
      return(candidate)
    }
    counter <- counter + 1L
  }
}

#' Which environment do the stage helpers read and write?
#'
#' A .qmd rendered or run by hand evaluates its chunks in the global
#' environment. The batch driver does not: it gives every LAS pair a fresh
#' environment of its own, so that a leftover object from the previous pair
#' cannot be picked up by one of the exists() guards this pipeline is full of.
#' Under the driver, everything the stages assign therefore lives THERE, and
#' not in globalenv.
#'
#' parent.env(environment()) is the environment this chunk was evaluated in:
#' globalenv() by hand, the pair environment under the driver. It is resolved
#' from where .stage_env was DEFINED rather than from where it is called, so it
#' returns the same environment inside a function as it does at chunk top
#' level, and nothing has to be told which mode it is in.
#'
#' This replaced a hardcoded .stage_env() throughout. Under the driver that
#' pointed at an environment the pipeline had never written to, so every object
#' in a save manifest read as missing and the checkpoint refused to write - the
#' Stage 1 failure that reported 30 missing required objects after a 64-minute
#' fit. (`run_counts` was the one it did find, and only because .tally()'s `<<-`
#' walked past the pair environment into globalenv. That is fixed below too.)
.stage_env <- function() parent.env(environment())

#' Is a batch driver running this file?
#'
#' The driver assigns .batch_mode <- TRUE into the pair environment before the
#' first chunk runs. It changes one thing, and only for the checkpoints: in a
#' batch, every stage runs in ONE session, so Stage N+1 already holds in memory
#' everything Stage N computed and never has to read it back off disk. The
#' checkpoints stop being load-bearing and become a record of the run.
#'
#' So under a batch: nothing in a manifest is required (a missing object is
#' reported and left out, and the pair carries on), a failed write is reported
#' rather than thrown, and .stage_load is a no-op. Run the same .qmd by hand and
#' every one of those is an error again, which is what it should be when the
#' checkpoint is the only thing carrying the stage forward.
.in_batch <- function() {
  isTRUE(mget(".batch_mode", envir = .stage_env(),
              ifnotfound = list(FALSE), inherits = FALSE)[[1]])
}

#' Read one of the driver's batch settings out of the pair environment.
#' Absent - i.e. this file is being run by hand - gives `default`.
.batch_opt <- function(name, default = NULL) {
  v <- mget(name, envir = .stage_env(), ifnotfound = list(NULL), inherits = FALSE)[[1]]
  if (is.null(v)) default else v
}

#' Where a checkpoint goes when a batch is running.
#'
#' Every save chunk in this pipeline assigns its .rds path and then USES it a
#' few lines later, in the same chunk. The driver applies its overrides AFTER a
#' chunk has run, so an override of stage1_rds_path arrives after the file has
#' already been written to the path hardcoded here - which means, without this,
#' every sample in a batch writes over the same hardcoded name.
#'
#' So in batch mode the destination is derived instead, from two values the
#' driver puts in the environment before anything runs: .batch_rds_dir and
#' .batch_sample. The label the save chunk already passes supplies the rest, so
#' "Stage 3.2" becomes <dir>/<sample>_Stage3_2.rds with nothing to keep in sync.
#'
#' NULL means: not in a batch, or the driver did not set the directory - use the
#' path the chunk asked for.
.batch_rds_path <- function(label) {
  dir <- .batch_opt(".batch_rds_dir")
  smp <- .batch_opt(".batch_sample")
  if (is.null(dir) || is.null(smp)) return(NULL)
  file.path(dir, sprintf("%s_%s.rds", smp,
                         gsub("\\.", "_", gsub("\\s+", "", label))))
}

#' Save a stage checkpoint. Errors on missing required objects - unless the
#' batch driver is running this file, in which case nothing is required, the
#' destination is the per-sample one derived above, and a failure to write is
#' reported rather than thrown. See .in_batch() above.
`%||%` <- function(a, b) if (is.null(a)) b else a

.stage_save <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  required <- unique(required)
  optional <- setdiff(unique(optional), required)

  # A batch may switch an individual checkpoint off by label - the driver's
  # keep_stage3_fit_checkpoint does exactly this for "Stage 3.1", whose chunk
  # must still RUN (the final Stage 3 manifest inherits from it) but need not
  # write a second large file per sample.
  if (!strict && label %in% .batch_opt(".batch_skip_saves", character(0))) {
    cat(sprintf("[%s save] Skipped - switched off for this batch.\n", label))
    return(invisible(NA_character_))
  }

  # inherits = FALSE, matching the mget() below, which has always been strict.
  # With inheritance on, an object could be FOUND in a parent environment and
  # then fail to be FETCHED, and under the driver the parent is globalenv -
  # i.e. the previous pair's leftovers.
  .have <- function(x)
    if (!length(x)) logical(0)
    else vapply(x, exists, logical(1), envir = envir, inherits = FALSE)

  missing_req <- required[!.have(required)]
  if (length(missing_req) > 0L) {
    if (strict)
      stop(sprintf(
        "[%s save] Cannot save - %d required object(s) missing from the environment:\n  %s",
        label, length(missing_req), paste(missing_req, collapse = "\n  ")))
    cat(sprintf(
      "[%s save] %d required object(s) missing - treated as optional and omitted (batch mode):\n  %s\n",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))
  }

  missing_opt <- optional[!.have(optional)]
  if (length(missing_opt) > 0L) {
    msg <- sprintf(
      "[%s save] %d optional object(s) missing - omitted from the checkpoint:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  "))
    if (strict) warning(msg) else cat(msg, "\n", sep = "")
  }

  present  <- c(setdiff(required, missing_req), setdiff(optional, missing_opt))
  n_opt_in <- length(setdiff(optional, missing_opt))
  dat      <- mget(present, envir = envir, inherits = FALSE)

  # ── The checkpoint describes itself ────────────────────────────────────────
  # The manifest travels INSIDE the file. Without it, loading a checkpoint means
  # first reconstructing the exact `required`/`optional` vectors it was written
  # with - and those are built by chunks, so a later checkpoint could only be
  # loaded after running the chunks that define an earlier one's manifest. That
  # made resuming from an arbitrary point in a fresh session impossible: the
  # Stage 3.2 chunk, for instance, stops outright unless .stage3_1_required is
  # already in scope.
  #
  # Stored under a name no manifest can collide with, and stripped again on
  # load, so it never reaches the session as an object.
  dat[[".__manifest__"]] <- list(
    required = setdiff(required, missing_req),
    optional = setdiff(optional, missing_opt),
    label    = label,
    saved_utc = format(Sys.time(), tz = "UTC", usetz = TRUE))

  .write <- function() {
    p <- path
    if (!strict) {
      bp <- .batch_rds_path(label)
      if (!is.null(bp)) p <- bp
    }
    dir.create(dirname(p), recursive = TRUE, showWarnings = FALSE)
    # .unique_path only interactively. A batch path already carries the sample
    # name, so it cannot collide with another sample, and a re-run of the same
    # sample should REPLACE its checkpoint rather than leave _001, _002 behind
    # for the next reader to guess between.
    if (strict) p <- .unique_path(p)
    saveRDS(dat, p)
    cat(sprintf("[%s save] Saved %d object(s) (%d required, %d optional) -> %s\n",
                label, length(dat), length(present) - n_opt_in, n_opt_in, p))
    p
  }

  if (strict) return(invisible(.write()))

  out <- tryCatch(.write(), error = function(e) {
    cat(sprintf("[%s save] Checkpoint NOT written: %s\n", label, conditionMessage(e)))
    cat(sprintf("[%s save] Batch mode - the objects are still in this session, so the run continues.\n",
                label))
    NA_character_
  })
  invisible(out)
}

#' Load a stage checkpoint. Warns (never errors) on content mismatches, and
#' under the batch driver does not read at all - see .in_batch() above.
.stage_load <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  # In a batch the objects this would read are already live in `envir`, put
  # there by the stage that just ran. Reading them back would at best be a
  # no-op. At worst it would be wrong: checkpoint paths are per-sample and
  # .unique_path() means a re-run writes _001 rather than replacing the
  # original, so the path named here can easily be an EARLIER run's file. And
  # if the write failed or was partial, there would be nothing to read and the
  # pair would die at the start of the next stage having done all the work.
  if (!strict) {
    cat(sprintf("[%s load] Skipped - batch mode: this stage's objects are already in the session.\n",
                label))
    return(invisible(character(0)))
  }

  if (!file.exists(path))
    stop(sprintf("[%s load] File not found:\n  %s", label, path))

  dat <- readRDS(path)

  # ── Prefer the manifest the file carries ───────────────────────────────────
  # A checkpoint written by .stage_save records what it holds, so a load needs
  # nothing from the session - no earlier chunk, no earlier checkpoint, no
  # reconstructed name vectors. `required` and `optional` become a CHECK against
  # what the caller expected rather than the definition of what to read, which
  # is what lets any stage be resumed from any checkpoint in a fresh session.
  #
  # Files written before manifests existed simply have none; those fall back to
  # the caller's vectors, exactly as before.
  mf <- dat[[".__manifest__"]]
  dat[[".__manifest__"]] <- NULL
  caller_req <- unique(required)
  caller_opt <- setdiff(unique(optional), caller_req)

  if (!is.null(mf)) {
    required <- unique(mf$required)
    optional <- setdiff(unique(mf$optional), required)
    src <- sprintf("its own manifest (written %s)", mf$saved_utc %||% "?")
  } else {
    required <- caller_req; optional <- caller_opt
    src <- "the caller's manifest - this file predates self-describing checkpoints"
  }
  expected <- c(required, optional)

  cat(sprintf("[%s load] Reading %d object(s) from: %s\n",
              label, length(dat), path))
  cat(sprintf("[%s load] Manifest: %s\n", label, src))

  # Named by the caller but absent from the file: the real incompatibility, and
  # the only one worth a warning when the file describes itself.
  if (!is.null(mf)) {
    unmet <- setdiff(caller_req, names(dat))
    if (length(unmet))
      warning(sprintf(
        "[%s load] %d object(s) this stage expects are not in the checkpoint:\n  %s",
        label, length(unmet), paste(unmet, collapse = "\n  ")))
  }

  missing_req <- required[!required %in% names(dat)]
  if (length(missing_req) > 0L)
    warning(sprintf(
      "[%s load] %d REQUIRED object(s) absent from the RDS - downstream chunks that use them will fail:\n  %s",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))

  missing_opt <- optional[!optional %in% names(dat)]
  if (length(missing_opt) > 0L)
    warning(sprintf(
      "[%s load] %d optional object(s) absent from the RDS:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  ")))

  # Unrequested objects are COUNTED, never listed. Every checkpoint from Stage 2
  # onward is cumulative, so this set is routinely hundreds of names long and
  # printing it buried the two messages that actually matter - an absent
  # required object, and an absent optional one - under a wall of text. The
  # count still appears in the summary line below, which is the only part of it
  # anyone reads.
  extra <- setdiff(names(dat), expected)

  # ── Publish what was loaded, for the next stage's save manifest ────────────
  # Every stage's SAVE manifest is built as "the previous stage's manifest, plus
  # my own objects", and the previous manifest is a vector defined by that
  # stage's LOAD chunk. Load a LATER checkpoint directly and that chunk never
  # runs, so the save dies on a missing .stageN_required - which is exactly the
  # resume-from-anywhere case these checkpoints now exist to support.
  #
  # Recording the manifest here breaks the dependency: .manifest_from() below
  # falls back to whatever was actually loaded, whichever chunk did it.
  assign(".stage_last_manifest",
         list(required = required, optional = optional,
              names = names(dat), label = label, path = path),
         envir = envir)

  list2env(dat, envir = envir)

  cat(sprintf("[%s load] Loaded %d object(s) | required present: %d/%d | optional present: %d/%d | unrequested: %d\n",
              label, length(dat),
              length(required) - length(missing_req), length(required),
              length(optional) - length(missing_opt), length(optional),
              length(extra)))
  invisible(names(dat))
}

#' The manifest a save chunk inherits from the stage before it.
#'
#' Looks for the named vector first - that is the normal path, when this
#' session's own load chunk built it. Falls back to the manifest carried by
#' whichever checkpoint was last loaded, which is what makes it possible to load
#' any checkpoint and still save the next one.
#'
#' `which` is "required" or "optional". Stops with something actionable rather
#' than returning character(0): a save that silently inherits nothing writes a
#' checkpoint missing every object from every earlier stage, and nothing would
#' notice until a later stage failed to find them.
.manifest_from <- function(name, which = c("required", "optional"),
                           envir = .stage_env()) {
  which <- match.arg(which)
  if (exists(name, envir = envir, inherits = FALSE))
    return(get(name, envir = envir, inherits = FALSE))

  lm <- if (exists(".stage_last_manifest", envir = envir, inherits = FALSE))
          get(".stage_last_manifest", envir = envir, inherits = FALSE) else NULL

  if (!is.null(lm)) {
    cat(sprintf("[manifest] %s not in the session - inheriting from the %s checkpoint that was loaded.\n",
                name, lm$label))
    # For "required", everything the loaded file actually held is the honest
    # inheritance: it is what this session has, regardless of how the file that
    # produced it split required from optional.
    return(if (identical(which, "required")) lm$names else character(0))
  }

  # ── Last resort: the session itself ────────────────────────────────────────
  # No named vector and no checkpoint loaded. That is the ordinary case for a
  # session built by RUNNING chunks rather than by loading - Stage 1 through
  # Stage 3 straight down the document, or an interactive session where the
  # objects were produced rather than read. There is nothing wrong with it, and
  # a save must not refuse.
  #
  # So fall back to what Stages 4 and 5 have always done: take the manifest from
  # the environment. Every non-function, non-dotted object in scope is what this
  # stage inherited, by definition - it is the same rule, just applied later.
  #
  # Only "required" gets this. Optional means "save it if it happens to exist",
  # which is meaningless once the list IS everything that exists.
  if (identical(which, "required")) {
    nm <- setdiff(ls(envir = envir), lsf.str(envir = envir))
    nm <- nm[!startsWith(nm, ".")]
    if (length(nm)) {
      cat(sprintf(paste0("[manifest] %s not found and no checkpoint was loaded - taking the manifest\n",
                         "           from the session instead (%d object(s)). This is the normal path\n",
                         "           when the stage was RUN rather than loaded.\n"), name, length(nm)))
      return(nm)
    }
  } else {
    return(character(0))
  }

  stop(sprintf(paste0("[manifest] %s not found, no checkpoint has been loaded, and the session is\n",
                      "  empty - there is nothing to save. Run this stage's chunks, or load a\n",
                      "  checkpoint, before the save chunk."), name))
}

#' Print the Stage 1 pivot point.
.report_pivot1 <- function() {
  if (exists("pivot1")) {
    cat(sprintf("pivot1 (Stage 1 footprint centre): (%.4f, %.4f, %.4f)\n",
                pivot1[1], pivot1[2], pivot1[3]))
  } else {
    cat("pivot1 NOT available.\n")
  }
}

# ============================================================
# Stages 2, 4 and 5 deliberately never load point clouds - they work purely on
# fitted ellipse geometry, which is what keeps them cheap to re-run. A handful
# of 3-D figures in those stages genuinely need raw points, so each of those
# chunks carries its own loader at the top, controlled by one toggle.
#
# Stage 3 carries it too, for a different reason: it DOES load the clouds at the
# top, but releases them again part-way through, and resuming from the Stage 3
# post-fit checkpoint skips the load entirely.
#
# Stages 1 and 6 do NOT have this block at all - they load the clouds up front
# and keep them for the whole stage, so nothing there would ever call it.
#
#   vis_load_clouds <- TRUE    read las_1 / las_2 from disk
#   vis_load_clouds <- FALSE   reuse whatever is already resident in memory
#
# Set it FALSE when the clouds are already loaded - you just ran Stage 3 in the
# same session, or an earlier figure in this stage already loaded them. Reading
# a multi-gigabyte LAS once per figure is slow and pointless. Leave it TRUE when
# running the stage standalone from its checkpoint, which is the normal case.
#
# ── FRAMES: why las_1 needs a rotation the moment it comes off disk ────────────
# las_1 sits on disk in the ORIGINAL scan frame. Stage 1A's rough vertical
# alignment (R_align_1 about align_pivot_1) was applied to las_1 IN MEMORY in
# Stage 1 and never written back - there is no writeLAS anywhere in Stage 1. So
# every measurement made downstream of Stage 1A - ellipse centres, per-stem axis
# vectors, pivot1, and the Stage 2 rotation solve R_xyz - live in the ALIGNED
# frame, not the disk frame. A figure that reads las_1 off disk and overlays any
# of that geometry is mixing two frames unless it first reapplies R_align_1.
#
# This loader therefore reconstructs the pipeline frame for las_1 as a stack of
# rigid steps, matching the Stage 2 composition (R_align_xyz_1) exactly:
#
#   x_aligned = R_align_1 (x_raw     - align_pivot_1) + align_pivot_1   (apply_align, Stage 1A)
#   x_final   = R_xyz     (x_aligned - pivot1)        + pivot1          (apply_rot,   Stage 2)
#   x_final  += t_trans_z                                              (apply_zshift, Stage 4)
#   x_final   = R_z-only  about pivot2 in XY                            (apply_zrot,   Stage 5)
#
# apply_align is the default-ON baseline: everything above stacks on the
# aligned frame, so leaving it on is what makes las_1 line up with the ellipses
# and vectors. Turn it OFF only to inspect the raw disk cloud deliberately.
# las_2 is the REFERENCE cloud - Stage 1 never rotates it and its ellipses are in
# the disk frame - so it is never transformed here.
#
# Transforms are applied to the RETURNED point frame, never baked back into the
# LAS object, so repeated calls (or a load = FALSE reuse) can never compound.
# ============================================================

#' Make point clouds available to one visualisation chunk.
#'
#' @param load         TRUE to read from disk; FALSE to reuse las_1 / las_2
#'                     already in the global environment.
#' @param las_1_path,las_2_path
#'                     Optional explicit source paths. When NULL (default) the
#'                     paths are taken from the global environment, so a caller
#'                     can either pass them or rely on the checkpoint. Only used
#'                     when load = TRUE.
#' @param apply_align  reapply the Stage 1A rough vertical alignment to las_1
#'                     (R_align_1 about align_pivot_1). Default TRUE: this is the
#'                     baseline frame every Stage 1/2 measurement lives in, so
#'                     without it the returned las_1 points do NOT match the
#'                     ellipses or axis vectors. Set FALSE only to see the raw
#'                     on-disk cloud. No-op (identity) when rough alignment was
#'                     not run; skipped with a note if the checkpoint lacks the
#'                     objects.
#' @param apply_rot    rebuild the Stage 2 XYZ rotation R_xyz (about pivot1),
#'                     stacked on the aligned frame.
#' @param apply_zshift rebuild the Stage 4 Z-translation.
#' @param apply_zrot   rebuild the Stage 5 Z-only rotation (about pivot2 in XY).
#' @param need_las2    FALSE to skip the reference cloud entirely.
#'
#' @return list(las_1, las_2, pts_1, pts_2, xyz_1, state). pts_* are plain data
#'   frames of X/Y/Z/TreeID with TreeID 0 and NA dropped (figures show stems).
#'   xyz_1 / pts_1 are in the transformed frame; the LAS objects stay on-disk.
.vis_point_clouds <- function(load,
                              las_1_path   = NULL,
                              las_2_path   = NULL,
                              apply_align  = TRUE,
                              apply_rot    = FALSE,
                              apply_zshift = FALSE,
                              apply_zrot   = FALSE,
                              need_las2    = TRUE,
                              quiet        = FALSE,
                              envir        = .stage_env()) {

  .say <- function(...) if (!quiet) cat(...)
  .g   <- function(nm) get(nm, envir = envir)
  .has <- function(nm) exists(nm, envir = envir)

  # Resolve source paths: explicit argument wins, else the global/checkpoint.
  .p1 <- if (!is.null(las_1_path)) las_1_path else
         if (.has("las_1_path")) .g("las_1_path") else NULL
  .p2 <- if (!is.null(las_2_path)) las_2_path else
         if (.has("las_2_path")) .g("las_2_path") else NULL

  if (isTRUE(load)) {

    if (is.null(.p1))
      stop("[vis loader] las_1_path not supplied and not in the environment - ",
           "pass it or load this stage's checkpoint first.")

    .say(sprintf("[vis loader] Reading las_1 from disk: %s\n", basename(.p1)))
    l1 <- lidR::readLAS(.p1)
    if (lidR::is.empty(l1)) stop("[vis loader] las_1 read as empty: ", .p1)
    l1 <- lidR::add_lasattribute(l1, x = seq_len(nrow(l1@data)),
                                 name = "point_id", desc = "Unique ID for las_1 points")
    l2 <- NULL
    if (need_las2) {
      if (is.null(.p2))
        stop("[vis loader] las_2_path not supplied and not in the environment.")
      .say(sprintf("[vis loader] Reading las_2 from disk: %s\n", basename(.p2)))
      l2 <- lidR::readLAS(.p2)
      if (lidR::is.empty(l2)) stop("[vis loader] las_2 read as empty: ", .p2)
      l2 <- lidR::add_lasattribute(l2, x = seq_len(nrow(l2@data)),
                                   name = "point_id", desc = "Unique ID for las_2 points")
    }

    assign("las_1", l1, envir = envir)
    if (!is.null(l2)) assign("las_2", l2, envir = envir)

  } else {

    if (!.has("las_1"))
      stop("[vis loader] vis_load_clouds = FALSE but las_1 is not in memory. Set it TRUE.")
    if (need_las2 && !.has("las_2"))
      stop("[vis loader] vis_load_clouds = FALSE but las_2 is not in memory. Set it TRUE.")

    l1 <- .g("las_1")
    l2 <- if (need_las2) .g("las_2") else NULL
    .say("[vis loader] Reusing the point clouds already in memory (no disk read).\n")
  }

  # ── Rebuild the transform state (las_1 only; las_2 is never transformed) ──────
  # The stack runs in pipeline order on the RETURNED frame: Stage 1A align first,
  # then the Stage 2 rotation, then the Stage 4/5 shifts. Each step is a rigid
  # transform about the pivot the corresponding stage used, matching the Stage 2
  # composition. IDENTICAL in Stages 2, 3, 4 and 5 - if it is edited, edit all four.
  xyz   <- as.matrix(l1@data[, c("X", "Y", "Z")])
  state <- character(0)

  # ── Stage 1A: rough vertical alignment (raw disk -> aligned frame) ────────────
  if (isTRUE(apply_align)) {
    if (!.has("R_align_1") || !.has("align_pivot_1")) {
      .say("[vis loader] R_align_1 / align_pivot_1 not in the checkpoint - ",
           "las_1 left in the raw disk frame (no Stage 1A alignment).\n")
    } else {
      Ra <- .g("R_align_1"); pa <- as.numeric(.g("align_pivot_1"))
      xyz <- sweep(t(Ra %*% t(sweep(xyz, 2, pa))), 2, -pa)
      # Angle of the rigid rotation, for the log only: acos((tr(R) - 1) / 2).
      .ang_a <- acos(max(-1, min(1, (sum(diag(Ra)) - 1) / 2))) * 180 / pi
      state  <- c(state, sprintf("stage-1A align %.4f deg about align_pivot_1", .ang_a))
    }
  }

  # ── Stage 2: XYZ rotation R_xyz about pivot1 (aligned -> final) ───────────────
  if (isTRUE(apply_rot)) {
    if (!.has("R_xyz") || !.has("pivot1"))
      stop("[vis loader] apply_rot = TRUE but R_xyz / pivot1 not found - run the ",
           "Stage 2 rotation solve first, or reload a checkpoint that carries them.")
    R  <- .g("R_xyz"); pv <- as.numeric(.g("pivot1"))
    xyz <- sweep(t(R %*% t(sweep(xyz, 2, pv))), 2, -pv)
    .ang_x <- if (.has("angle_xyz")) .g("angle_xyz") else
              acos(max(-1, min(1, (sum(diag(R)) - 1) / 2))) * 180 / pi
    state <- c(state, sprintf("xyz-rot %.4f deg about pivot1", .ang_x))
  }

  if (isTRUE(apply_zshift)) {
    tz <- if (.has("t_trans_z_applied")) .g("t_trans_z_applied") else .g("t_trans_z")
    xyz[, 3] <- xyz[, 3] + tz
    state <- c(state, sprintf("z-shift %+.4f m", tz))
  }

  if (isTRUE(apply_zrot)) {
    # Same sign flip the Stage 5 / Stage 6 builders use: z_rotation_deg is held
    # in the standard CCW-positive convention, the 2x2 fill below evaluates to
    # the matrix for -a, so the angle is negated exactly once here.
    a  <- -.g("z_rotation_deg") * pi / 180
    Rz <- matrix(c(cos(a), -sin(a), sin(a), cos(a)), 2, 2)
    pv <- if (.has("pivot2")) .g("pivot2") else .g("pivot1")
    xc <- xyz[, 1] - pv[1]; yc <- xyz[, 2] - pv[2]
    xyz[, 1] <- Rz[1, 1] * xc + Rz[1, 2] * yc + pv[1]
    xyz[, 2] <- Rz[2, 1] * xc + Rz[2, 2] * yc + pv[2]
    state <- c(state, sprintf("z-rot %+.4f deg", .g("z_rotation_deg")))
  }

  pts_1 <- data.frame(X = xyz[, 1], Y = xyz[, 2], Z = xyz[, 3], TreeID = l1@data$TreeID)
  pts_1 <- pts_1[!is.na(pts_1$TreeID) & pts_1$TreeID != 0L, ]

  pts_2 <- NULL
  if (!is.null(l2)) {
    pts_2 <- data.frame(X = l2@data$X, Y = l2@data$Y, Z = l2@data$Z, TreeID = l2@data$TreeID)
    pts_2 <- pts_2[!is.na(pts_2$TreeID) & pts_2$TreeID != 0L, ]
  }

  .say(sprintf("[vis loader] las_1 state: %s\n",
               if (length(state)) paste(state, collapse = " + ") else "as loaded (no transform)"))
  .say(sprintf("[vis loader] stem points available - las_1: %s | las_2: %s\n",
               format(nrow(pts_1), big.mark = ","),
               if (is.null(pts_2)) "not loaded" else format(nrow(pts_2), big.mark = ",")))

  invisible(list(las_1 = l1, las_2 = l2, pts_1 = pts_1, pts_2 = pts_2,
                 xyz_1 = xyz, state = state))
}

# ============================================================
# Run tally - one cumulative record of how many things survived each filter
#
# Every stage prints its own counts, but those printouts vanish with the render
# and cannot be compared across runs. .tally() writes the same numbers into a
# single named list, `run_counts`, that travels in the checkpoints and is read
# out whole by the Stage 6 error report. A run then answers "how many pairs
# reached the Z-translation" without anyone re-reading five HTML files.
#
# Cumulative and idempotent: each call updates only the keys it names, so
# re-running one chunk refreshes that chunk's numbers and leaves the rest alone.
# It never errors - a tally is bookkeeping, and a missing count must not be able
# to stop a stage from saving.
# ============================================================
.tally <- function(...) {
  vals <- list(...)
  e    <- .stage_env()
  base <- if (exists("run_counts", envir = e, inherits = FALSE))
            get("run_counts", envir = e) else list()
  assign("run_counts", utils::modifyList(base, vals), envir = e)
  invisible(get("run_counts", envir = e))
}


#' TRUE when a cli bar can actually be seen.
#'
#' All four conditions matter: cli absent, output being knitted, a message sink
#' in the way (the batch driver), or a terminal that cannot redraw - any one of
#' them and the bar is invisible, mangled, or both.
.pb_cli_usable <- function() {
  isNamespaceLoaded("cli") || requireNamespace("cli", quietly = TRUE)
  requireNamespace("cli", quietly = TRUE) &&
    !isTRUE(getOption("knitr.in.progress")) &&
    sink.number(type = "message") == 0L &&
    isTRUE(try(cli::is_dynamic_tty(), silent = TRUE))
}

.pb_fmt_dur <- function(s) {
  if (!is.finite(s)) return("--")
  if (s < 90) sprintf("%.0fs", s) else if (s < 5400) sprintf("%.1fm", s / 60)
  else sprintf("%.1fh", s / 3600)
}

#' Start a bar. Returns an opaque handle for .pb_tick()/.pb_done(), or NULL.
#'
#' .envir = parent.frame() is load-bearing for the cli branch: cli ties a bar's
#' lifetime to a frame, so a bar created with the default .envir would be torn
#' down the instant this function returned and every later tick would silently
#' do nothing. That was a real bug here once.
.pb_new <- function(total, label) {
  total <- as.integer(total)
  if (!isTRUE(show_progress_bars) || is.na(total) || total < 1L) return(NULL)

  if (.pb_cli_usable()) {
    id <- cli::cli_progress_bar(
      name   = label,
      total  = total,
      format = paste0(label,
                      " {cli::pb_bar} {cli::pb_current}/{cli::pb_total}",
                      " | {cli::pb_elapsed} | {cli::pb_eta} left"),
      .envir = parent.frame())
    return(structure(list(kind = "cli", id = id), class = ".pb"))
  }

  # An ENVIRONMENT, not a list: a tick inside lapply() has to update the
  # caller's counter, and a list passed by value would not.
  st <- new.env(parent = emptyenv())
  st$i <- 0L; st$total <- total; st$label <- label
  st$t0 <- proc.time()[["elapsed"]]; st$last_t <- -Inf; st$last_i <- -1L
  structure(list(kind = "lines", st = st), class = ".pb")
}

.pb_tick <- function(pb, by = 1L) {
  if (is.null(pb)) return(invisible(NULL))
  if (identical(pb$kind, "cli")) {
    cli::cli_progress_update(id = pb$id, inc = by)
    return(invisible(NULL))
  }
  st <- pb$st
  st$i <- st$i + by
  now <- proc.time()[["elapsed"]]
  step <- max(1L, as.integer(ceiling(st$total * progress_step_frac)))
  due  <- st$i >= st$total || st$i == 1L ||
          (st$i - st$last_i) >= step ||
          (now - st$last_t) >= progress_min_seconds
  if (!due) return(invisible(NULL))
  st$last_i <- st$i; st$last_t <- now
  el   <- now - st$t0
  frac <- st$i / st$total
  eta  <- if (st$i > 0L) el / st$i * (st$total - st$i) else NA_real_
  nfill <- as.integer(round(20 * frac))
  cat(sprintf("  %s [%s%s] %d/%d %3.0f%% | %s elapsed | ~%s left\n",
              st$label, strrep("#", nfill), strrep("-", 20L - nfill),
              st$i, st$total, 100 * frac, .pb_fmt_dur(el), .pb_fmt_dur(eta)))
  flush.console()
  invisible(NULL)
}

.pb_done <- function(pb) {
  if (is.null(pb)) return(invisible(NULL))
  if (identical(pb$kind, "cli")) {
    cli::cli_progress_done(id = pb$id)
  } else if (pb$st$i < pb$st$total) {
    # The loop ended early (an epoch with nothing to do, say). Say so rather
    # than leaving a bar stopped at 12/25 as though something had hung.
    cat(sprintf("  %s: finished after %d of %d | %s\n", pb$st$label, pb$st$i,
                pb$st$total, .pb_fmt_dur(proc.time()[["elapsed"]] - pb$st$t0)))
  }
  invisible(NULL)
}
```

## 0.3 Load the Stage 4 Checkpoint

```{r}
#| label: params-load_Stage5
#| code-summary: "Parameters - Load the Stage 4 checkpoint (no point clouds are read in this stage)"


# Only the Stage 4 checkpoint is read here. The Stage 5 OUTPUT path
# (stage5_rds_path) is defined in the Stage 5 save chunk, where it is used.
stage4_rds_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/RunR/exports/SaveLoad/I_SNBG_F1_ai_134p77_2020_2024_TLS_TLS_Stage4.rds"

.par_apply("params-load_Stage5")
```

```{r}
#| label: load_Stage5
#| code-summary: "Load the Stage 4 checkpoint (no point clouds are read in this stage)"
#| code-fold: false

# Stage 5 uses only a handful of Stage 4 objects. .stage_load pulls EVERYTHING in
# the RDS via list2env regardless of this list (the manifest only drives the
# present/absent warnings), so rather than mirror Stage 4's full manifest here -
# which drifts the moment an upstream stage changes - Stage 5 requests just what
# the XY-translation solve consumes and lets the loader bring the rest in
# silently; the loader counts the unrequested objects rather than listing them.
.stage4_required <- c(
  # geometry / frame the solve works in
  "pivot1", "pivot2",
  # the Z-shift already applied to the centres upstream (context / reporting)
  "t_trans_z_applied",
  # the Z-only rotation SOLVED in Stage 4 but NOT yet applied - Stage 5 applies
  # it to the las_1 centres before measuring the residual XY offset
  "z_rotation_deg", "z_rotation_rad",
  # windowed average ellipse centres per epoch + the valid pair set
  "avg_centres_1", "avg_centres_2", "pairs_zrot_valid",
  # stems the Stage 4 angular-difference MAD filter rejected (excluded from XY)
  "removed_trees_zr",
  # pair set carried through Stage 4 (reporting)
  "pairs_stage4"
)
.stage4_optional <- c(
  # useful context, absent on some runs
  "s5_signal_registry", "rotation_summary_zr",
  "ellipses_zrot_1", "ellipses_zrot_2",
  "las_1_path", "las_2_path"
)

.stage_load(stage4_rds_path, .stage4_required, .stage4_optional, label = "Stage 4")

cat("\n── Carried in from Stages 1-4 ─────────────────────────────────────────────\n")
if (exists(".report_pivot1")) .report_pivot1()
cat(sprintf("pivot2 (post Z-shift): (%.4f, %.4f, %.4f)\n", pivot2[1], pivot2[2], pivot2[3]))
cat(sprintf("Z-shift already applied to the ellipse centres: %+.4f m\n", t_trans_z_applied))
cat(sprintf("Z-only rotation solved upstream (applied in 5.1): %+.4f deg\n", z_rotation_deg))
cat(sprintf("Pairs available (pairs_stage4): %d\n", nrow(pairs_stage4)))
cat("No point-cloud data is loaded in this stage.\n")
```

# Stage 5 XY-translation


## 5.1 Assemble the Rotated Centres for the XY Solve

```{r}
#| label: assemble-centres-xyt
#| code-summary: "Restrict the rotated centres to pairs that survived every Stage 5 filter"

# ─────────────────────────────────────────────────────────────────────────────
# INPUT: pairs_zrot_valid (Stage 4: pairs with a valid windowed centre both sides)
# avg_centres_1 (Stage 4: las_1 windowed mean centres, PRE-rotation)
# avg_centres_2 (Stage 4: las_2 windowed mean centres, reference)
# z_rotation_deg (Stage 4: the Z-only yaw, SOLVED but not yet applied)
# pivot2 (Stage 4: pivot in the post-Z-shift frame)
# removed_trees_zr (Stage 4: stems the angular-diff MAD filter rejected)
# OUTPUT: pairs_xyt, avg_centres_xyt_1, avg_centres_xyt_2
# ─────────────────────────────────────────────────────────────────────────────
# ARCHAIC-DEPENDENCY FIX. The old Stage 4 pre-applied the Z-rotation to the las_1
# centres and handed Stage 5 an already-rotated `avg_centres_1_zrot` plus a
# `pairs_centres_ok` set. The new Stage 4 follows the solve-then-apply split: it
# SOLVES the yaw (z_rotation_deg) but records rather than applies it, and its
# valid-pair object is `pairs_zrot_valid`. So Stage 5 now (a) reads the new
# names, and (b) APPLIES the Z-rotation to avg_centres_1 here - about pivot2 in
# XY, with the exact sign convention the cloud builders use (a = -z_rotation_deg)
# - to produce the rotated las_1 centres the XY-displacement solve needs.

.have <- function(nm) exists(nm, envir = .stage_env(), inherits = FALSE)
# exists() is told WHERE to look rather than relying on its default. Bare, it
# resolves from the frame of whatever forced it - and here that is inside
# stopifnot(), not this chunk. Under the batch driver, where chunks run in a
# per-pair environment rather than globalenv, a lookup that escapes to the
# global search path reports these six as absent however plainly they are here.
# Stage 3's load chunk failed exactly that way inside a vapply().
stopifnot(
  .have("pairs_zrot_valid"),
  .have("avg_centres_1"),
  .have("avg_centres_2"),
  .have("z_rotation_deg"),
  .have("pivot2"),
  .have("removed_trees_zr")
)

# --- Apply the solved Z-only rotation to the las_1 centres (about pivot2, XY) ---
# ang_diff = angle_2 - angle_1 is the rotation carrying las_1 TO las_2, so aligning
# las_1 onto las_2 applies +z_rotation_deg (NOT -). z_rotation_deg is in degrees;
# convert to radians for cos/sin. las_2 is the fixed reference and is never rotated.
.a_zrot <- z_rotation_deg * pi / 180        # was -z_rotation_deg (wrong direction)
.Rz     <- matrix(c(cos(.a_zrot), -sin(.a_zrot),
                    sin(.a_zrot),  cos(.a_zrot)), nrow = 2, byrow = TRUE)

avg_centres_1_zrot <- avg_centres_1 |>
  dplyr::mutate(
    .xc = cx - pivot2[1],
    .yc = cy - pivot2[2],
    cx  = .Rz[1, 1] * .xc + .Rz[1, 2] * .yc + pivot2[1],
    cy  = .Rz[2, 1] * .xc + .Rz[2, 2] * .yc + pivot2[2]
  ) |>
  dplyr::select(-.xc, -.yc)

# --- Restrict to pairs that survived every Stage 4 filter -----------------------
# Start from the valid-centre pair set, then exclude the trees the Stage 4
# angular-difference MAD filter rejected: a stem whose lines disagreed with the
# network consensus is exactly the stem whose centre is least trustworthy for
# measuring a translation.
pairs_xyt <- pairs_zrot_valid |>
  dplyr::filter(
    !as.character(TreeID_1) %in% removed_trees_zr,
    !as.character(TreeID_2) %in% removed_trees_zr
  )

avg_centres_xyt_1 <- avg_centres_1_zrot |>
  dplyr::filter(TreeID %in% pairs_xyt$TreeID_1)
avg_centres_xyt_2 <- avg_centres_2 |>
  dplyr::filter(TreeID %in% pairs_xyt$TreeID_2)

# Final: both epochs must have a valid mean centre
pairs_xyt <- pairs_xyt |>
  dplyr::filter(
    TreeID_1 %in% avg_centres_xyt_1$TreeID,
    TreeID_2 %in% avg_centres_xyt_2$TreeID
  )

cat(sprintf("Pairs with valid windowed centres (pairs_zrot_valid): %d\n", nrow(pairs_zrot_valid)))
cat(sprintf("Z-rotation applied to las_1 centres:                  %+.4f deg (about pivot2)\n", z_rotation_deg))
cat(sprintf("Pairs after the Stage 4 angular-diff MAD exclusion:   %d\n", nrow(pairs_xyt)))
cat(sprintf("Tracked centres - las_1: %d | las_2: %d tree(s)\n",
            nrow(avg_centres_xyt_1), nrow(avg_centres_xyt_2)))

if (nrow(pairs_xyt) < 1L)
  stop("No pairs available for the XY translation solve in Stage 5.")

```


### Vis - 3D stem slices after XYZ rotation, Z-shift and Z-only rotation

```{r}
#| label: params-inspect-pairs-post-zrotation-3D
#| code-summary: "Parameters - inspect-pairs-post-zrotation-3D"


# ── Point clouds for this figure ──────────────────────────────────────────────
# This stage does not load point clouds at the top, so this figure loads its own.
# vis_load_clouds <- TRUE read las_1 / las_2 from disk and apply the
# Stage 1A alignment + every solved transform reapplied
# vis_load_clouds <- FALSE reuse the clouds already in memory (no disk read)
# The loader also rebuilds the transform state this figure expects, so the
# points shown match the pipeline stage rather than the file on disk.
vis_load_clouds <- TRUE

.par_apply("params-inspect-pairs-post-zrotation-3D")
```

```{r}
#| label: inspect-pairs-post-zrotation-3D
#| fig-cap: "3D inspection of temporal tree pairs after Z-translation, full XYZ rotation, and Z-only rotation."

.vpc <- .vis_point_clouds(
  load = vis_load_clouds,
  apply_rot     = TRUE,
  apply_zshift  = TRUE,
  apply_zrot    = TRUE
)

las1_pts_zrot <- .vpc$pts_1   # las_1: all three transforms rebuilt

#| eval: false
if (params$make_plots) {

  inspect_z_min   <- 1
  inspect_z_max   <- 3
  points_per_tree <- 1000

  las2_pts_zrot_vis <- data.frame(
    X      = las_2@data$X,
    Y      = las_2@data$Y,
    Z      = las_2@data$Z,
    TreeID = las_2@data$TreeID
  )
  las2_pts_zrot_vis <- las2_pts_zrot_vis[
    !is.na(las2_pts_zrot_vis$TreeID) & las2_pts_zrot_vis$TreeID != 0L,
  ]

  extract_inspect_slice_df <- function(df, tree_ids) {
    df <- df[df$TreeID %in% tree_ids, ]
    df |>
      dplyr::group_by(TreeID) |>
      dplyr::mutate(Z_rel = Z - min(Z)) |>
      dplyr::filter(Z_rel >= inspect_z_min, Z_rel <= inspect_z_max) |>
      dplyr::ungroup()
  }

  inspect_1 <- extract_inspect_slice_df(las1_pts_zrot,     pairs_zrot_valid$TreeID_1) |>
    dplyr::mutate(cloud = "las_1 (tilt + Z-rot + Z-trans)")
  inspect_2 <- extract_inspect_slice_df(las2_pts_zrot_vis, pairs_zrot_valid$TreeID_2) |>
    dplyr::mutate(cloud = "las_2")

  inspect_df <- dplyr::bind_rows(inspect_1, inspect_2) |>
    dplyr::group_by(cloud, TreeID) |>
    dplyr::group_modify(~ {
      n_pts  <- nrow(.x)
      n_take <- min(points_per_tree, n_pts)
      .x[sample.int(n_pts, n_take, replace = FALSE), ]
    }) |>
    dplyr::ungroup()

  cat("Inspection points:", nrow(inspect_df), "\n")

  cx <- mean(inspect_df$X)
  cy <- mean(inspect_df$Y)
  inspect_df <- inspect_df |>
    dplyr::mutate(X_local = X - cx, Y_local = Y - cy)

  plot_ly(
    data = inspect_df,
    x = ~X_local, y = ~Y_local, z = ~Z,
    color = ~cloud,
    colors = c(
      "las_1 (tilt + Z-rot + Z-trans)" = COL_1,
      "las_2"                           = COL_2
    ),
    type = "scatter3d", mode = "markers",
    marker = list(size = 2, opacity = 0.6),
    text = ~paste0("TreeID: ", TreeID, "<br>Cloud: ", cloud),
    hoverinfo = "text"
  ) |>
    layout(
      title = "Temporal pairs after Z-translation + full XYZ rotation + Z-only rotation (1–3 m slice)",
      scene = list(
        xaxis = list(title = "X local (m)"),
        yaxis = list(title = "Y local (m)"),
        zaxis = list(title = "Z (m)"),
        aspectmode = "data"
      )
    )
}
```

## 5.2 XY Displacement and Outlier Filter
```{r}
#| label: params-stem-centres-xyt-comparison
#| code-summary: "Parameters - Tracked stem centres: compute per-pair XY displacement, MAD outlier removal on dx/dy"


xyt_mad_k <- 4

# ─────────────────────────────────────────────────────────────────────────────
# INPUT: avg_centres_xyt_1, avg_centres_xyt_2, pairs_xyt (from 5.1)
# OUTPUT: comp, comp_all, stem_centres_xyt
# ─────────────────────────────────────────────────────────────────────────────

# xyt_mad_k is set immediately above.

# --- Compute per-pair XY displacement ---
# avg_centres_xyt_1: las_1 centres with every transform tracked through so far
# (XYZ rotation -> Z-translation -> Z-only rotation)
# avg_centres_xyt_2: las_2 centres – the fixed reference epoch
# Whatever XY offset remains between them is the residual translation.

.par_apply("params-stem-centres-xyt-comparison")
```

```{r}
#| label: stem-centres-xyt-comparison
#| code-summary: "Tracked stem centres: compute per-pair XY displacement, MAD outlier removal on dx/dy"

comp <- avg_centres_xyt_1 |>

  dplyr::left_join(
    pairs_xyt |>
      dplyr::left_join(avg_centres_xyt_2 |> dplyr::rename(cx2 = cx, cy2 = cy),
                       by = c("TreeID_2" = "TreeID")),
    by = c("TreeID" = "TreeID_1")
  ) |>
  dplyr::filter(!is.na(cx), !is.na(cx2)) |>
  dplyr::mutate(
    dx_xyt = cx2 - cx,
    dy_xyt = cy2 - cy
  ) |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4)))

# --- MAD outlier removal on dx_xyt and dy_xyt ---
mad_remove_displacement <- function(df, cols = c("dx_xyt", "dy_xyt"), k) {
  keep <- rep(TRUE, nrow(df))
  for (col in cols) {
    v    <- df[[col]]
    med  <- median(v, na.rm = TRUE)
    madv <- mad(v, na.rm = TRUE, constant = 1)
    if (madv < 1e-10) next   # all values identical – keep all
    keep <- keep & abs(v - med) <= k * madv
  }
  df[keep, ]
}

comp_all <- comp                                          # full table for display
comp     <- mad_remove_displacement(comp, k = xyt_mad_k) # filtered table for translation

stem_centres_xyt <- comp

cat(sprintf("MAD multiplier (xyt_mad_k): %.1f\n", xyt_mad_k))
cat(sprintf("Pairs before MAD filter: %d | after: %d\n", nrow(comp_all), nrow(comp)))

cat("\nAll pairs (before MAD filter):\n")
comp_all |>
  dplyr::select(TreeID, TreeID_2,
                cx_las1 = cx, cy_las1 = cy,
                cx_las2 = cx2, cy_las2 = cy2,
                dx_xyt, dy_xyt) |>
  print()

cat("\nPairs used for translation (after MAD filter):\n")
comp |>
  dplyr::select(TreeID, TreeID_2,
                cx_las1 = cx, cy_las1 = cy,
                cx_las2 = cx2, cy_las2 = cy2,
                dx_xyt, dy_xyt) |>
  print()

# The pairs that survived every Stage 5 filter AND the final dx/dy MAD cut are
# exactly the ones that contributed to t_trans_xy. Recording them under their
# own name keeps the provenance of the reported translation explicit.

pairs_stage5 <- pairs_xyt |>
  dplyr::semi_join(
    comp |> dplyr::select(TreeID_1 = TreeID, TreeID_2),
    by = c("TreeID_1", "TreeID_2")
  )

cat(sprintf("pairs_stage5: %d pair(s) contributed to the final XY translation.\n",
            nrow(pairs_stage5)))
.tally(n_pairs_stage5_in = nrow(pairs_xyt),
       n_pairs_xy_kept   = nrow(pairs_stage5))

# Provenance chain, robust to which pair objects this pipeline build carries.
# The current Stage 3 ends at pairs_ax_fitted (the old pairs_stage3 / manual-
# removal step was retired), so fall back to that name when pairs_stage3 is
# absent; likewise guard the earlier stages.
.chain_n <- function(nm, alt = NULL) {
  # Explicit envir: this runs inside a function, so a bare exists()/get() would
  # search the function frame and then whatever lies beyond it. It happens to
  # reach the right place here, but only by way of the enclosure - and a stale
  # object of the same name in globalenv would satisfy it just as well.
  .e <- .stage_env()
  if (exists(nm, envir = .e, inherits = FALSE)) nrow(get(nm, envir = .e))
  else if (!is.null(alt) && exists(alt, envir = .e, inherits = FALSE)) nrow(get(alt, envir = .e))
  else NA_integer_
}
.n1 <- .chain_n("pairs_stage1")
.n2 <- .chain_n("pairs_stage2")
.n3 <- .chain_n("pairs_stage3", alt = "pairs_ax_fitted")
.n4 <- .chain_n("pairs_stage4")
.n5 <- nrow(pairs_stage5)
cat(sprintf("  Chain: stage1 %s -> stage2 %s -> stage3 %s -> stage4 %s -> stage5 %d\n",
            .n1, .n2, .n3, .n4, .n5))



```

## 5.3 Compute the XY Translation

```{r}
#| label: compute-xy-translation-new
#| code-summary: "Compute XY translation from tracked stem centres (unweighted mean)"
# ─────────────────────────────────────────────────────────────────────────────
# INPUT: comp, stem_centres_xyt (from 5.2)
# OUTPUT: t_trans_xy
# ─────────────────────────────────────────────────────────────────────────────

# --- Compute XY translation (unweighted mean) ---
t_trans_xy <- c(
  mean(comp$dx_xyt, na.rm = TRUE),
  mean(comp$dy_xyt, na.rm = TRUE)
)

# --- Diagnostics ---
cat("Using unweighted mean XY translation\n")
cat("\nPer-pair XY offsets:\n")
print(
  data.frame(
    TreeID   = comp$TreeID,
    TreeID_2 = comp$TreeID_2,
    dx_xyt   = comp$dx_xyt,
    dy_xyt   = comp$dy_xyt
  ),
  row.names = FALSE
)

cat(sprintf("\nXY translation (t_trans_xy) from tracked stem centres:\n"))
cat(sprintf("  t_trans_xy[1]  (x-shift applied to las_1) : %+.5f m  (%+.2f cm)\n",
            t_trans_xy[1], t_trans_xy[1] * 100))
cat(sprintf("  t_trans_xy[2]  (y-shift applied to las_1) : %+.5f m  (%+.2f cm)\n",
            t_trans_xy[2], t_trans_xy[2] * 100))
cat(sprintf("  Based on %d pairs.\n", nrow(stem_centres_xyt)))

if (nrow(stem_centres_xyt) < 3L) {
  warning(paste0(
    "Fewer than 3 pairs contribute to t_trans_xy. ",
    "Treat the result with caution."
  ))
}
```


## Save/Load – Stage 5

```{r}
#| label: params-save-load_Stage5
#| code-summary: "Parameters - Save or load the complete Stage 5 object set"


# The Stage 5 OUTPUT path lives here (the load chunk reads only the Stage 4 path).
stage5_rds_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/SaveLoad/Stage5_BOS_F1_V1S4_P1S3_4mS2_15mS1.rds"

stage5_save <- TRUE
stage5_load <- FALSE

.par_apply("params-save-load_Stage5")
```

```{r}
#| label: save-load_Stage5
#| code-summary: "Save or load the complete Stage 5 object set"

# A batch decides these itself, whatever they are left set to above. Both are
# read a few lines below, in this same chunk, so the driver's overrides land
# too late to reach them - and a stage left switched to LOAD after an
# interactive session would otherwise make a batch write no checkpoint at all
# and say nothing about it. Saving is what a batch is for; loading is a no-op
# in one, since every stage runs in the same session.
if (.in_batch()) { stage5_save <- TRUE; stage5_load <- FALSE }

# Stage 5 checkpoint = everything the Stage 4 RDS carried (still in the
# environment after the load chunk) + Stage 5's own products. The old manifest
# hand-listed ~300 Stage 1-4 names (many long retired) AND objects from when
# Stage 5 also did the Z-window / line-network / Z-rotation solve - all of which
# moved to Stage 4. Instead, snapshot the current environment's data objects
# (dropping functions and internal dot-names) and add only Stage 5's new names,
# so this can never drift from upstream again.
.stage5_new <- c(
  # 5.1 rotated las_1 centres + the XY-solve pair/centre sets
  "avg_centres_1_zrot", "pairs_xyt", "avg_centres_xyt_1", "avg_centres_xyt_2",
  # 5.2 per-pair XY displacement + MAD filter
  "xyt_mad_k", "comp", "comp_all", "stem_centres_xyt", "pairs_stage5",
  "run_counts",
  # 5.3 the solved XY translation - the stage's headline output
  "t_trans_xy"
)

# carry forward whatever came in from Stage 4 (data objects only, no helpers)
.carried <- setdiff(ls(envir = .stage_env()), lsf.str(envir = .stage_env()))
.carried <- .carried[!startsWith(.carried, ".")]

.stage5_required <- unique(c(.carried, .stage5_new))
.stage5_optional <- character(0)

if (stage5_save) {
  .stage_save(stage5_rds_path, .stage5_required, .stage5_optional, label = "Stage 5")
}

if (stage5_load) {
  .stage_load(stage5_rds_path, .stage5_required, .stage5_optional, label = "Stage 5")
  cat(sprintf("\nStage 5 restored: Z-rotation %+.4f deg | XY (%+.5f, %+.5f) m | %d pair(s)\n",
              z_rotation_deg, t_trans_xy[1], t_trans_xy[2], nrow(pairs_stage5)))
  if (exists(".report_pivot1")) .report_pivot1()
}
```

Next step

Run `Stage6_Export.qmd`. It reloads the point clouds, composes the Z-only
rotation with the XYZ rotation, applies the result about `pivot1`, adds the XYZ
translation, and writes the aligned cloud.
