---
title: "Forest Point Cloud Alignment – Stage 4"
subtitle: "Solve the Z-only rotation from ellipse-centre line networks"
author: "Point Cloud Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: flatly
    highlight-style: github
    fig-width: 9
    fig-height: 6
    self-contained: true
execute:
  warning: false
  message: false
  eval: false
params:
  make_plots: true
---


# Overview – Stage 4 of 6

Stage 4 **solves the Z-only (yaw) rotation** that best aligns the two epochs in the horizontal plane. It first applies the vertical offset `t_trans_z` that Stage 3 solved but left unapplied – bringing both epochs into one shared absolute-height frame – and then estimates the yaw from the geometry of the ellipse-centre network.

The setup (Steps 4.1-4.3) applies `t_trans_z` to:

1. **the ellipse centres** fitted in Stage 3 – every absolute-height field of the `las_1` ellipses moves by `t_trans_z`, producing `ellipses_zshift_1` / `ellipses_zshift_2`;
2. **`pivot1`** – producing **`pivot2`**, the pivot in the post-Z-shift frame.

The solve (Steps 4.5-4.9) then works purely on the shifted ellipse centres:

3. **average each tree’s ellipse centres** to one representative XY point per stem, per epoch (4.6);
4. **build an inter-tree line network** – every pair of stems defines an edge – and measure, for each edge, the angular difference between the epoch-1 and epoch-2 versions of that edge (4.7). A pure yaw rotates every edge by the same angle, so those per-edge differences all estimate the same rotation;
5. **remove outlier stems** whose edges disagree with the consensus, iteratively by MAD (4.8);
6. **solve the rotation** as the mean of the surviving edge-wise differences (4.9), recorded as `z_rotation_deg`.

No point clouds are loaded for the solve. The rotation is solved and recorded here but **applied** in Stage 5/6, following the same solve-then-apply split Stage 3 used for the translation. `z_rotation_deg` is the stage’s headline output.

Which fields move, and which deliberately do not

Stage 3's ellipses carry heights in two different frames, and only one of them
is affected by a rigid vertical translation:

- **Absolute (world) heights** – `center_z`, `global_stem_origin_z`,
  `stem_len_origin`, `stem_len_abs`. These are real elevations in the shared
  frame, so every one of them gains `t_trans_z`.
- **Stem-relative axis coordinates** – `stem_len` and its alias `slice_z`.
  These measure distance *along that stem's own axis from that stem's own
  origin*. Translating the whole stem vertically moves the origin by exactly
  the same amount, so the relative coordinate is unchanged. Shifting it as well
  would double-count the offset.

`slice_z_global` is (re)built from the post-shift `stem_len_abs`, giving every
ellipse – from both epochs – a directly comparable elevation in one shared
frame. That single column is what Stage 5's vertical-profile logic reads.

`las_2` is the fixed reference cloud: it is never shifted, so its absolute
fields pass through untouched and only gain the `slice_z_global` column.

# Stage 0 - Setup


## 0.1 Load libraries
```{r}
#| label: setup
#| code-summary: "Load libraries and set plot theme"

library(lidR)       # LAS I/O and point cloud processing
library(ggplot2)    # Plotting
library(dplyr)      # Data wrangling
library(plotly)     # Interactive 3D figures
library(tidyr)      # pivot_longer for the before/after centre plot
library(base64enc)  # Embed rendered plots into HTML galleries
library(htmltools)  # Assemble the HTML gallery output

# ── Driver-supplied parameter overrides ──────────────────────────────────────
# Every params-* chunk in this file ends with .par_apply("<its label>"). Run the
# file by hand and that call does nothing: the values the chunk just assigned
# stand. Run it under the batch driver, which puts a .batch_params list in the
# environment before the first chunk, and any entry keyed to that chunk label is
# assigned here - AFTER the defaults, so the driver's value is the one that
# survives into the functional chunk that follows.
#
# Why this exists. Parameters used to sit in the same chunk as the code that
# consumed them, so the driver's overrides - which land after a chunk finishes -
# arrived too late to matter, and every pair silently used the .qmd's hardcoded
# value. Splitting each chunk in two fixes the ordering; this helper is what the
# driver reaches through.
#
# The defaults stay in this file rather than moving into the driver. They are
# the single definition, the driver names only what a batch changes, and there
# is no second copy of ~670 parameters to drift out of step with this one.
#
# envir = parent.frame() so a call at chunk top level lands in the environment
# the chunk is being evaluated in - globalenv by hand, the pair environment
# under the driver - without needing .stage_env(), which is defined further down.
.par_apply <- function(label, envir = parent.frame()) {
  bp <- mget(".batch_params", envir = envir, ifnotfound = list(NULL),
             inherits = TRUE)[[1]]
  if (is.null(bp) || is.null(bp[[label]])) return(invisible(character(0)))
  vals <- bp[[label]]
  if (!length(vals)) return(invisible(character(0)))
  if (is.null(names(vals)) || any(!nzchar(names(vals))))
    stop(sprintf("[.par_apply] .batch_params[['%s']] must be a fully NAMED list.", label))
  # A name the chunk did not define is almost always a typo or a rename in the
  # .qmd that the driver has not caught up with. Silence there would mean the
  # batch runs on the default while looking like it was overridden, so it stops.
  unknown <- names(vals)[!vapply(names(vals), exists, logical(1),
                                 envir = envir, inherits = FALSE)]
  if (length(unknown))
    stop(sprintf(paste0("[.par_apply] %s: %d override(s) name nothing this chunk assigns:\n  %s\n",
                        "  Fix the name in the driver's .build_params(), or remove it."),
                 label, length(unknown), paste(unknown, collapse = "\n  ")))
  for (nm in names(vals)) assign(nm, vals[[nm]], envir = envir)
  cat(sprintf("  [params] %s <- driver (%s)\n", label, paste(names(vals), collapse = ", ")))
  invisible(names(vals))
}
```

## 0.2 Global Helpers
```{r}
#| label: params-helpers
#| code-summary: "Parameters - Colours, ggplot theme, stage save/load utilities, and the on-demand point-cloud loader"

# ============================================================
# Colours
# ============================================================
COL_1     <- "#2196F3"
COL_2     <- "#FF5722"
COL_MATCH <- "#D9F9A5"

# ============================================================
# Progress bars
#
# Thin wrappers around cli, using EXACTLY the idiom Stage 3's axis-RANSAC fit
# has always used, because that one works when these documents are run on their
# own. Their only job is to save repeating the same six lines at every long
# loop.
#
# THE ONE THING THAT MUST NOT BE CHANGED HERE is `.envir = parent.frame()`.
# cli ties a progress bar's lifetime to a frame and tears the bar down when
# that frame exits. .pb_new() is a function, so a bar created with cli's
# default `.envir` would be created and immediately destroyed the moment
# .pb_new returned - the loop then ticks a bar that no longer exists and
# nothing is ever drawn. parent.frame() hands the bar to the CALLER's frame,
# which is the loop, and is the difference between this working and silently
# doing nothing. (An earlier revision passed globalenv() here, which is why the
# bars added through this helper never appeared while the hand-written ones in
# 3.3, 3.6 and 3.9 did.)
#
# Under the batch driver these will NOT be visible: cli draws on stderr, and
# the driver holds a message sink, which cannot be split. That is expected and
# is not worked around here - the driver prints its own progress instead.
# ============================================================
# Progress bars
# ============================================================
# One set of helpers, two renderers, chosen at the moment a bar is created.
#
#   cli    a redrawing bar, when there is a terminal able to show one
#   lines  one ordinary printed line per update, when there is not
#
# The choice is not cosmetic. cli draws on STDERR, and the batch driver holds a
# sink(type = "message"), which cannot split - so under the driver a cli bar
# goes to the log file and nowhere else, arriving as carriage returns and
# escape codes wrapped around each other. That is what the last run's log shows
# for every fitting step except the library's own: minutes of work with nothing
# on screen, and an unreadable smear in the log afterwards.
#
# The printed-line renderer is what the RANSAC library's cat handler already
# does, and it is the one visible line in that log. Same format here, so a fit
# reports identically whether the progress comes from the library or from a
# loop in this file.
show_progress_bars <- TRUE

# Print at most one line per 5% of the total or per 15 seconds, whichever comes
# first, plus always the first and last. A 20-stem fit gives ~20 lines; a
# 5000-slice loop does not give 5000.
progress_step_frac    <- 0.05
progress_min_seconds  <- 15

.par_apply("params-helpers")
```

```{r}
#| label: helpers
#| code-summary: "Colours, ggplot theme, stage save/load utilities, and the on-demand point-cloud loader"

# ============================================================
# ggplot theme
# ============================================================
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      legend.position  = "bottom"
    )
)
.write_las_snapshot <- function(las_template, xyz_mat, output_path, stage_label) {
  las_snap        <- las_template
  las_snap@data$X <- xyz_mat[, 1]
  las_snap@data$Y <- xyz_mat[, 2]
  las_snap@data$Z <- xyz_mat[, 3]

  # Make output_path unique if the file already exists (appends _001, _002, ...)
  if (file.exists(output_path)) {
    base_noext <- tools::file_path_sans_ext(output_path)
    ext        <- tools::file_ext(output_path)
    counter    <- 1L
    repeat {
      candidate <- sprintf("%s_%03d.%s", base_noext, counter, ext)
      if (!file.exists(candidate)) { output_path <- candidate; break }
      counter <- counter + 1L
    }
    cat(sprintf("Output file already exists - using unique name: %s\n", output_path))
  }

  writeLAS(las_snap, output_path)
  cat(sprintf("[%s] Saved %s points -> %s\n",
              stage_label, format(nrow(xyz_mat), big.mark = ","), output_path))
  invisible(output_path)
}
angle_from_x_ccw <- function(dx, dy) atan2(dy, dx) * 180 / pi


# ============================================================
# Stage save/load utilities – shared by every stage file.
#
# Contract (identical in all six stage files):
# SAVE → hard ERROR if any *required* object is missing from the
# environment, so a checkpoint can never be written half-empty.
# A *warning* is raised for any missing *optional* object
# (diagnostic products that may legitimately not have been run);
# those are simply omitted from the file.
# LOAD → never errors on content. A *warning* is raised for each
# requested object that is absent from the RDS. Objects present in
# the RDS but not requested are COUNTED in the summary line and
# never listed. Everything found in the file is loaded either way,
# so nothing is silently discarded.
# PATHS → a save never overwrites: if the target path exists, a numeric
# suffix (_001, _002, ...) is appended until the name is free.
# ============================================================

#' Return `path` if free, otherwise path_001 / path_002 / ... (first free one).
.unique_path <- function(path) {
  if (!file.exists(path)) return(path)
  base_noext <- tools::file_path_sans_ext(path)
  ext        <- tools::file_ext(path)
  counter    <- 1L
  repeat {
    candidate <- sprintf("%s_%03d.%s", base_noext, counter, ext)
    if (!file.exists(candidate)) {
      cat(sprintf("Target already exists - using unique name: %s\n", candidate))
      return(candidate)
    }
    counter <- counter + 1L
  }
}

#' Which environment do the stage helpers read and write?
#'
#' A .qmd rendered or run by hand evaluates its chunks in the global
#' environment. The batch driver does not: it gives every LAS pair a fresh
#' environment of its own, so that a leftover object from the previous pair
#' cannot be picked up by one of the exists() guards this pipeline is full of.
#' Under the driver, everything the stages assign therefore lives THERE, and
#' not in globalenv.
#'
#' parent.env(environment()) is the environment this chunk was evaluated in:
#' globalenv() by hand, the pair environment under the driver. It is resolved
#' from where .stage_env was DEFINED rather than from where it is called, so it
#' returns the same environment inside a function as it does at chunk top
#' level, and nothing has to be told which mode it is in.
#'
#' This replaced a hardcoded .stage_env() throughout. Under the driver that
#' pointed at an environment the pipeline had never written to, so every object
#' in a save manifest read as missing and the checkpoint refused to write - the
#' Stage 1 failure that reported 30 missing required objects after a 64-minute
#' fit. (`run_counts` was the one it did find, and only because .tally()'s `<<-`
#' walked past the pair environment into globalenv. That is fixed below too.)
.stage_env <- function() parent.env(environment())

#' Is a batch driver running this file?
#'
#' The driver assigns .batch_mode <- TRUE into the pair environment before the
#' first chunk runs. It changes one thing, and only for the checkpoints: in a
#' batch, every stage runs in ONE session, so Stage N+1 already holds in memory
#' everything Stage N computed and never has to read it back off disk. The
#' checkpoints stop being load-bearing and become a record of the run.
#'
#' So under a batch: nothing in a manifest is required (a missing object is
#' reported and left out, and the pair carries on), a failed write is reported
#' rather than thrown, and .stage_load is a no-op. Run the same .qmd by hand and
#' every one of those is an error again, which is what it should be when the
#' checkpoint is the only thing carrying the stage forward.
.in_batch <- function() {
  isTRUE(mget(".batch_mode", envir = .stage_env(),
              ifnotfound = list(FALSE), inherits = FALSE)[[1]])
}

#' Read one of the driver's batch settings out of the pair environment.
#' Absent - i.e. this file is being run by hand - gives `default`.
.batch_opt <- function(name, default = NULL) {
  v <- mget(name, envir = .stage_env(), ifnotfound = list(NULL), inherits = FALSE)[[1]]
  if (is.null(v)) default else v
}

#' Where a checkpoint goes when a batch is running.
#'
#' Every save chunk in this pipeline assigns its .rds path and then USES it a
#' few lines later, in the same chunk. The driver applies its overrides AFTER a
#' chunk has run, so an override of stage1_rds_path arrives after the file has
#' already been written to the path hardcoded here - which means, without this,
#' every sample in a batch writes over the same hardcoded name.
#'
#' So in batch mode the destination is derived instead, from two values the
#' driver puts in the environment before anything runs: .batch_rds_dir and
#' .batch_sample. The label the save chunk already passes supplies the rest, so
#' "Stage 3.2" becomes <dir>/<sample>_Stage3_2.rds with nothing to keep in sync.
#'
#' NULL means: not in a batch, or the driver did not set the directory - use the
#' path the chunk asked for.
.batch_rds_path <- function(label) {
  dir <- .batch_opt(".batch_rds_dir")
  smp <- .batch_opt(".batch_sample")
  if (is.null(dir) || is.null(smp)) return(NULL)
  file.path(dir, sprintf("%s_%s.rds", smp,
                         gsub("\\.", "_", gsub("\\s+", "", label))))
}

#' Save a stage checkpoint. Errors on missing required objects - unless the
#' batch driver is running this file, in which case nothing is required, the
#' destination is the per-sample one derived above, and a failure to write is
#' reported rather than thrown. See .in_batch() above.
`%||%` <- function(a, b) if (is.null(a)) b else a

.stage_save <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  required <- unique(required)
  optional <- setdiff(unique(optional), required)

  # A batch may switch an individual checkpoint off by label - the driver's
  # keep_stage3_fit_checkpoint does exactly this for "Stage 3.1", whose chunk
  # must still RUN (the final Stage 3 manifest inherits from it) but need not
  # write a second large file per sample.
  if (!strict && label %in% .batch_opt(".batch_skip_saves", character(0))) {
    cat(sprintf("[%s save] Skipped - switched off for this batch.\n", label))
    return(invisible(NA_character_))
  }

  # inherits = FALSE, matching the mget() below, which has always been strict.
  # With inheritance on, an object could be FOUND in a parent environment and
  # then fail to be FETCHED, and under the driver the parent is globalenv -
  # i.e. the previous pair's leftovers.
  .have <- function(x)
    if (!length(x)) logical(0)
    else vapply(x, exists, logical(1), envir = envir, inherits = FALSE)

  missing_req <- required[!.have(required)]
  if (length(missing_req) > 0L) {
    if (strict)
      stop(sprintf(
        "[%s save] Cannot save - %d required object(s) missing from the environment:\n  %s",
        label, length(missing_req), paste(missing_req, collapse = "\n  ")))
    cat(sprintf(
      "[%s save] %d required object(s) missing - treated as optional and omitted (batch mode):\n  %s\n",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))
  }

  missing_opt <- optional[!.have(optional)]
  if (length(missing_opt) > 0L) {
    msg <- sprintf(
      "[%s save] %d optional object(s) missing - omitted from the checkpoint:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  "))
    if (strict) warning(msg) else cat(msg, "\n", sep = "")
  }

  present  <- c(setdiff(required, missing_req), setdiff(optional, missing_opt))
  n_opt_in <- length(setdiff(optional, missing_opt))
  dat      <- mget(present, envir = envir, inherits = FALSE)

  # ── The checkpoint describes itself ────────────────────────────────────────
  # The manifest travels INSIDE the file. Without it, loading a checkpoint means
  # first reconstructing the exact `required`/`optional` vectors it was written
  # with - and those are built by chunks, so a later checkpoint could only be
  # loaded after running the chunks that define an earlier one's manifest. That
  # made resuming from an arbitrary point in a fresh session impossible: the
  # Stage 3.2 chunk, for instance, stops outright unless .stage3_1_required is
  # already in scope.
  #
  # Stored under a name no manifest can collide with, and stripped again on
  # load, so it never reaches the session as an object.
  dat[[".__manifest__"]] <- list(
    required = setdiff(required, missing_req),
    optional = setdiff(optional, missing_opt),
    label    = label,
    saved_utc = format(Sys.time(), tz = "UTC", usetz = TRUE))

  .write <- function() {
    p <- path
    if (!strict) {
      bp <- .batch_rds_path(label)
      if (!is.null(bp)) p <- bp
    }
    dir.create(dirname(p), recursive = TRUE, showWarnings = FALSE)
    # .unique_path only interactively. A batch path already carries the sample
    # name, so it cannot collide with another sample, and a re-run of the same
    # sample should REPLACE its checkpoint rather than leave _001, _002 behind
    # for the next reader to guess between.
    if (strict) p <- .unique_path(p)
    saveRDS(dat, p)
    cat(sprintf("[%s save] Saved %d object(s) (%d required, %d optional) -> %s\n",
                label, length(dat), length(present) - n_opt_in, n_opt_in, p))
    p
  }

  if (strict) return(invisible(.write()))

  out <- tryCatch(.write(), error = function(e) {
    cat(sprintf("[%s save] Checkpoint NOT written: %s\n", label, conditionMessage(e)))
    cat(sprintf("[%s save] Batch mode - the objects are still in this session, so the run continues.\n",
                label))
    NA_character_
  })
  invisible(out)
}

#' Load a stage checkpoint. Warns (never errors) on content mismatches, and
#' under the batch driver does not read at all - see .in_batch() above.
.stage_load <- function(path, required, optional = character(0),
                        label = "stage", envir = .stage_env(),
                        strict = !.in_batch()) {

  # In a batch the objects this would read are already live in `envir`, put
  # there by the stage that just ran. Reading them back would at best be a
  # no-op. At worst it would be wrong: checkpoint paths are per-sample and
  # .unique_path() means a re-run writes _001 rather than replacing the
  # original, so the path named here can easily be an EARLIER run's file. And
  # if the write failed or was partial, there would be nothing to read and the
  # pair would die at the start of the next stage having done all the work.
  if (!strict) {
    cat(sprintf("[%s load] Skipped - batch mode: this stage's objects are already in the session.\n",
                label))
    return(invisible(character(0)))
  }

  if (!file.exists(path))
    stop(sprintf("[%s load] File not found:\n  %s", label, path))

  dat <- readRDS(path)

  # ── Prefer the manifest the file carries ───────────────────────────────────
  # A checkpoint written by .stage_save records what it holds, so a load needs
  # nothing from the session - no earlier chunk, no earlier checkpoint, no
  # reconstructed name vectors. `required` and `optional` become a CHECK against
  # what the caller expected rather than the definition of what to read, which
  # is what lets any stage be resumed from any checkpoint in a fresh session.
  #
  # Files written before manifests existed simply have none; those fall back to
  # the caller's vectors, exactly as before.
  mf <- dat[[".__manifest__"]]
  dat[[".__manifest__"]] <- NULL
  caller_req <- unique(required)
  caller_opt <- setdiff(unique(optional), caller_req)

  if (!is.null(mf)) {
    required <- unique(mf$required)
    optional <- setdiff(unique(mf$optional), required)
    src <- sprintf("its own manifest (written %s)", mf$saved_utc %||% "?")
  } else {
    required <- caller_req; optional <- caller_opt
    src <- "the caller's manifest - this file predates self-describing checkpoints"
  }
  expected <- c(required, optional)

  cat(sprintf("[%s load] Reading %d object(s) from: %s\n",
              label, length(dat), path))
  cat(sprintf("[%s load] Manifest: %s\n", label, src))

  # Named by the caller but absent from the file: the real incompatibility, and
  # the only one worth a warning when the file describes itself.
  if (!is.null(mf)) {
    unmet <- setdiff(caller_req, names(dat))
    if (length(unmet))
      warning(sprintf(
        "[%s load] %d object(s) this stage expects are not in the checkpoint:\n  %s",
        label, length(unmet), paste(unmet, collapse = "\n  ")))
  }

  missing_req <- required[!required %in% names(dat)]
  if (length(missing_req) > 0L)
    warning(sprintf(
      "[%s load] %d REQUIRED object(s) absent from the RDS - downstream chunks that use them will fail:\n  %s",
      label, length(missing_req), paste(missing_req, collapse = "\n  ")))

  missing_opt <- optional[!optional %in% names(dat)]
  if (length(missing_opt) > 0L)
    warning(sprintf(
      "[%s load] %d optional object(s) absent from the RDS:\n  %s",
      label, length(missing_opt), paste(missing_opt, collapse = "\n  ")))

  # Unrequested objects are COUNTED, never listed. Every checkpoint from Stage 2
  # onward is cumulative, so this set is routinely hundreds of names long and
  # printing it buried the two messages that actually matter - an absent
  # required object, and an absent optional one - under a wall of text. The
  # count still appears in the summary line below, which is the only part of it
  # anyone reads.
  extra <- setdiff(names(dat), expected)

  # ── Publish what was loaded, for the next stage's save manifest ────────────
  # Every stage's SAVE manifest is built as "the previous stage's manifest, plus
  # my own objects", and the previous manifest is a vector defined by that
  # stage's LOAD chunk. Load a LATER checkpoint directly and that chunk never
  # runs, so the save dies on a missing .stageN_required - which is exactly the
  # resume-from-anywhere case these checkpoints now exist to support.
  #
  # Recording the manifest here breaks the dependency: .manifest_from() below
  # falls back to whatever was actually loaded, whichever chunk did it.
  assign(".stage_last_manifest",
         list(required = required, optional = optional,
              names = names(dat), label = label, path = path),
         envir = envir)

  list2env(dat, envir = envir)

  cat(sprintf("[%s load] Loaded %d object(s) | required present: %d/%d | optional present: %d/%d | unrequested: %d\n",
              label, length(dat),
              length(required) - length(missing_req), length(required),
              length(optional) - length(missing_opt), length(optional),
              length(extra)))
  invisible(names(dat))
}

#' The manifest a save chunk inherits from the stage before it.
#'
#' Looks for the named vector first - that is the normal path, when this
#' session's own load chunk built it. Falls back to the manifest carried by
#' whichever checkpoint was last loaded, which is what makes it possible to load
#' any checkpoint and still save the next one.
#'
#' `which` is "required" or "optional". Stops with something actionable rather
#' than returning character(0): a save that silently inherits nothing writes a
#' checkpoint missing every object from every earlier stage, and nothing would
#' notice until a later stage failed to find them.
.manifest_from <- function(name, which = c("required", "optional"),
                           envir = .stage_env()) {
  which <- match.arg(which)
  if (exists(name, envir = envir, inherits = FALSE))
    return(get(name, envir = envir, inherits = FALSE))

  lm <- if (exists(".stage_last_manifest", envir = envir, inherits = FALSE))
          get(".stage_last_manifest", envir = envir, inherits = FALSE) else NULL

  if (!is.null(lm)) {
    cat(sprintf("[manifest] %s not in the session - inheriting from the %s checkpoint that was loaded.\n",
                name, lm$label))
    # For "required", everything the loaded file actually held is the honest
    # inheritance: it is what this session has, regardless of how the file that
    # produced it split required from optional.
    return(if (identical(which, "required")) lm$names else character(0))
  }

  # ── Last resort: the session itself ────────────────────────────────────────
  # No named vector and no checkpoint loaded. That is the ordinary case for a
  # session built by RUNNING chunks rather than by loading - Stage 1 through
  # Stage 3 straight down the document, or an interactive session where the
  # objects were produced rather than read. There is nothing wrong with it, and
  # a save must not refuse.
  #
  # So fall back to what Stages 4 and 5 have always done: take the manifest from
  # the environment. Every non-function, non-dotted object in scope is what this
  # stage inherited, by definition - it is the same rule, just applied later.
  #
  # Only "required" gets this. Optional means "save it if it happens to exist",
  # which is meaningless once the list IS everything that exists.
  if (identical(which, "required")) {
    nm <- setdiff(ls(envir = envir), lsf.str(envir = envir))
    nm <- nm[!startsWith(nm, ".")]
    if (length(nm)) {
      cat(sprintf(paste0("[manifest] %s not found and no checkpoint was loaded - taking the manifest\n",
                         "           from the session instead (%d object(s)). This is the normal path\n",
                         "           when the stage was RUN rather than loaded.\n"), name, length(nm)))
      return(nm)
    }
  } else {
    return(character(0))
  }

  stop(sprintf(paste0("[manifest] %s not found, no checkpoint has been loaded, and the session is\n",
                      "  empty - there is nothing to save. Run this stage's chunks, or load a\n",
                      "  checkpoint, before the save chunk."), name))
}

#' Print the Stage 1 pivot point.
.report_pivot1 <- function() {
  if (exists("pivot1")) {
    cat(sprintf("pivot1 (Stage 1 footprint centre): (%.4f, %.4f, %.4f)\n",
                pivot1[1], pivot1[2], pivot1[3]))
  } else {
    cat("pivot1 NOT available.\n")
  }
}

# ============================================================
# Stages 2, 4 and 5 deliberately never load point clouds - they work purely on
# fitted ellipse geometry, which is what keeps them cheap to re-run. A handful
# of 3-D figures in those stages genuinely need raw points, so each of those
# chunks carries its own loader at the top, controlled by one toggle.
#
# Stage 3 carries it too, for a different reason: it DOES load the clouds at the
# top, but releases them again part-way through, and resuming from the Stage 3
# post-fit checkpoint skips the load entirely.
#
# Stages 1 and 6 do NOT have this block at all - they load the clouds up front
# and keep them for the whole stage, so nothing there would ever call it.
#
#   vis_load_clouds <- TRUE    read las_1 / las_2 from disk
#   vis_load_clouds <- FALSE   reuse whatever is already resident in memory
#
# Set it FALSE when the clouds are already loaded - you just ran Stage 3 in the
# same session, or an earlier figure in this stage already loaded them. Reading
# a multi-gigabyte LAS once per figure is slow and pointless. Leave it TRUE when
# running the stage standalone from its checkpoint, which is the normal case.
#
# ── FRAMES: why las_1 needs a rotation the moment it comes off disk ────────────
# las_1 sits on disk in the ORIGINAL scan frame. Stage 1A's rough vertical
# alignment (R_align_1 about align_pivot_1) was applied to las_1 IN MEMORY in
# Stage 1 and never written back - there is no writeLAS anywhere in Stage 1. So
# every measurement made downstream of Stage 1A - ellipse centres, per-stem axis
# vectors, pivot1, and the Stage 2 rotation solve R_xyz - live in the ALIGNED
# frame, not the disk frame. A figure that reads las_1 off disk and overlays any
# of that geometry is mixing two frames unless it first reapplies R_align_1.
#
# This loader therefore reconstructs the pipeline frame for las_1 as a stack of
# rigid steps, matching the Stage 2 composition (R_align_xyz_1) exactly:
#
#   x_aligned = R_align_1 (x_raw     - align_pivot_1) + align_pivot_1   (apply_align, Stage 1A)
#   x_final   = R_xyz     (x_aligned - pivot1)        + pivot1          (apply_rot,   Stage 2)
#   x_final  += t_trans_z                                              (apply_zshift, Stage 4)
#   x_final   = R_z-only  about pivot2 in XY                            (apply_zrot,   Stage 5)
#
# apply_align is the default-ON baseline: everything above stacks on the
# aligned frame, so leaving it on is what makes las_1 line up with the ellipses
# and vectors. Turn it OFF only to inspect the raw disk cloud deliberately.
# las_2 is the REFERENCE cloud - Stage 1 never rotates it and its ellipses are in
# the disk frame - so it is never transformed here.
#
# Transforms are applied to the RETURNED point frame, never baked back into the
# LAS object, so repeated calls (or a load = FALSE reuse) can never compound.
# ============================================================

#' Make point clouds available to one visualisation chunk.
#'
#' @param load         TRUE to read from disk; FALSE to reuse las_1 / las_2
#'                     already in the global environment.
#' @param las_1_path,las_2_path
#'                     Optional explicit source paths. When NULL (default) the
#'                     paths are taken from the global environment, so a caller
#'                     can either pass them or rely on the checkpoint. Only used
#'                     when load = TRUE.
#' @param apply_align  reapply the Stage 1A rough vertical alignment to las_1
#'                     (R_align_1 about align_pivot_1). Default TRUE: this is the
#'                     baseline frame every Stage 1/2 measurement lives in, so
#'                     without it the returned las_1 points do NOT match the
#'                     ellipses or axis vectors. Set FALSE only to see the raw
#'                     on-disk cloud. No-op (identity) when rough alignment was
#'                     not run; skipped with a note if the checkpoint lacks the
#'                     objects.
#' @param apply_rot    rebuild the Stage 2 XYZ rotation R_xyz (about pivot1),
#'                     stacked on the aligned frame.
#' @param apply_zshift rebuild the Stage 4 Z-translation.
#' @param apply_zrot   rebuild the Stage 5 Z-only rotation (about pivot2 in XY).
#' @param need_las2    FALSE to skip the reference cloud entirely.
#'
#' @return list(las_1, las_2, pts_1, pts_2, xyz_1, state). pts_* are plain data
#'   frames of X/Y/Z/TreeID with TreeID 0 and NA dropped (figures show stems).
#'   xyz_1 / pts_1 are in the transformed frame; the LAS objects stay on-disk.
.vis_point_clouds <- function(load,
                              las_1_path   = NULL,
                              las_2_path   = NULL,
                              apply_align  = TRUE,
                              apply_rot    = FALSE,
                              apply_zshift = FALSE,
                              apply_zrot   = FALSE,
                              need_las2    = TRUE,
                              quiet        = FALSE,
                              envir        = .stage_env()) {

  .say <- function(...) if (!quiet) cat(...)
  .g   <- function(nm) get(nm, envir = envir)
  .has <- function(nm) exists(nm, envir = envir)

  # Resolve source paths: explicit argument wins, else the global/checkpoint.
  .p1 <- if (!is.null(las_1_path)) las_1_path else
         if (.has("las_1_path")) .g("las_1_path") else NULL
  .p2 <- if (!is.null(las_2_path)) las_2_path else
         if (.has("las_2_path")) .g("las_2_path") else NULL

  if (isTRUE(load)) {

    if (is.null(.p1))
      stop("[vis loader] las_1_path not supplied and not in the environment - ",
           "pass it or load this stage's checkpoint first.")

    .say(sprintf("[vis loader] Reading las_1 from disk: %s\n", basename(.p1)))
    l1 <- lidR::readLAS(.p1)
    if (lidR::is.empty(l1)) stop("[vis loader] las_1 read as empty: ", .p1)
    l1 <- lidR::add_lasattribute(l1, x = seq_len(nrow(l1@data)),
                                 name = "point_id", desc = "Unique ID for las_1 points")
    l2 <- NULL
    if (need_las2) {
      if (is.null(.p2))
        stop("[vis loader] las_2_path not supplied and not in the environment.")
      .say(sprintf("[vis loader] Reading las_2 from disk: %s\n", basename(.p2)))
      l2 <- lidR::readLAS(.p2)
      if (lidR::is.empty(l2)) stop("[vis loader] las_2 read as empty: ", .p2)
      l2 <- lidR::add_lasattribute(l2, x = seq_len(nrow(l2@data)),
                                   name = "point_id", desc = "Unique ID for las_2 points")
    }

    assign("las_1", l1, envir = envir)
    if (!is.null(l2)) assign("las_2", l2, envir = envir)

  } else {

    if (!.has("las_1"))
      stop("[vis loader] vis_load_clouds = FALSE but las_1 is not in memory. Set it TRUE.")
    if (need_las2 && !.has("las_2"))
      stop("[vis loader] vis_load_clouds = FALSE but las_2 is not in memory. Set it TRUE.")

    l1 <- .g("las_1")
    l2 <- if (need_las2) .g("las_2") else NULL
    .say("[vis loader] Reusing the point clouds already in memory (no disk read).\n")
  }

  # ── Rebuild the transform state (las_1 only; las_2 is never transformed) ──────
  # The stack runs in pipeline order on the RETURNED frame: Stage 1A align first,
  # then the Stage 2 rotation, then the Stage 4/5 shifts. Each step is a rigid
  # transform about the pivot the corresponding stage used, matching the Stage 2
  # composition. IDENTICAL in Stages 2, 3, 4 and 5 - if it is edited, edit all four.
  xyz   <- as.matrix(l1@data[, c("X", "Y", "Z")])
  state <- character(0)

  # ── Stage 1A: rough vertical alignment (raw disk -> aligned frame) ────────────
  if (isTRUE(apply_align)) {
    if (!.has("R_align_1") || !.has("align_pivot_1")) {
      .say("[vis loader] R_align_1 / align_pivot_1 not in the checkpoint - ",
           "las_1 left in the raw disk frame (no Stage 1A alignment).\n")
    } else {
      Ra <- .g("R_align_1"); pa <- as.numeric(.g("align_pivot_1"))
      xyz <- sweep(t(Ra %*% t(sweep(xyz, 2, pa))), 2, -pa)
      # Angle of the rigid rotation, for the log only: acos((tr(R) - 1) / 2).
      .ang_a <- acos(max(-1, min(1, (sum(diag(Ra)) - 1) / 2))) * 180 / pi
      state  <- c(state, sprintf("stage-1A align %.4f deg about align_pivot_1", .ang_a))
    }
  }

  # ── Stage 2: XYZ rotation R_xyz about pivot1 (aligned -> final) ───────────────
  if (isTRUE(apply_rot)) {
    if (!.has("R_xyz") || !.has("pivot1"))
      stop("[vis loader] apply_rot = TRUE but R_xyz / pivot1 not found - run the ",
           "Stage 2 rotation solve first, or reload a checkpoint that carries them.")
    R  <- .g("R_xyz"); pv <- as.numeric(.g("pivot1"))
    xyz <- sweep(t(R %*% t(sweep(xyz, 2, pv))), 2, -pv)
    .ang_x <- if (.has("angle_xyz")) .g("angle_xyz") else
              acos(max(-1, min(1, (sum(diag(R)) - 1) / 2))) * 180 / pi
    state <- c(state, sprintf("xyz-rot %.4f deg about pivot1", .ang_x))
  }

  if (isTRUE(apply_zshift)) {
    tz <- if (.has("t_trans_z_applied")) .g("t_trans_z_applied") else .g("t_trans_z")
    xyz[, 3] <- xyz[, 3] + tz
    state <- c(state, sprintf("z-shift %+.4f m", tz))
  }

  if (isTRUE(apply_zrot)) {
    # Same sign flip the Stage 5 / Stage 6 builders use: z_rotation_deg is held
    # in the standard CCW-positive convention, the 2x2 fill below evaluates to
    # the matrix for -a, so the angle is negated exactly once here.
    a  <- -.g("z_rotation_deg") * pi / 180
    Rz <- matrix(c(cos(a), -sin(a), sin(a), cos(a)), 2, 2)
    pv <- if (.has("pivot2")) .g("pivot2") else .g("pivot1")
    xc <- xyz[, 1] - pv[1]; yc <- xyz[, 2] - pv[2]
    xyz[, 1] <- Rz[1, 1] * xc + Rz[1, 2] * yc + pv[1]
    xyz[, 2] <- Rz[2, 1] * xc + Rz[2, 2] * yc + pv[2]
    state <- c(state, sprintf("z-rot %+.4f deg", .g("z_rotation_deg")))
  }

  pts_1 <- data.frame(X = xyz[, 1], Y = xyz[, 2], Z = xyz[, 3], TreeID = l1@data$TreeID)
  pts_1 <- pts_1[!is.na(pts_1$TreeID) & pts_1$TreeID != 0L, ]

  pts_2 <- NULL
  if (!is.null(l2)) {
    pts_2 <- data.frame(X = l2@data$X, Y = l2@data$Y, Z = l2@data$Z, TreeID = l2@data$TreeID)
    pts_2 <- pts_2[!is.na(pts_2$TreeID) & pts_2$TreeID != 0L, ]
  }

  .say(sprintf("[vis loader] las_1 state: %s\n",
               if (length(state)) paste(state, collapse = " + ") else "as loaded (no transform)"))
  .say(sprintf("[vis loader] stem points available - las_1: %s | las_2: %s\n",
               format(nrow(pts_1), big.mark = ","),
               if (is.null(pts_2)) "not loaded" else format(nrow(pts_2), big.mark = ",")))

  invisible(list(las_1 = l1, las_2 = l2, pts_1 = pts_1, pts_2 = pts_2,
                 xyz_1 = xyz, state = state))
}

# ============================================================
# Run tally - one cumulative record of how many things survived each filter
#
# Every stage prints its own counts, but those printouts vanish with the render
# and cannot be compared across runs. .tally() writes the same numbers into a
# single named list, `run_counts`, that travels in the checkpoints and is read
# out whole by the Stage 6 error report. A run then answers "how many pairs
# reached the Z-translation" without anyone re-reading five HTML files.
#
# Cumulative and idempotent: each call updates only the keys it names, so
# re-running one chunk refreshes that chunk's numbers and leaves the rest alone.
# It never errors - a tally is bookkeeping, and a missing count must not be able
# to stop a stage from saving.
# ============================================================
.tally <- function(...) {
  vals <- list(...)
  e    <- .stage_env()
  base <- if (exists("run_counts", envir = e, inherits = FALSE))
            get("run_counts", envir = e) else list()
  assign("run_counts", utils::modifyList(base, vals), envir = e)
  invisible(get("run_counts", envir = e))
}


#' TRUE when a cli bar can actually be seen.
#'
#' All four conditions matter: cli absent, output being knitted, a message sink
#' in the way (the batch driver), or a terminal that cannot redraw - any one of
#' them and the bar is invisible, mangled, or both.
.pb_cli_usable <- function() {
  isNamespaceLoaded("cli") || requireNamespace("cli", quietly = TRUE)
  requireNamespace("cli", quietly = TRUE) &&
    !isTRUE(getOption("knitr.in.progress")) &&
    sink.number(type = "message") == 0L &&
    isTRUE(try(cli::is_dynamic_tty(), silent = TRUE))
}

.pb_fmt_dur <- function(s) {
  if (!is.finite(s)) return("--")
  if (s < 90) sprintf("%.0fs", s) else if (s < 5400) sprintf("%.1fm", s / 60)
  else sprintf("%.1fh", s / 3600)
}

#' Start a bar. Returns an opaque handle for .pb_tick()/.pb_done(), or NULL.
#'
#' .envir = parent.frame() is load-bearing for the cli branch: cli ties a bar's
#' lifetime to a frame, so a bar created with the default .envir would be torn
#' down the instant this function returned and every later tick would silently
#' do nothing. That was a real bug here once.
.pb_new <- function(total, label) {
  total <- as.integer(total)
  if (!isTRUE(show_progress_bars) || is.na(total) || total < 1L) return(NULL)

  if (.pb_cli_usable()) {
    id <- cli::cli_progress_bar(
      name   = label,
      total  = total,
      format = paste0(label,
                      " {cli::pb_bar} {cli::pb_current}/{cli::pb_total}",
                      " | {cli::pb_elapsed} | {cli::pb_eta} left"),
      .envir = parent.frame())
    return(structure(list(kind = "cli", id = id), class = ".pb"))
  }

  # An ENVIRONMENT, not a list: a tick inside lapply() has to update the
  # caller's counter, and a list passed by value would not.
  st <- new.env(parent = emptyenv())
  st$i <- 0L; st$total <- total; st$label <- label
  st$t0 <- proc.time()[["elapsed"]]; st$last_t <- -Inf; st$last_i <- -1L
  structure(list(kind = "lines", st = st), class = ".pb")
}

.pb_tick <- function(pb, by = 1L) {
  if (is.null(pb)) return(invisible(NULL))
  if (identical(pb$kind, "cli")) {
    cli::cli_progress_update(id = pb$id, inc = by)
    return(invisible(NULL))
  }
  st <- pb$st
  st$i <- st$i + by
  now <- proc.time()[["elapsed"]]
  step <- max(1L, as.integer(ceiling(st$total * progress_step_frac)))
  due  <- st$i >= st$total || st$i == 1L ||
          (st$i - st$last_i) >= step ||
          (now - st$last_t) >= progress_min_seconds
  if (!due) return(invisible(NULL))
  st$last_i <- st$i; st$last_t <- now
  el   <- now - st$t0
  frac <- st$i / st$total
  eta  <- if (st$i > 0L) el / st$i * (st$total - st$i) else NA_real_
  nfill <- as.integer(round(20 * frac))
  cat(sprintf("  %s [%s%s] %d/%d %3.0f%% | %s elapsed | ~%s left\n",
              st$label, strrep("#", nfill), strrep("-", 20L - nfill),
              st$i, st$total, 100 * frac, .pb_fmt_dur(el), .pb_fmt_dur(eta)))
  flush.console()
  invisible(NULL)
}

.pb_done <- function(pb) {
  if (is.null(pb)) return(invisible(NULL))
  if (identical(pb$kind, "cli")) {
    cli::cli_progress_done(id = pb$id)
  } else if (pb$st$i < pb$st$total) {
    # The loop ended early (an epoch with nothing to do, say). Say so rather
    # than leaving a bar stopped at 12/25 as though something had hung.
    cat(sprintf("  %s: finished after %d of %d | %s\n", pb$st$label, pb$st$i,
                pb$st$total, .pb_fmt_dur(proc.time()[["elapsed"]] - pb$st$t0)))
  }
  invisible(NULL)
}
```

## 0.3 Load the Stage 3 Checkpoint

```{r}
#| label: params-load_Stage4
#| code-summary: "Parameters - Load the Stage 3 checkpoint (no point clouds are read in this stage)"


# Only the last Stage 3 export is read here. The Stage 4 OUTPUT path
# (stage4_rds_path) is defined in the Stage 4 save chunk, where it is used.
stage3_2_rds_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/RunR/exports/SaveLoad/I_SNBG_F1_ai_2020_2024_TLS_TLS_Stage3_2.rds"

.par_apply("params-load_Stage4")
```

```{r}
#| label: load_Stage4
#| code-summary: "Load the Stage 3 checkpoint (no point clouds are read in this stage)"
#| code-fold: false

# Stage 4 uses only a handful of Stage 3 objects. .stage_load pulls EVERYTHING in
# the RDS via list2env regardless of this list (the manifest only drives the
# present/absent warnings), so rather than mirror Stage 3.2's ~300-name manifest
# here - which would drift the moment Stage 3 changes - Stage 4 requests just what
# it consumes and lets the loader bring the rest in silently. The unrequested
# (loaded anyway)" note from .stage_load is expected and harmless: those objects
# are still in the environment for the Stage 4 save chunk to re-checkpoint.
.stage4_needs <- c(
  # geometry the shift operates on
  "pivot1",
  "ellipses_ax_1_valid", "ellipses_ax_2_valid",
  # pair universe kept after the Stage 2 XYZ rotation (the set the Z-rot solve
  # runs on) and the narrower axis-fit set (used for the z-translation pair list)
  "pairs_stage2",
  "pairs_ax_fitted",
  # axes for the tilt diagnostic
  "vecs_axis_1_rot", "vecs_xy_2",
  # the recorded (unapplied) Z-translation from Steps 3.17/3.18
  "t_trans_z"
)
.stage4_needs_opt <- c(
  # present in this Stage 3 build, useful for context, absent on some runs
  "s5_signal_registry",
  "s58_pair_tbl", "s58_pair_shift",
  "s5_pm_true_shift",
  "apply_mode_3_18", "custom_z_shift_m"
)

.stage_load(stage3_2_rds_path, .stage4_needs, .stage4_needs_opt, label = "Stage 3.2")

cat("\n\u2500\u2500 Carried in from Stages 1-3 \u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\u2500\n")
if (exists(".report_pivot1")) .report_pivot1()

cat(sprintf("Pairs kept after XYZ rotation (pairs_ax_fitted): %d\n", nrow(pairs_ax_fitted)))
if (exists("s58_pair_tbl") && !is.null(s58_pair_tbl))
  cat(sprintf("  of which %d contributed to the Z-translation estimate (Step 3.16)\n",
              nrow(s58_pair_tbl)))
cat(sprintf("Z-translation recorded by Stage 3: %+.4f m (not yet applied)\n", t_trans_z))
cat("No point-cloud data is loaded in this stage.\n")
```


# Stage 4 - Z- Rotation

## 4.1 Confirm or Override the Z-Shift

```{r}
#| label: params-zshift-toggle
#| code-summary: "Parameters - Toggle: use the Stage 3 Z-shift, or override it with a custom value"


# apply_mode_zshift = "auto" → apply t_trans_z exactly as selected in Stage 3
# Step 3.15 (default).
# apply_mode_zshift = "custom" → apply a hand-entered value instead. Useful for
# testing a value read out of CloudCompare, or
# for reproducing a manual alignment.
#
# Either way the value actually applied is stored as t_trans_z_applied, so the
# checkpoint always records what was really used rather than what was proposed.

apply_mode_zshift <- "auto"      # "auto" or "custom"
custom_z_shift_stage4 <- -0.7294 # metres; only used when apply_mode_zshift == "custom"

.par_apply("params-zshift-toggle")
```

```{r}
#| label: zshift-toggle
#| code-summary: "Toggle: use the Stage 3 Z-shift, or override it with a custom value"

if (identical(apply_mode_zshift, "custom")) {
  t_trans_z_applied <- custom_z_shift_stage4
  cat(sprintf(
    "Z-translation OVERRIDDEN with a custom value: %+.4f m\n  (Stage 3 selected %+.4f m)\n",
    t_trans_z_applied, t_trans_z
  ))
} else {
  t_trans_z_applied <- t_trans_z
  cat(sprintf("Applying the Stage 3 Z-translation: %+.4f m (%+.2f cm)\n",
              t_trans_z_applied, t_trans_z_applied * 100))
}
```

## 4.2 Shift `pivot1` to `pivot2`

```{r}
#| label: params-pivot2
#| code-summary: "Parameters - Translate pivot1 vertically by the applied Z-shift -> pivot2"


# pivot1 lives in the post-rotation frame (Stage 3 rotated the cloud about it).
# Once the whole cloud moves vertically by t_trans_z_applied, the pivot moves
# with it – otherwise Stage 5 would rotate the shifted ellipse centres about a
# point that no longer sits at the cloud's footprint centre.
#
# Only Z changes: a vertical translation cannot move a horizontal centre. That
# is why pivot2[1:2] == pivot1[1:2] exactly, and it is the reason Stage 6 can
# legitimately compose both rotations about pivot1 alone (see the Stage 6
# equivalence note).

pivot2 <- pivot1 + c(0, 0, t_trans_z_applied)

.par_apply("params-pivot2")
```

```{r}
#| label: pivot2
#| code-summary: "Translate pivot1 vertically by the applied Z-shift -> pivot2"

names(pivot2) <- c("X", "Y", "Z")

cat("── pivot1 -> pivot2 ───────────────────────────────────────────────────────\n")
cat(sprintf("  pivot1: (%.4f, %.4f, %.4f)\n", pivot1[1], pivot1[2], pivot1[3]))
cat(sprintf("  shift : (%+.4f, %+.4f, %+.4f)\n", 0, 0, t_trans_z_applied))
cat(sprintf("  pivot2: (%.4f, %.4f, %.4f)\n", pivot2[1], pivot2[2], pivot2[3]))
```

## 4.3 Shift the Ellipse Centres

```{r}
#| label: params-apply-zshift-ellipses
#| code-summary: "Parameters - Apply the Z-shift to the las_1 ellipse centres -> ellipses_zshift_1 / _2"


# ── Absolute vs. stem-relative height fields ─────────────────────────────────
# Only genuine elevations move. See the stage overview for the full reasoning;
# in short, stem_len / slice_z are measured from each stem's OWN origin, and
# that origin translates by the same amount as the stem, so they are invariant
# under a rigid vertical shift. Shifting them too would apply the offset twice.
.abs_z_fields <- c("center_z", "global_stem_origin_z",
                   "stem_len_origin", "stem_len_abs")

.par_apply("params-apply-zshift-ellipses")
```

```{r}
#| label: apply-zshift-ellipses
#| code-summary: "Apply the Z-shift to the las_1 ellipse centres -> ellipses_zshift_1 / _2"

#' Shift every absolute-height field of one tree's ellipse table by dz, then
#' (re)build slice_z_global from the post-shift stem_len_abs.
.shift_tree_ellipses <- function(df, tid, dz) {
  if (is.null(df) || nrow(df) == 0L) return(NULL)
  df$TreeID <- as.integer(tid)
  for (f in .abs_z_fields) {
    if (f %in% names(df)) df[[f]] <- df[[f]] + dz
  }
  # Shared-frame elevation of each ellipse. Built from stem_len_abs so that the
  # two epochs are directly comparable: las_1 has been shifted by dz, las_2 has
  # dz = 0, and both end up on the same absolute scale.
  if ("stem_len_abs" %in% names(df)) {
    df$slice_z_global <- df$stem_len_abs
  } else {
    df$slice_z_global <- NA_real_
  }
  df
}

# ── las_1: shifted by t_trans_z_applied ──────────────────────────────────────
ellipses_zshift_1 <- setNames(
  lapply(names(ellipses_ax_1_valid), function(tid)
    .shift_tree_ellipses(ellipses_ax_1_valid[[tid]], tid, t_trans_z_applied)),
  names(ellipses_ax_1_valid)
)
ellipses_zshift_1 <- Filter(Negate(is.null), ellipses_zshift_1)

# ── las_2: the fixed reference cloud – dz = 0, fields pass through unchanged ──
ellipses_zshift_2 <- setNames(
  lapply(names(ellipses_ax_2_valid), function(tid)
    .shift_tree_ellipses(ellipses_ax_2_valid[[tid]], tid, 0)),
  names(ellipses_ax_2_valid)
)
ellipses_zshift_2 <- Filter(Negate(is.null), ellipses_zshift_2)

# ── Flattened las_1 table, kept for inspection and downstream convenience ────
ellipses_zt_1_shifted <- dplyr::bind_rows(ellipses_zshift_1)

cat(sprintf(
  "Z-shift applied: %+.4f m (%+.2f cm)\n",
  t_trans_z_applied, t_trans_z_applied * 100
))
cat(sprintf(
  "  ellipses_zshift_1: %d tree(s), %d ellipse row(s) shifted\n",
  length(ellipses_zshift_1), nrow(ellipses_zt_1_shifted)
))
cat(sprintf(
  "  ellipses_zshift_2: %d tree(s), %d ellipse row(s) (reference epoch, unshifted)\n",
  length(ellipses_zshift_2),
  sum(vapply(ellipses_zshift_2, nrow, integer(1)))
))

# ── Verification: the two epochs should now overlap in absolute height ───────
.rng <- function(lst) {
  v <- unlist(lapply(lst, function(d) d$slice_z_global), use.names = FALSE)
  v <- v[is.finite(v)]
  if (length(v) == 0L) return(c(NA_real_, NA_real_))
  range(v)
}
.r1 <- .rng(ellipses_zshift_1)
.r2 <- .rng(ellipses_zshift_2)
cat("\n── slice_z_global ranges after the shift ─────────────────────────────────\n")
cat(sprintf("  las_1 (shifted):   [%.3f, %.3f] m\n", .r1[1], .r1[2]))
cat(sprintf("  las_2 (reference): [%.3f, %.3f] m\n", .r2[1], .r2[2]))
cat(sprintf("  Overlap:           [%.3f, %.3f] m (%.3f m deep)\n",
            max(.r1[1], .r2[1]), min(.r1[2], .r2[2]),
            max(0, min(.r1[2], .r2[2]) - max(.r1[1], .r2[1]))))
```

## 4.4 Stem Tilt Annotation (diagnostic)

```{r}
#| label: stage4-tilt-annotation
#| code-summary: "Report per-epoch stem tilt from the axis vectors (diagnostic only)"

# Deviation from vertical, in degrees, for each stem axis. Purely descriptive:
# nothing downstream filters on it. las_1 uses the ROTATED axes produced in
# Stage 3.2, so this reports tilt as it stands after the XYZ rotation.
tilt_fn <- function(vx, vy, vz) {
  len <- sqrt(vx^2 + vy^2 + vz^2)
  if (is.na(len) || len < 1e-8 || is.na(vz)) return(NA_real_)
  acos(pmax(-1, pmin(1, abs(vz) / len))) * 180 / pi
}

stem_tilt_1 <- data.frame(
  TreeID   = vecs_axis_1_rot$TreeID,
  tilt_deg = mapply(tilt_fn, vecs_axis_1_rot$vx_rot,
                    vecs_axis_1_rot$vy_rot, vecs_axis_1_rot$vz_rot)
)
stem_tilt_2 <- data.frame(
  TreeID   = vecs_xy_2$TreeID,
  tilt_deg = mapply(tilt_fn, vecs_xy_2$vx, vecs_xy_2$vy, vecs_xy_2$vz)
)

tilt_report <- function(tilt_vec, label) {
  t_clean <- tilt_vec[is.finite(tilt_vec)]
  if (length(t_clean) == 0) {
    cat(sprintf("%s tilt report: no valid tilt values\n", label)); return(invisible(NULL))
  }
  cat(sprintf(
    "%s tilt report (deg): min=%.2f | median=%.2f | mean=%.2f | max=%.2f | n=%d\n",
    label, min(t_clean), median(t_clean), mean(t_clean), max(t_clean),
    length(t_clean)
  ))
}
tilt_report(stem_tilt_1$tilt_deg, "las_1 (post-rotation axes)")
tilt_report(stem_tilt_2$tilt_deg, "las_2 (reference axes)")
```

## 4.5 Z-overlap filter

```{r}
#| label: params-pairs-stage4
#| code-summary: "Parameters - Per-pair z-rotation window from ellipse-CENTRE overlap: zero-point = higher of the two lowest centres, then keep centres in [zero+zrot_bot, zero+zrot_top]"


# ── What this does ────────────────────────────────────────────────────────────
# Builds the pair list AND the per-pair ellipse subset the Z-rotation solve runs
# on, working purely on ellipse-CENTRE heights - a kept ellipse was fit in the
# stem-perpendicular frame, so its centre height IS its location; there is no
# slice extent to compute.
#
# Pair universe: every pair kept after the XYZ rotation (pairs_stage2), narrowed
# to those that actually have Stage-3 ellipses on both sides. Using pairs_stage2
# (not the z-translation subset) keeps the solve on as many XYZrot survivors as
# have geometry.
#
# Per pair, on the shifted centre heights (slice_z_global):
# 1. OVERLAP ZERO-POINT. z0 = max(min(z1), min(z2)) - the higher of the two
# stems' LOWEST centres. Whichever stem has the higher lowest ellipse sets
# the floor and sits exactly on it; the other stem's nearest ellipse above
# the floor is, by construction, within one slice step (staggering). The
# overlap top is z_top = min(max(z1), max(z2)).
# 2. WINDOW. Take the subsection [z0 + zrot_bot, z0 + zrot_top] measured from
# the zero-point. With zrot_bot = 0 the zero-point ellipse itself is
# included. Keep, per stem, the centres whose slice_z_global falls in that
# band (with a half-slice-step tolerance so a centre sitting essentially on
# an edge is not lost to rounding).
# A pair survives when BOTH stems keep >= 1 ellipse in the window and the overlap
# is real (z_top >= z0). The kept ellipse subsets are what the 4.3 centre-average
# consumes.

# ── Window parameters ─────────────────────────────────────────────────────────
zrot_bot   <- 0.00   # window bottom, metres above the overlap zero-point
zrot_top   <- 1.00   # window top,    metres above the overlap zero-point

.par_apply("params-pairs-stage4")
```

```{r}
#| label: pairs-stage4
#| code-summary: "Per-pair z-rotation window from ellipse-CENTRE overlap: zero-point = higher of the two lowest centres, then keep centres in [zero+zrot_bot, zero+zrot_top]"

s5_slice_step_m <- if (exists("s5_slice_step")) s5_slice_step else 0.10  # fit slice spacing
.half_step <- s5_slice_step_m / 2                                        # edge tolerance

# ── Pair universe: XYZrot survivors with Stage-3 ellipses on both sides ───────
pairs_zrot_ok <- if (exists("pairs_stage2")) pairs_stage2 else pairs_ax_fitted
.base_pairs <- pairs_zrot_ok |>
  dplyr::filter(as.character(TreeID_1) %in% names(ellipses_zshift_1),
                as.character(TreeID_2) %in% names(ellipses_zshift_2))
n_base <- nrow(.base_pairs)
.tally(n_pairs_stage4_in = n_base)

# centre heights (shifted) for one stem, sorted ascending; NA-safe
.centre_z <- function(df) {
  if (is.null(df) || nrow(df) == 0L) return(numeric(0))
  v <- df$slice_z_global
  sort(v[is.finite(v)])
}

# ── Per-pair overlap window + kept-ellipse subset ─────────────────────────────
kept_1 <- list(); kept_2 <- list(); rows <- list()

for (i in seq_len(n_base)) {
  t1 <- as.character(.base_pairs$TreeID_1[i])
  t2 <- as.character(.base_pairs$TreeID_2[i])
  d1 <- ellipses_zshift_1[[t1]]
  d2 <- ellipses_zshift_2[[t2]]
  z1 <- .centre_z(d1); z2 <- .centre_z(d2)

  if (!length(z1) || !length(z2)) {
    rows[[i]] <- data.frame(TreeID_1 = .base_pairs$TreeID_1[i],
                            TreeID_2 = .base_pairs$TreeID_2[i],
                            z0 = NA_real_, z_top = NA_real_,
                            win_lo = NA_real_, win_hi = NA_real_,
                            n1 = 0L, n2 = 0L, kept = FALSE,
                            reason = "no centres on one side")
    next
  }

  # overlap on centre heights: zero-point = higher of the two lowest centres
  z0    <- max(min(z1), min(z2))
  z_top <- min(max(z1), max(z2))

  if (!(z_top + .half_step >= z0)) {           # no real overlap
    rows[[i]] <- data.frame(TreeID_1 = .base_pairs$TreeID_1[i],
                            TreeID_2 = .base_pairs$TreeID_2[i],
                            z0 = z0, z_top = z_top,
                            win_lo = NA_real_, win_hi = NA_real_,
                            n1 = 0L, n2 = 0L, kept = FALSE,
                            reason = "no centre overlap")
    next
  }

  # window measured from the zero-point; clamp the top to the actual overlap top
  win_lo <- z0 + zrot_bot
  win_hi <- min(z0 + zrot_top, z_top)

  # keep centres in [win_lo, win_hi], with a half-step edge tolerance so the
  # zero-point ellipse (and any centre sitting on an edge) is retained
  in_win <- function(df, z) {
    if (is.null(df) || nrow(df) == 0L) return(df[0, , drop = FALSE])
    v <- df$slice_z_global
    df[is.finite(v) & v >= (win_lo - .half_step) & v <= (win_hi + .half_step), ,
       drop = FALSE]
  }
  k1 <- in_win(d1); k2 <- in_win(d2)

  keep <- nrow(k1) > 0L && nrow(k2) > 0L
  if (keep) { kept_1[[t1]] <- k1; kept_2[[t2]] <- k2 }

  rows[[i]] <- data.frame(TreeID_1 = .base_pairs$TreeID_1[i],
                          TreeID_2 = .base_pairs$TreeID_2[i],
                          z0 = z0, z_top = z_top,
                          win_lo = win_lo, win_hi = win_hi,
                          n1 = nrow(k1), n2 = nrow(k2), kept = keep,
                          reason = if (keep) NA_character_ else "empty window on one side")
}

stage4_overlap_tbl <- dplyr::bind_rows(rows)
pairs_stage4 <- .base_pairs[stage4_overlap_tbl$kept %in% TRUE, , drop = FALSE]
.tally(n_pairs_zoverlap_kept = nrow(pairs_stage4))

# the window-restricted ellipse subsets the Z-rot solve (4.3) consumes
ellipses_zrot_1 <- kept_1
ellipses_zrot_2 <- kept_2

# ── Report ────────────────────────────────────────────────────────────────────
n_keep <- nrow(pairs_stage4)
cat("Stage 4 z-rotation pair window:\n")
cat(sprintf("  XYZrot pairs with Stage-3 ellipses (pairs_zrot_ok): %d\n", n_base))
cat(sprintf("  window: [z0 + %.2f, z0 + %.2f] m above each pair's overlap zero-point\n",
            zrot_bot, zrot_top))
cat(sprintf("  slice step %.3f m (edge tolerance +/- %.3f m)\n",
            s5_slice_step_m, .half_step))
cat(sprintf("  pairs kept (both sides non-empty in window): %d\n", n_keep))

n_drop <- n_base - n_keep
if (n_drop > 0L) {
  dd <- stage4_overlap_tbl[!(stage4_overlap_tbl$kept %in% TRUE),
                           c("TreeID_1", "TreeID_2", "z0", "z_top", "n1", "n2", "reason"),
                           drop = FALSE]
  dd$z0 <- round(dd$z0, 3); dd$z_top <- round(dd$z_top, 3)
  cat(sprintf("  dropped %d pair(s):\n", n_drop))
  print(dd, row.names = FALSE)
}
```

### Vis - ellipse centre elevations before and after the shift

```{r}
#| label: fig-zshift-elevations
#| fig-cap: "Absolute ellipse-centre elevation per PAIR, las_1 and las_2 side by side, before and after the Stage 4 Z-shift. Golden band = that pair's z-rotation window."
#| fig-height: 7

if (params$make_plots) {
  # map each stem to the pair it belongs to, so the two epochs share one x slot.
  # las_1 is keyed by TreeID_1, las_2 by TreeID_2; both resolve to the same pair.
  .pair_key   <- paste0(pairs_stage4$TreeID_1, "\u2192", pairs_stage4$TreeID_2)
  .k1 <- setNames(.pair_key, as.character(pairs_stage4$TreeID_1))  # TreeID_1 -> pair
  .k2 <- setNames(.pair_key, as.character(pairs_stage4$TreeID_2))  # TreeID_2 -> pair

  .gather <- function(lst, cloud_lab, dz_back, keymap) {
    dplyr::bind_rows(lapply(names(lst), function(tid) {
      d <- lst[[tid]]
      if (is.null(d) || nrow(d) == 0L) return(NULL)
      pk <- keymap[[tid]]
      if (is.null(pk)) return(NULL)          # stem not in the Stage 4 pair list
      tibble::tibble(
        pair   = pk,
        TreeID = as.integer(tid),
        cloud  = cloud_lab,
        after  = d$slice_z_global,
        before = d$slice_z_global - dz_back
      )
    }))
  }

  z_before_after <- dplyr::bind_rows(
    .gather(ellipses_zshift_1, "las_1", t_trans_z_applied, .k1),
    .gather(ellipses_zshift_2, "las_2", 0,                 .k2)
  ) |>
    tidyr::pivot_longer(c(before, after), names_to = "state", values_to = "z") |>
    dplyr::mutate(state = factor(state, levels = c("before", "after")))

  # order pairs on the x-axis by their post-shift median elevation so the panel
  # reads left-to-right; keeps the two epochs of a pair locked to one slot.
  .ord <- z_before_after |>
    dplyr::filter(state == "after") |>
    dplyr::group_by(pair) |>
    dplyr::summarise(mz = median(z, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(mz)
  z_before_after$pair <- factor(z_before_after$pair, levels = .ord$pair)

  # ── Per-pair z-rotation window band (golden) ────────────────────────────────
  # The window is per-pair (each pair has its own overlap zero-point z0), so the
  # band sits at a DIFFERENT elevation in every x-slot and is drawn only in the
  # "after" facet, where the window is defined. win_lo / win_hi come straight
  # from the pair-list chunk's stage4_overlap_tbl.
  .band <- stage4_overlap_tbl |>
    dplyr::filter(.data$kept %in% TRUE,
                  is.finite(.data$win_lo), is.finite(.data$win_hi)) |>
    dplyr::mutate(pair = paste0(TreeID_1, "\u2192", TreeID_2)) |>
    dplyr::filter(pair %in% levels(z_before_after$pair)) |>
    dplyr::mutate(
      pair  = factor(pair, levels = levels(z_before_after$pair)),
      xi    = as.integer(pair),          # discrete x position of this pair's slot
      state = factor("after", levels = c("before", "after"))
    )

  gg <- ggplot(z_before_after,
               aes(x = pair, y = z, colour = cloud,
                   group = interaction(pair, cloud)))

  # golden band behind the points, one rect per pair spanning its own window
  if (nrow(.band) > 0L) {
    gg <- gg +
      geom_rect(data = .band, inherit.aes = FALSE,
                aes(xmin = xi - 0.45, xmax = xi + 0.45,
                    ymin = win_lo,   ymax = win_hi),
                fill = "#d4a017", alpha = 0.25)
  }

  gg +
    # dodge the two epochs within each pair's slot so they sit adjacent, not stacked
    geom_point(size = 0.7, alpha = 0.5,
               position = position_dodge(width = 0.6)) +
    facet_wrap(~ state, ncol = 1, scales = "fixed") +
    scale_colour_manual(values = c("las_1" = COL_1, "las_2" = COL_2), name = NULL) +
    labs(
      title    = sprintf("Ellipse-centre elevations by pair, Z-shift = %+.4f m", t_trans_z_applied),
      subtitle = sprintf("Each x slot is one pair; las_1/las_2 dodged. Golden band = z-rot window [z0+%.2f, z0+%.2f] m (after only).",
                         zrot_bot, zrot_top),
      x = "pair (TreeID_1 \u2192 TreeID_2, ordered by post-shift elevation)",
      y = "slice_z_global (m)"
    ) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6))
}
```


### Vis - point clouds after Z-translation

Stage 4 works on ellipse geometry alone, so this figure loads and transforms
its own copy of the clouds. The loader rebuilds the Stage 2 rotation and the
Z-shift applied above, so what is drawn is the cloud as it stands at the end of
this stage.

### Vis - point clouds after Z-translation
```{r}
#| label: params-fig-zt-3D
#| code-summary: "Parameters - fig-zt-3D"


# ── Point clouds for this figure ──────────────────────────────────────────────
# This stage does not load point clouds at the top, so this figure loads its own.
# vis_load_clouds <- TRUE read las_1 / las_2 from disk and apply the
# Stage 1A alignment + the Stage 2 rotation reapplied
# vis_load_clouds <- FALSE reuse the clouds already in memory (no disk read)
# The loader also rebuilds the transform state this figure expects, so the
# points shown match the pipeline stage rather than the file on disk.
vis_load_clouds <- TRUE

.par_apply("params-fig-zt-3D")
```

```{r}
#| label: fig-zt-3D
#| fig-cap: "Paired tree point clouds after full XYZ rotation + Z-translation (0-5 m stem)."

.vpc <- .vis_point_clouds(
  load = vis_load_clouds,
  apply_rot     = TRUE,
  apply_zshift  = TRUE
)

las1_pts_zt <- .vpc$pts_1   # las_1: rotation + Z-shift rebuilt
las2_pts_zt <- .vpc$pts_2   # las_2: reference epoch, untransformed


if (params$make_plots) {

  max_pts_zt   <- 1000L   # <- custom point count per stem, adjust as needed
  zt_slice_top <- 5       # metres above each stem's local Z minimum

  stem_pts_zt <- bind_rows(lapply(seq_len(nrow(pairs_stage4)), function(i) {
    tid1 <- pairs_stage4$TreeID_1[i]
    tid2 <- pairs_stage4$TreeID_2[i]

    pts1_all <- las1_pts_zt[las1_pts_zt$TreeID == tid1, ]
    if (nrow(pts1_all) == 0) return(NULL)
    z0_1 <- min(pts1_all$Z)
    pts1 <- pts1_all[pts1_all$Z - z0_1 <= zt_slice_top, ]
    if (nrow(pts1) > max_pts_zt) pts1 <- pts1[sample.int(nrow(pts1), max_pts_zt), ]
    pts1 <- pts1 |> mutate(cloud = "las_1 (xyz-rot + z-shift)", TreeID_pair = tid1)

    pts2_all <- las2_pts_zt[las2_pts_zt$TreeID == tid2, ]
    if (nrow(pts2_all) == 0) return(NULL)
    z0_2 <- min(pts2_all$Z)
    pts2 <- pts2_all[pts2_all$Z - z0_2 <= zt_slice_top, ]
    if (nrow(pts2) > max_pts_zt) pts2 <- pts2[sample.int(nrow(pts2), max_pts_zt), ]
    pts2 <- pts2 |> mutate(cloud = "las_2", TreeID_pair = tid1)

    bind_rows(pts1, pts2)
  }))

  global_cx_zt <- mean(stem_pts_zt$X[stem_pts_zt$cloud == "las_2"], na.rm = TRUE)
  global_cy_zt <- mean(stem_pts_zt$Y[stem_pts_zt$cloud == "las_2"], na.rm = TRUE)

  stem_pts_zt <- stem_pts_zt |>
    mutate(X_local = X - global_cx_zt, Y_local = Y - global_cy_zt)

  plot_ly(
    data = stem_pts_zt, x = ~X_local, y = ~Y_local, z = ~Z,
    color = ~cloud,
    colors = c("las_1 (xyz-rot + z-shift)" = COL_1, "las_2" = COL_2),
    type = "scatter3d", mode = "markers",
    marker = list(size = 2, opacity = 0.6),
    text = ~paste0("Cloud: ", cloud, "<br>Pair: ", TreeID_pair, "<br>Z: ", round(Z, 3)),
    hoverinfo = "text"
  ) |>
    layout(
      title = sprintf(
        "Full XYZ rotation + Z-translation – %d pairs (0-%d m, %d pts/stem)",
        nrow(pairs_stage4), zt_slice_top, max_pts_zt
      ),
      scene = list(
        xaxis = list(title = "X local (m)"),
        yaxis = list(title = "Y local (m)"),
        zaxis = list(title = "Z (m)"),
        aspectmode = "data",
        camera = list(eye = list(x = 1.5, y = 1.5, z = 0.8))
      ),
      legend = list(title = list(text = "Cloud"))
    )
}
```



## 4.6 Average Ellipse Centres per Tree

```{r}
#| label: params-avg-ellipse-centres-zrot
#| code-summary: "Parameters - Collapse each tree's WINDOWED ellipse subset to one representative XY centre per epoch"


# ─────────────────────────────────────────────────────────────────────────────
# INPUT: pairs_zrot_ok (= pairs_stage2), ellipses_zrot_1, ellipses_zrot_2
# OUTPUT: avg_centres_1, avg_centres_2, pairs_zrot_valid
# ─────────────────────────────────────────────────────────────────────────────
# The yaw solve works on ONE point per stem per epoch: the mean of that stem's
# ellipse centres WITHIN the z-rotation window [z0+zrot_bot, z0+zrot_top]. Those
# windowed subsets are ellipses_zrot_1 / ellipses_zrot_2, built per pair by the
# 4.2 Z-overlap chunk - NOT the full-stem ellipses_zshift_*. Averaging only the
# windowed centres is the whole point of the bot/top range: the rotation is
# estimated from the band where the two epochs actually overlap.
#
# The PAIR UNIVERSE is still the full XYZrot-survivor set (pairs_stage2). Because
# ellipses_zrot_* is keyed only for pairs that survived the window with a
# non-empty subset on both sides, any pairs_stage2 pair outside that set has no
# entry here and is dropped by compute_avg_centre as "not found" - so the
# effective set is exactly "XYZrot pairs that also have a non-empty windowed
# ellipse subset on both sides".
pairs_zrot_ok <- pairs_stage2

.par_apply("params-avg-ellipse-centres-zrot")
```

```{r}
#| label: avg-ellipse-centres-zrot
#| code-summary: "Collapse each tree's WINDOWED ellipse subset to one representative XY centre per epoch"

.zrot_pairs <- pairs_zrot_ok |>

  dplyr::mutate(TreeID_1 = as.integer(TreeID_1),
                TreeID_2 = as.integer(TreeID_2))

# per-tree mean centre (and mean shape), ID-safe against the named ellipse list
compute_avg_centre <- function(ellipse_list, tree_ids) {
  results <- lapply(as.integer(tree_ids), function(tid_int) {
    tid_chr <- as.character(tid_int)
    df      <- ellipse_list[[tid_chr]]
    if (is.null(df))   return(list(row = NULL, invalid_reason = "not found in windowed ellipse list"))
    if (nrow(df) == 0) return(list(row = NULL, invalid_reason = "windowed subset is empty"))
    df <- df[!is.na(df$center_x) & !is.na(df$center_y), ]
    if (nrow(df) == 0) return(list(row = NULL, invalid_reason = "all windowed centres are NA"))
    a_mean <- mean(df$semi_a, na.rm = TRUE); b_mean <- mean(df$semi_b, na.rm = TRUE)
    list(
      row = data.frame(
        TreeID    = tid_int,
        cx        = mean(df$center_x, na.rm = TRUE),
        cy        = mean(df$center_y, na.rm = TRUE),
        a_mean    = a_mean, b_mean = b_mean,
        ang_mean  = mean(df$angle_deg, na.rm = TRUE),
        area_mean = if ("ell_area" %in% names(df)) mean(df$ell_area, na.rm = TRUE) else NA_real_,
        r_mean    = sqrt(a_mean * b_mean),
        n_win     = nrow(df)            # how many windowed ellipses backed this centre
      ),
      invalid_reason = NA_character_)
  })
  valid_rows   <- dplyr::bind_rows(lapply(results, `[[`, "row"))
  invalid_flag <- vapply(results, function(x) !is.na(x$invalid_reason), logical(1))
  invalid_df   <- data.frame(
    TreeID = as.integer(tree_ids)[invalid_flag],
    reason = vapply(results[invalid_flag], `[[`, character(1), "invalid_reason"))
  list(centres = valid_rows, invalid = invalid_df)
}

res_1 <- compute_avg_centre(ellipses_zrot_1, .zrot_pairs$TreeID_1)
res_2 <- compute_avg_centre(ellipses_zrot_2, .zrot_pairs$TreeID_2)
avg_centres_1 <- res_1$centres
avg_centres_2 <- res_2$centres

cat(sprintf("Average centres computed from the z-rot window [z0+%.2f, z0+%.2f] m - las_1: %d tree(s) | las_2: %d tree(s)\n",
            zrot_bot, zrot_top, nrow(avg_centres_1), nrow(avg_centres_2)))

if (nrow(res_1$invalid) > 0) {
  cat(sprintf("\nlas_1 trees with no usable windowed centre (%d):\n", nrow(res_1$invalid)))
  print(res_1$invalid, row.names = FALSE)
}
if (nrow(res_2$invalid) > 0) {
  cat(sprintf("\nlas_2 trees with no usable windowed centre (%d):\n", nrow(res_2$invalid)))
  print(res_2$invalid, row.names = FALSE)
}

# keep only pairs with a valid windowed centre on BOTH sides
pairs_zrot_valid <- .zrot_pairs |>
  dplyr::filter(TreeID_1 %in% avg_centres_1$TreeID,
                TreeID_2 %in% avg_centres_2$TreeID)

n_drop <- nrow(.zrot_pairs) - nrow(pairs_zrot_valid)
if (n_drop > 0) cat(sprintf("\n%d pair(s) dropped for a missing/empty windowed centre.\n", n_drop))
cat(sprintf("Pairs with valid windowed avg centres in both epochs: %d / %d\n",
            nrow(pairs_zrot_valid), nrow(.zrot_pairs)))
```

## 4.7 Create Line Network

```{r}
#| label: build-line-network-zrot
#| code-summary: "Build the inter-tree line network on windowed averaged centres and measure per-edge angular difference between epochs"

# Every unordered pair of kept stems is an EDGE. For each edge, take the las_1
# direction (from the two las_1 windowed-average centres) and the las_2 direction
# (from the corresponding las_2 centres) and measure the angle between them.
# Under a pure yaw, every edge rotates by the same angle, so ang_diff is a
# per-edge estimate of the rotation; averaging the clean ones (4.5-4.7) solves it.
centers_zr_1 <- avg_centres_1 |> dplyr::filter(TreeID %in% pairs_zrot_valid$TreeID_1)
centers_zr_2 <- avg_centres_2 |> dplyr::filter(TreeID %in% pairs_zrot_valid$TreeID_2)
pair_map_zr  <- pairs_zrot_valid |> dplyr::select(TreeID_1, TreeID_2)

tree_ids_zr <- centers_zr_1$TreeID
edge_idx_zr <- which(upper.tri(matrix(NA, length(tree_ids_zr), length(tree_ids_zr))),
                     arr.ind = TRUE)

edges_zr1 <- data.frame(
  TreeID_1_a = tree_ids_zr[edge_idx_zr[, 1]],
  TreeID_1_b = tree_ids_zr[edge_idx_zr[, 2]]
) |>
  dplyr::left_join(centers_zr_1 |> dplyr::rename(cx1_a = cx, cy1_a = cy),
                   by = c("TreeID_1_a" = "TreeID")) |>
  dplyr::left_join(centers_zr_1 |> dplyr::rename(cx1_b = cx, cy1_b = cy),
                   by = c("TreeID_1_b" = "TreeID"))

edges_zr2 <- edges_zr1 |>
  dplyr::left_join(pair_map_zr |> dplyr::rename(TreeID_1_a = TreeID_1, TreeID_2_a = TreeID_2),
                   by = "TreeID_1_a") |>
  dplyr::left_join(pair_map_zr |> dplyr::rename(TreeID_1_b = TreeID_1, TreeID_2_b = TreeID_2),
                   by = "TreeID_1_b") |>
  dplyr::left_join(centers_zr_2 |> dplyr::rename(cx2_a = cx, cy2_a = cy),
                   by = c("TreeID_2_a" = "TreeID")) |>
  dplyr::left_join(centers_zr_2 |> dplyr::rename(cx2_b = cx, cy2_b = cy),
                   by = c("TreeID_2_b" = "TreeID"))

line_df_zr <- edges_zr2 |>
  dplyr::filter(!is.na(cx1_a), !is.na(cx1_b), !is.na(cx2_a), !is.na(cx2_b)) |>
  dplyr::mutate(
    dx1 = cx1_b - cx1_a, dy1 = cy1_b - cy1_a, angle_1 = angle_from_x_ccw(dx1, dy1),
    dx2 = cx2_b - cx2_a, dy2 = cy2_b - cy2_a, angle_2 = angle_from_x_ccw(dx2, dy2),
    ang_diff = angle_2 - angle_1,
    mx = (cx1_a + cx1_b) / 2, my = (cy1_a + cy1_b) / 2
  )

# fold angular differences into (-180, 180] so a wrap at +/-180 does not distort
# the median/mean of the solve
line_df_zr$ang_diff <- ((line_df_zr$ang_diff + 180) %% 360) - 180

cat(sprintf("Line network: %d edge(s) across %d stem(s)\n",
            nrow(line_df_zr), length(tree_ids_zr)))
cat(sprintf("Edge-wise angular difference (deg): median %.3f | mean %.3f | sd %.3f\n",
            median(line_df_zr$ang_diff), mean(line_df_zr$ang_diff), sd(line_df_zr$ang_diff)))
```

### Vis - ellipse-centre line angles

```{r}
#| label: fig-ellipse-lines-angles
#| fig-cap: "las_1 centre network (best 6 stems by kept ellipse count); each edge coloured by angular difference to the corresponding las_2 edge."
#| fig-height: 7
#| fig-width: 7

if (params$make_plots && nrow(line_df_zr) > 0) {
  # rank stems by how many WINDOWED ellipses backed their centre (reliability
  # proxy), reading the windowed subsets - not the full stem
  ell_count_zr <- vapply(as.character(pairs_zrot_valid$TreeID_1), function(tid_chr) {
    df <- ellipses_zrot_1[[tid_chr]]; if (is.null(df)) 0L else nrow(df)
  }, integer(1))
  names(ell_count_zr) <- as.character(pairs_zrot_valid$TreeID_1)
  top6_ids <- names(sort(ell_count_zr, decreasing = TRUE))[seq_len(min(6L, length(ell_count_zr)))]

  line_df_zr_top6 <- line_df_zr |>
    dplyr::filter(as.character(TreeID_1_a) %in% top6_ids &
                  as.character(TreeID_1_b) %in% top6_ids)
  ellipses_top6_pts <- dplyr::bind_rows(lapply(top6_ids, function(tid_chr) {
    df <- ellipses_zrot_1[[tid_chr]]
    if (is.null(df)) return(NULL)
    d <- df[!is.na(df$center_x), c("center_x", "center_y")]
    dplyr::rename(d, X = center_x, Y = center_y)
  }))
  centers_zr_1_top6 <- centers_zr_1 |> dplyr::filter(as.character(TreeID) %in% top6_ids)

  print(
    ggplot() +
      geom_point(data = ellipses_top6_pts, aes(X, Y),
                 colour = COL_1, size = 0.4, alpha = 0.5) +
      geom_segment(data = line_df_zr_top6,
                   aes(x = cx1_a, y = cy1_a, xend = cx1_b, yend = cy1_b, colour = ang_diff),
                   linewidth = 0.7, alpha = 0.9) +
      geom_point(data = centers_zr_1_top6, aes(cx, cy), colour = "black", size = 2) +
      geom_text(data = line_df_zr_top6, aes(mx, my, label = sprintf("%.2f\u00b0", ang_diff)),
                size = 3, vjust = -0.4, colour = "black") +
      scale_colour_gradient2(low = "firebrick", mid = "grey60", high = COL_2,
                             midpoint = 0, name = "delta angle\n(deg)") +
      coord_equal() +
      labs(title = "las_1 centre network (best 6 stems) - edge-wise angular differences to las_2",
           x = "X (m)", y = "Y (m)") +
      theme_bw(base_size = 11)
  )
}
```

## 4.8 Angular diff outlier filter

```{r}
#| label: params-ellipse-outlier-removal
#| code-summary: "Parameters - Iteratively remove the stem whose edges disagree most with the consensus (MAD on ang_diff)"


# A stem with a bad windowed centre corrupts EVERY edge it participates in, so
# outliers are removed one STEM at a time. Each round: flag edges whose ang_diff
# is > mad_thresh_zr MADs from the median, count how many flagged edges each stem
# touches, drop the worst-offending stem, and repeat until no edge is flagged.
#
# TIE-BREAK. When several stems touch the same number of flagged edges, the one
# whose own edges sit furthest from the consensus goes. That used to be decided
# by which removal most reduced sd(ang_diff), which is undefined once a
# candidate removal leaves fewer than two edges - on a small network every
# candidate scored NA, which.max() returned integer(0), and indexing the edge
# frame with a zero-length vector silently emptied it. A four-stem network then
# reported "1 stem removed, 0 kept". Mean absolute deviation from the median is
# defined for a single edge, so the tie-break can no longer collapse.
mad_thresh_zr    <- 4

# FLOOR. Below this many stems the network cannot support an outlier test at
# all: with 3 stems there are 3 edges and one bad stem touches two of them, so
# "the consensus" is a single edge. The loop stops rather than eating the set.
zr_min_stems     <- 3L

edges_zr         <- line_df_zr
removed_trees_zr <- character()
n_trees_start    <- length(unique(c(line_df_zr$TreeID_1_a, line_df_zr$TreeID_1_b)))
zr_stop_reason   <- "no outlying edge"

.par_apply("params-ellipse-outlier-removal")
```

```{r}
#| label: ellipse-outlier-removal
#| code-summary: "Iteratively remove the stem whose edges disagree most with the consensus (MAD on ang_diff)"

# Stems currently represented in the edge set.
.zr_stems <- function(e) unique(c(as.character(e$TreeID_1_a),
                                  as.character(e$TreeID_1_b)))

repeat {
  if (nrow(edges_zr) < 2L) { zr_stop_reason <- "fewer than 2 edges left"; break }

  .n_now <- length(.zr_stems(edges_zr))
  if (.n_now <= zr_min_stems) {
    zr_stop_reason <- sprintf("floor reached (%d stems, zr_min_stems = %d)",
                              .n_now, zr_min_stems)
    break
  }

  med_zr <- median(edges_zr$ang_diff); mad_val_zr <- mad(edges_zr$ang_diff)
  if (is.na(mad_val_zr) || mad_val_zr < 1e-10) {
    zr_stop_reason <- "MAD is zero - every edge agrees"; break
  }
  edges_zr$outlier <- abs(edges_zr$ang_diff - med_zr) / mad_val_zr > mad_thresh_zr
  if (!any(edges_zr$outlier)) { zr_stop_reason <- "no outlying edge"; break }

  tree_counts_zr <- table(c(as.character(edges_zr$TreeID_1_a[edges_zr$outlier]),
                            as.character(edges_zr$TreeID_1_b[edges_zr$outlier])))
  worst_zr <- names(tree_counts_zr)[tree_counts_zr == max(tree_counts_zr)]

  if (length(worst_zr) == 1L) {
    tree_to_remove_zr <- worst_zr
  } else {
    # Mean |ang_diff - median| over the edges that stem touches. Always defined
    # (a stem in the edge set touches at least one edge), so the selection below
    # always returns exactly one name.
    dev_zr <- vapply(worst_zr, function(tid) {
      touch <- as.character(edges_zr$TreeID_1_a) == tid |
               as.character(edges_zr$TreeID_1_b) == tid
      if (!any(touch)) return(-Inf)
      mean(abs(edges_zr$ang_diff[touch] - med_zr))
    }, numeric(1))
    tree_to_remove_zr <- worst_zr[order(-dev_zr, worst_zr)][1L]
  }

  # Belt and braces: a zero-length or NA name here would index the edge frame
  # with a zero-length vector and wipe it out without removing anything. If it
  # ever happens the loop stops and says so, rather than reporting an empty
  # network as a result.
  if (length(tree_to_remove_zr) != 1L || is.na(tree_to_remove_zr)) {
    zr_stop_reason <- "could not identify a single worst stem"; break
  }

  # Would this removal take the network below the floor? Then it is not a
  # removal worth making - keep the stem and stop.
  .keep <- !(as.character(edges_zr$TreeID_1_a) == tree_to_remove_zr |
             as.character(edges_zr$TreeID_1_b) == tree_to_remove_zr)
  .keep[is.na(.keep)] <- FALSE
  if (length(.zr_stems(edges_zr[.keep, , drop = FALSE])) < zr_min_stems) {
    zr_stop_reason <- sprintf("removing stem %s would leave fewer than %d stems",
                              tree_to_remove_zr, zr_min_stems)
    break
  }

  removed_trees_zr <- c(removed_trees_zr, tree_to_remove_zr)
  edges_zr <- edges_zr[.keep, , drop = FALSE]
}

clean_edges_zr   <- edges_zr
removed_edges_zr <- line_df_zr[line_df_zr$TreeID_1_a %in% removed_trees_zr |
                               line_df_zr$TreeID_1_b %in% removed_trees_zr, ]
n_trees_kept <- length(unique(c(clean_edges_zr$TreeID_1_a, clean_edges_zr$TreeID_1_b)))

# The inter-stem line filter removes STEMS, not edges, so the surviving pair set
# is every pair whose las_1 stem still appears in a clean edge. Recorded as a
# pair count so it lines up with every other number in the tally.
pairs_stage4_lines <- pairs_stage4[
  as.character(pairs_stage4$TreeID_1) %in%
    as.character(unique(c(clean_edges_zr$TreeID_1_a, clean_edges_zr$TreeID_1_b))), ,
  drop = FALSE]
.tally(n_stems_lines_start = n_trees_start,
       n_stems_lines_kept  = n_trees_kept,
       n_pairs_lines_kept  = nrow(pairs_stage4_lines))

cat(sprintf("Trees entering MAD filter:   %d\n", n_trees_start))
cat(sprintf("Trees removed by MAD filter: %d\n", length(removed_trees_zr)))
cat(sprintf("Trees kept after MAD filter: %d\n", n_trees_kept))
cat(sprintf("  stopped because: %s\n", zr_stop_reason))
if (length(removed_trees_zr) > 0)
  cat(sprintf("  Removed TreeIDs: %s\n", paste(removed_trees_zr, collapse = ", ")))

# entering - removed must equal kept. If it does not, stems left the network
# without being removed by name, which is the failure this chunk used to have.
if (n_trees_kept != n_trees_start - length(removed_trees_zr))
  warning(sprintf(paste0("[4.8] stem bookkeeping does not balance: %d entered, %d removed, ",
                         "%d kept. Edges were lost without a stem being named."),
                  n_trees_start, length(removed_trees_zr), n_trees_kept))
if (n_trees_kept == 0L)
  warning("[4.8] the angular-difference filter left NO stems - Stage 4's z-rotation cannot be solved from an empty edge set.")
cat(sprintf("Clean edges: %d | Removed edges: %d\n",
            nrow(clean_edges_zr), nrow(removed_edges_zr)))
```

### Vis - angular correction distribution

```{r}
#| label: fig-angular-diff-whisker
#| fig-cap: "Edge-wise angular differences after outlier removal. Green = kept, red = removed; black line = mean (the solved rotation)."

if (params$make_plots && nrow(clean_edges_zr) > 0) {
  print(
    ggplot() +
      geom_point(data = clean_edges_zr, aes(x = "delta angle", y = ang_diff),
                 colour = "#00AA00", size = 3, alpha = 1.0) +
      { if (nrow(removed_edges_zr) > 0)
          geom_point(data = removed_edges_zr, aes(x = "delta angle", y = ang_diff),
                     colour = "red", size = 1, alpha = 0.3) } +
      stat_summary(data = clean_edges_zr, aes(x = "delta angle", y = ang_diff),
                   fun.min = min, fun.max = max, geom = "errorbar", width = 0.15,
                   colour = "red", linewidth = 0.9, linetype = "dotted") +
      geom_hline(yintercept = mean(clean_edges_zr$ang_diff),
                 colour = "black", linetype = "dotted", linewidth = 1) +
      labs(title = "Angular differences (kept vs removed edges)",
           subtitle = "Green = kept, red = removed; black dotted = mean = solved rotation",
           x = "", y = "Angular difference (deg)") +
      theme_minimal(base_size = 12)
  )
}
```

## 4.9 Z rot solve

```{r}
#| label: z-only-rotation-solve
#| code-summary: "Solve the Z-only rotation as the mean of the cleaned edge-wise angular differences"

# The solve: a pure yaw rotates every edge by the same angle, so the best single
# estimate is the mean of the surviving (outlier-free) edge-wise differences,
# all measured from windowed centres. Recorded as z_rotation_deg / z_rotation_rad;
# Stage 5/6 apply it.
if (nrow(clean_edges_zr) == 0L)
  stop("Z-rot solve: no clean edges survived the outlier filter - cannot solve the rotation.")

z_rotation_deg <- mean(clean_edges_zr$ang_diff)
z_rotation_rad <- z_rotation_deg * pi / 180

rotation_summary_zr <- tibble::tibble(
  rotation_deg    = z_rotation_deg,
  rotation_rad    = z_rotation_rad,
  n_trees_start   = n_trees_start,
  n_trees_kept    = n_trees_kept,
  n_trees_removed = length(removed_trees_zr),
  n_edges_used    = nrow(clean_edges_zr),
  n_edges_removed = nrow(removed_edges_zr)
)

cat(sprintf("Z-only rotation solved: %.4f deg (%.6f rad)\n", z_rotation_deg, z_rotation_rad))
print(as.data.frame(rotation_summary_zr), row.names = FALSE)
```

### Export LAS - Post ZShift

```{r}
#| label: params-export-zshift-cloud
#| code-summary: "Parameters - Write las_1 with the full XYZ rotation + Stage 4 Z-translation baked in, to a .las snapshot"


# Stage 4 works on ellipse geometry and never loads point clouds at the top, so
# - like the 3D figure above - this export loads its own copy on demand and
# rebuilds the transform state. .vis_point_clouds returns xyz_1, the las_1 points
# with the rotation AND the Z-shift already applied. The Z-only rotation solved
# in 4.7 is NOT baked in here (it is applied in Stage 5/6); this snapshot is the
# post-translation cloud, matching where the ellipse geometry stands.
#
# export_zshift_cloud <- TRUE/FALSE master switch
# vis_load_clouds <- TRUE/FALSE read from disk vs reuse in-memory clouds
export_zshift_cloud <- TRUE
vis_load_clouds     <- TRUE

# NOTE: this default points at the OLD "111 Documents Artemis/124 Master Thesis"
# tree, not the current 1_Artemis/11_PaqLab one. Interactively it will fail on a
# missing directory until you repoint it.
output_path_zshift <- "C:/Users/ascsh/Documents/111 Documents Artemis/124 Master Thesis/Alignment/Correction_Matrix/2026-07-14_Axis/GZ/Export_Ztrans_BOS_F1_P1.las"

.par_apply("params-export-zshift-cloud")
```

```{r}
#| label: export-zshift-cloud
#| code-summary: "Write las_1 with the full XYZ rotation + Stage 4 Z-translation baked in, to a .las snapshot"

# Off in a batch unless asked for - same reasoning as Stage 3's rotated-cloud
# snapshot, and doubly so here: this export RELOADS both clouds from disk to
# rebuild the transform state, which is real time per pair for a file nobody
# opens. Stage 6 writes the aligned cloud that actually matters.
if (.in_batch()) {
  export_zshift_cloud <- isTRUE(.batch_opt(".batch_export_stage_clouds", FALSE))
  if (export_zshift_cloud)
    output_path_zshift <- file.path(.batch_opt(".batch_export_dir", dirname(output_path_zshift)),
                                    paste0(.batch_opt(".batch_sample", "sample"),
                                           "_Ztrans.las"))
}

if (isTRUE(export_zshift_cloud)) {
  .vpc_exp <- .vis_point_clouds(load = vis_load_clouds,
                                apply_rot = TRUE, apply_zshift = TRUE, need_las2 = FALSE)
  .write_las_snapshot(las_template = .vpc_exp$las_1, xyz_mat = .vpc_exp$xyz_1,
                      output_path = output_path_zshift,
                      stage_label = "4.8 xyz-rot + z-translation")
} else {
  cat("export-zshift-cloud: export_zshift_cloud = FALSE - skipping .las export.\n")
}
```

## Save/Load – Stage 4

```{r}
#| label: params-save-load_Stage4
#| code-summary: "Parameters - Save or load the complete Stage 4 object set"


# The Stage 4 OUTPUT path lives here (the load chunk reads only the Stage 3 path).
stage4_rds_path <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/SaveLoad/Stage4_BOS_F1_ai_manualcontinue.rds"

stage4_save <- TRUE
stage4_load <- FALSE

.par_apply("params-save-load_Stage4")
```

```{r}
#| label: save-load_Stage4
#| code-summary: "Save or load the complete Stage 4 object set"

# A batch decides these itself, whatever they are left set to above. Both are
# read a few lines below, in this same chunk, so the driver's overrides land
# too late to reach them - and a stage left switched to LOAD after an
# interactive session would otherwise make a batch write no checkpoint at all
# and say nothing about it. Saving is what a batch is for; loading is a no-op
# in one, since every stage runs in the same session.
if (.in_batch()) { stage4_save <- TRUE; stage4_load <- FALSE }

# Stage 4 checkpoint = everything the Stage 3.2 RDS carried (still in the
# environment after the load chunk) + Stage 4's own products. Snapshot the
# current environment's data objects (dropping functions and internal dot-names)
# and add the Stage 4 names explicitly, so this never drifts from upstream.
.stage4_new <- c(
  # 4.1 applied shift + toggle
  "t_trans_z_applied", "apply_mode_zshift", "custom_z_shift_stage4",
  # 4.2 shifted pivot
  "pivot2",
  # 4.2a shifted ellipses (full stem)
  "ellipses_zshift_1", "ellipses_zshift_2", "ellipses_zt_1_shifted",
  # 4.2b tilt diagnostic
  "stem_tilt_1", "stem_tilt_2",
  # 4.2 z-rotation window: params, per-pair overlap table, windowed subsets, kept pairs
  "zrot_bot", "zrot_top", "stage4_overlap_tbl",
  "ellipses_zrot_1", "ellipses_zrot_2", "pairs_stage4",
  # 4.3-4.4 solve inputs (centres from the windowed subsets)
  "pairs_zrot_ok", "avg_centres_1", "avg_centres_2", "pairs_zrot_valid", "line_df_zr",
  # 4.5 outlier filter products
  "clean_edges_zr", "removed_edges_zr", "removed_trees_zr", "mad_thresh_zr",
  "pairs_stage4_lines",
  # 4.7 the solved rotation - the stage's headline output
  "z_rotation_deg", "z_rotation_rad", "rotation_summary_zr"
)

# carry forward whatever came in from Stage 3.2 (data objects only, no helpers)
.carried <- setdiff(ls(envir = .stage_env()), lsf.str(envir = .stage_env()))
.carried <- .carried[!startsWith(.carried, ".")]

.stage4_required <- unique(c(.carried, .stage4_new))
.stage4_optional <- character(0)

if (stage4_save) {
  .stage_save(stage4_rds_path, .stage4_required, .stage4_optional, label = "Stage 4")
}

if (stage4_load) {
  .stage_load(stage4_rds_path, .stage4_required, .stage4_optional, label = "Stage 4")
  cat(sprintf("\nStage 4 restored: Z-shift %+.4f m | Z-rotation %+.4f deg | %d pair(s)\n",
              t_trans_z_applied, z_rotation_deg, nrow(pairs_stage4)))
  if (exists(".report_pivot1")) .report_pivot1()
  cat(sprintf("pivot2: (%.4f, %.4f, %.4f)\n", pivot2[1], pivot2[2], pivot2[3]))
}
```

Next step

Run `Stage5_ZRotationXY.qmd`. It loads this checkpoint, rebuilds each stem's
vertical profile from the shifted ellipse centres, and solves the Z-only
rotation and the residual XY translation.
