---
title: "Test Sample Generator - Known-Truth Misalignment"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
params:
  make_plots: true
---

# Overview - Test Sample Generator

Every stage of the alignment pipeline reports what it *solved*. None of them can
report what it got *wrong*, because on real data there is nothing to check the
answer against. This document manufactures that missing reference.

The input is a single **reference file**, `las_1`, already known to be correctly
aligned - by construction or by careful manual work. A known transform is then
applied to a copy of `las_1`, deliberately knocking it out of alignment by an
amount that is recorded, while the untouched `las_1` remains the reference. The
result is a new `.las` that looks like
an ordinary unaligned sample to the pipeline, and a `.csv` that says exactly what
the pipeline is supposed to find.

**One run produces a sweep, not a sample.** Give it a list of reference files and
a vector of tilt magnitudes and it generates every combination - one `.las`, one
forward matrix, one backward matrix and one truth row per (reference x tilt) -
plus a single consolidated Reference CSV covering every sample in the run and a
ready-made `las_pairs` for the batch driver. Everything else about the forward
transform is held constant, so the tilt is the only variable and any change in
the pipeline's error across the sweep is attributable to it.

**The direction of the numbers matters and is the easiest thing to get wrong.**
The parameters entered below are the *forward* transform - the damage done to a
correct cloud. What the pipeline has to solve for is the *inverse* - the repair.
They are not the same numbers, and for the rotation they are not even the same
numbers with a minus sign in front: inverting a tilt-plus-yaw rotation turns the
tilt direction around by roughly 180 degrees *and* shifts it by the yaw. Both
sets are computed and both are written to the CSV, under names that say which is
which. **The `target_*` columns are the ones to score the pipeline against.**

The eight error terms this exists to support

Once the pipeline has run on the generated sample, each of these is a
subtraction against a `target_*` column:

| Error term | Compared against |
|---|---|
| full XYZ rotation | `target_rot_total_deg` (and the full matrix `target_R_*`) |
| XY-only rotation (tilt) | `target_rot_xy_mag_deg`, `target_rot_xy_dir_deg` |
| Z-only rotation (yaw) | `target_rot_z_deg` |
| X-only shift | `target_shift_x_m` |
| Y-only shift | `target_shift_y_m` |
| Z-only shift | `target_shift_z_m` |
| XY shift | `target_shift_xy_m` |
| XYZ shift | `target_shift_xyz_m` |

Rotation errors are **not** taken as differences of the reported angles - that is
wrong whenever more than one axis is involved, because angles about different
axes do not subtract. They are taken from the residual rotation
$R_{\text{err}} = R_{\text{recovered}} R_{\text{target}}^{\top}$, which is the
rotation still left over after the pipeline's answer has been applied. Its total
angle, its tilt and its yaw are the three rotation errors. Stage 6 does this.

The pivot travels with the sample

A rotation is meaningless without the point it turned about, and the pipeline
can use the shifted  pivot point used here.  the shifted pivot is written into the exported cloud
as a single extra point carrying a `PivotPoint` scalar field set to 1. It is
visible in CloudCompare, it survives the round trip, and it lets any later
comparison be re-expressed about the correct origin. It is also written to the
CSV, which is the copy to trust - a LAS stores coordinates as scaled integers, so
the point in the cloud is quantised to the file's scale factor.

The marker box - a reference made of geometry rather than of matrices

The eight terms above are all read off a transform. That is a statement about a
matrix, and a matrix is only as meaningful as the frame and the pivot it is
expressed in. So this generator also plants a physical reference in the cloud:
the eight vertices of the reference cloud's bounding box, written into the
sample **twice**.

| `TargetBox` | `BoxCorner` | what it is |
|---|---|---|
| 1 | 1-8 | the eight corners where they belong - never transformed |
| 2 | 1-8 | the same eight corners, put through the forward transform |

Corner *n* of box 2 is the same corner as corner *n* of box 1, so the two boxes
difference without any nearest-neighbour guesswork. Once the pipeline has run,
Stage 6 applies the transform it solved to the box-2 corners; a perfect answer
lands each one exactly on its box-1 twin. What is left is a displacement in
metres, measured on the geometry, with no pivot convention anywhere in it. That
is what Stage 6 now reports as the primary shift residual - the matrix-based
terms above are still computed and still written, beside it.

Order of operations

1. shift `las_1` by `(dx, dy, dz)`
2. move the pivot by the same `(dx, dy, dz)`
3. compose the rotation from the tilt and yaw parameters
4. rotate the shifted cloud about the shifted pivot

Shift first, then rotate about the moved pivot. Written out:

$$p' = R\,\big(p + t - c_{\text{shifted}}\big) + c_{\text{shifted}},
\qquad c_{\text{shifted}} = c + t$$

Because the pivot is carried along by the shift, the two operations do not
interfere: the rotation sees the cloud in exactly the same relative position to
its pivot as before the shift. That is what makes the inverse clean.

# Stage 0 - Setup

## 0.1 Load libraries

```{r}
#| label: libraries
#| code-summary: "Load libraries and set the plot theme"

suppressPackageStartupMessages({
  library(lidR)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(plotly)
})
```

## 0.2 Global Helpers

```{r}
#| label: global-helpers
#| code-summary: "Colours, ggplot theme, non-overwriting paths, rotation builders and the exact decomposer"

# ============================================================
# Colours
# ============================================================
COL_1 <- "#4C72B0"   # las_1 - the reference cloud (also the one that gets damaged)
COL_P <- "#BF9ACA"   # the pivot point
COL_B <- "#C44E52"   # the truth marker box (TargetBox = 1)

# ============================================================
# ggplot theme
# ============================================================
theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title       = element_text(face = "bold"),
          strip.text       = element_text(face = "bold"))
)

# ============================================================
# Non-overwriting output paths
# Identical in behaviour to the pipeline's .unique_path: a write never
# clobbers an existing file, it appends _001, _002, ... until the name is
# free. Repeated here rather than sourced so this document stands alone.
# ============================================================
.unique_path <- function(path) {
  if (!file.exists(path)) return(path)
  base_noext <- tools::file_path_sans_ext(path)
  ext        <- tools::file_ext(path)
  counter    <- 1L
  repeat {
    candidate <- sprintf("%s_%03d.%s", base_noext, counter, ext)
    if (!file.exists(candidate)) {
      cat(sprintf("Target already exists - using unique name: %s\n", candidate))
      return(candidate)
    }
    counter <- counter + 1L
  }
}

# ============================================================
# Rotation builders
#
# THE TILT BUILDER IS COPIED FROM THE PIPELINE, deliberately and byte for
# byte (Stage 3, .tilt_rotation_matrix). If this document built its tilt a
# different way, every truth value it wrote would be expressed in a
# convention the pipeline does not use, and the errors it produced would be
# measuring the mismatch between two conventions rather than the pipeline's
# accuracy. If the pipeline's builder is ever edited, edit this one too.
#
# Convention, verified rather than assumed: for R = .tilt_rotation_matrix(m, d),
#   R %*% c(0, 0, 1) = (sin m cos d, sin m sin d, cos m)
# so the tilt magnitude is exactly m and the azimuth of the rotated +Z axis is
# exactly d, CCW from +X. That is the same quantity the pipeline's own
# .rot_specs() reports as tilt / tilt_dir, so a value entered here comes back
# out of the pipeline's reports under the same name and meaning.
#
# (The pipeline's comment on this function claims points lying in direction_deg
# move in +Z. They move in -Z; the +Z AXIS is what tips toward direction_deg.
# The matrix is right, the comment is not - and it is the matrix that both
# documents share, so nothing downstream is affected.)
# ============================================================
.tilt_rotation_matrix <- function(magnitude_deg, direction_deg) {
  m <- magnitude_deg * pi / 180
  d <- direction_deg * pi / 180
  axis <- c(-sin(d), cos(d), 0)          # horizontal axis, perpendicular to direction
  ax <- axis[1]; ay <- axis[2]; az <- axis[3]
  K <- matrix(c(  0, -az,  ay,
                 az,   0, -ax,
                -ay,  ax,   0), nrow = 3, byrow = TRUE)
  diag(3) + sin(m) * K + (1 - cos(m)) * (K %*% K)
}

# Yaw: a rotation of `deg` about +Z, CCW seen from above, matching the sign
# convention Stage 5 / Stage 6 use for z_rotation_deg.
.yaw_rotation_matrix <- function(deg) {
  a <- deg * pi / 180
  R <- diag(3)
  R[1:2, 1:2] <- matrix(c(cos(a), -sin(a),
                          sin(a),  cos(a)), nrow = 2, byrow = TRUE)
  R
}

# ============================================================
# The composition order, and why this one
#
#     R = tilt(m, d) %*% yaw(a)          -> yaw applied FIRST, then tilt
#
# Not an arbitrary pick. Under this order the third column of R is
# tilt(m, d) %*% c(0,0,1) untouched by the yaw, so m and d read straight back
# off the finished matrix and the numbers entered in Step 0.4 are literally
# the numbers the matrix carries. Under the other order (tilt first, then yaw)
# the yaw rotates the tilt direction, and a matrix built from d = 20 reports a
# tilt direction of d + a. Both are valid rotations; only one lets a user
# check the composed matrix against what they typed.
# ============================================================
.compose_rotation <- function(xy_mag_deg, xy_dir_deg, z_deg) {
  .tilt_rotation_matrix(xy_mag_deg, xy_dir_deg) %*% .yaw_rotation_matrix(z_deg)
}

# ============================================================
# EXACT decomposition, the inverse of .compose_rotation
#
# Given any rotation R, returns the unique (m, d, a) with
# R = tilt(m, d) %*% yaw(a). This is a true factorisation, not a description:
# the three numbers rebuild R to machine precision, and the caller can check
# that (rebuild_err below). That distinction matters for the inverse
# transform - a set of angles that merely DESCRIBES the repair is not something
# anyone can apply.
#
#   m = acos(R[3,3])                third column is tilt(m,d) %*% z
#   d = atan2(R[2,3], R[1,3])
#   a = the residual yaw, read off tilt(m,d)^T %*% R
#
# Degenerate case: at m = 0 the tilt direction is undefined (there is no tilt
# to point anywhere). d is reported as NA rather than as the 0 that atan2(0,0)
# happens to return, because a direction of "0 degrees" and "no direction" are
# different claims and only one of them is true.
# ============================================================
.decompose_rotation <- function(R, dir_tol_deg = 1e-9) {
  m <- acos(max(-1, min(1, R[3, 3]))) * 180 / pi
  d <- if (m < dir_tol_deg) NA_real_ else (atan2(R[2, 3], R[1, 3]) * 180 / pi) %% 360
  M <- .tilt_rotation_matrix(m, if (is.na(d)) 0 else d) %*% .yaw_rotation_matrix(0)
  M <- t(M) %*% R
  a <- atan2(M[2, 1], M[1, 1]) * 180 / pi
  total <- acos(max(-1, min(1, (sum(diag(R)) - 1) / 2))) * 180 / pi

  rebuild <- .compose_rotation(m, if (is.na(d)) 0 else d, a)
  list(xy_mag_deg  = m,
       xy_dir_deg  = d,
       z_deg       = a,
       total_deg   = total,
       rebuild_err = max(abs(rebuild - R)))
}

# ============================================================
# Flatteners and the matrix writer
# Used by the per-sample generator in Stage T. Defined here, once, because a
# helper defined inside the loop is a helper redefined 28 times.
# ============================================================

# Matrices flattened to R_fwd_11 ... R_fwd_33 in ROW-major order, so they read
# back with a single matrix(byrow = TRUE).
.flat <- function(M, prefix) {
  nm <- as.vector(t(outer(seq_len(nrow(M)), seq_len(ncol(M)),
                          function(i, j) sprintf("%s_%d%d", prefix, i, j))))
  stats::setNames(as.list(as.vector(t(M))), nm)
}

# The marker corners: box1_c1_x ... box2_c8_z. These are the EXACT coordinates -
# the copies in the .las went through the file's scale factor and are quantised.
# Stage 6 reads these in preference to the cloud's.
.flat_box <- function(M, prefix) {
  nm <- sprintf("%s_c%d_%s", prefix,
                rep(seq_len(nrow(M)), each = 3L), rep(c("x", "y", "z"), nrow(M)))
  stats::setNames(as.list(as.vector(t(M))), nm)
}

# Full double precision (17 significant digits, so the file round-trips
# exactly), space-separated, no header - the same shape as the TLS-registration
# matrix files elsewhere in the pipeline.
.write_matrix_txt <- function(M, path) {
  utils::write.table(format(M, digits = 17, scientific = FALSE), path,
                     sep = " ", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

# A rotation matrix, or not. Catches a scale, a shear or a reflection before
# it is baked into a sample that will then be used to judge the pipeline.
.validate_rotation <- function(R, nm = "R") {
  if (!is.matrix(R) || !identical(dim(R), c(3L, 3L)))
    stop(sprintf("[%s] is not a 3x3 matrix.", nm))
  if (!isTRUE(all.equal(t(R) %*% R, diag(3), tolerance = 1e-9, check.attributes = FALSE)))
    stop(sprintf("[%s] is not a rotation (R'R != I) - a scale or shear here would resize the stems.", nm))
  if (det(R) < 0)
    stop(sprintf("[%s] has determinant %.6f - a reflection, not a rotation.", nm, det(R)))
  invisible(TRUE)
}
```

## 0.3 Reference Files and the Tilt Sweep

One run generates **one sample per (reference file × tilt magnitude)**. Four
references and seven tilts is twenty-eight samples, each with its own `.las`,
its own pair of matrix files, its own truth row, and a name that says which
combination it is.

The seven identifier fields are read off each **reference** cloud's filename
rather than typed in beside it. The reference file already carries them:

```
R_SNBG_F1_a_2020_2024_TLS_TLS.las
│  │    │  │  │    │    │   └── Scanner_2
│  │    │  │  │    │    └────── Scanner_1
│  │    │  │  │    └─────────── Epoch_2
│  │    │  │  └──────────────── Epoch_1
│  │    │  └─────────────────── Sample_Type_ID
│  │    └────────────────────── Cluster_ID
│  └─────────────────────────── Sample_location_ID
└────────────────────────────── R = reference
```

and each generated sample is a damaged copy of that reference, so it takes the
same seven fields under an `I_` prefix - `I` for input, the thing the pipeline
is handed - plus the tilt magnitude that was applied to it, inserted after
`Sample_Type_ID`:

```
R_SNBG_F1_a_2020_2024_TLS_TLS.las         one reference, untouched
├── I_SNBG_F1_a_0p00_2020_2024_TLS_TLS.las
├── I_SNBG_F1_a_1p04_2020_2024_TLS_TLS.las
├── I_SNBG_F1_a_29p71_2020_2024_TLS_TLS.las
└── ...                     └── tg_rot_xy_mag, degrees off vertical
```

The tilt is the one field that is *not* a property of the reference file - it
is a property of the generation run, and it is the dial being swept. Putting it
in the filename is what lets a directory of samples off one reference be told
apart at a glance, and sorted by difficulty without opening any of them.

Two consequences worth stating. The reference name has seven fields and a sample
name has eight, so they do not round-trip: a sample name is built from a
reference's fields, never parsed back into one. And the tilt is written `45p59`
rather than `45.59`, because a period in a basename invites every tool that
splits on it to find an extension where there is none.

**Everything is validated before anything is generated.** Every reference file
is checked for existence and parsed, and the full job table is built and checked
for duplicate names, before the first cloud is read. A run of twenty-eight
samples should not discover on sample nineteen that a filename does not parse.

```{r}
#| label: sample-inputs
#| code-summary: "The reference clouds, and how their names are read"
#| code-fold: false

# ── Reference clouds: ALREADY KNOWN TO BE CORRECTLY ALIGNED ──────────────────
# Each is loaded once, left untouched on disk, used for its own marker box, and
# is the file whose name supplies the seven identifier fields. A copy of it is
# what gets damaged to build each generated sample.
reference_dir <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/Samples"

reference_las_names <- c(
  "R_SNBG_F1-F1006-F1125_a_2020_2024_TLS_TLS"
)
reference_las_paths <- file.path(reference_dir, paste0(reference_las_names, ".las"))

# ── Where the identifiers come from ──────────────────────────────────────────
# "parse"  -> split them out of each reference's basename. The normal mode.
# "manual" -> use the seven values below for EVERY reference, which only makes
#             sense with a single reference file. For one not yet named to the
#             convention.
sample_name_mode <- "parse"

reference_prefix <- "R"   # stripped from the front of a reference basename
sample_prefix    <- "I"   # I for input: what the pipeline is handed

# Used only when sample_name_mode is "manual". Left in the parsed order so the
# two modes cannot disagree about which field is which.
manual_ids <- list(
  Sample_location_ID = "SNBG",
  Cluster_ID         = "F1",
  Sample_Type_ID     = "a",
  Epoch_1            = "2020",
  Epoch_2            = "2024",
  Scanner_1          = "TLS",
  Scanner_2          = "TLS"
)

id_order <- c("Sample_location_ID", "Cluster_ID", "Sample_Type_ID",
              "Epoch_1", "Epoch_2", "Scanner_1", "Scanner_2")

#' The seven fields from one reference basename.
#'
#' Returns a named list in id_order. Stops rather than falling back: a fallback
#' would name a sample after a file it did not come from, and that name goes
#' into the truth CSV and from there into every downstream report.
.parse_ids <- function(path) {
  if (identical(sample_name_mode, "manual")) return(manual_ids[id_order])

  ref_base <- tools::file_path_sans_ext(basename(path))

  # The prefix is stripped only if it is actually there, so a reference already
  # named bare parses the same way. A file named "Reference_..." is not silently
  # mistaken for a prefix plus six fields - the field count catches it.
  stem  <- sub(sprintf("^%s_", reference_prefix), "", ref_base)
  parts <- strsplit(stem, "_", fixed = TRUE)[[1]]

  if (length(parts) != length(id_order))
    stop(sprintf(paste0("[0.3] A reference filename does not parse into %d fields.\n",
                        "  file    : %s\n  stem    : %s\n  found   : %d field(s): %s\n",
                        "  expected: %s\n",
                        "  Rename it to %s_<%s>.las, or set sample_name_mode <- \"manual\"."),
                 length(id_order), ref_base, stem, length(parts),
                 paste(parts, collapse = ", "),
                 paste(id_order, collapse = ", "),
                 reference_prefix, paste(id_order, collapse = ">_<")),
         call. = FALSE)

  # An empty field collapses two underscores into a name that no longer splits
  # into its parts; a space inside one breaks it differently. Flagged rather
  # than corrected - renaming a sample silently is worse than a warning.
  if (any(!nzchar(parts)) || any(grepl("[[:space:]]", parts)))
    warning(sprintf(paste0("Reference '%s' has an empty or space-containing field - the ",
                           "sample name will not split back into its parts. The CSV ",
                           "columns still will."), ref_base), call. = FALSE)

  stats::setNames(as.list(parts), id_order)
}

if (!identical(sample_name_mode, "parse") && !identical(sample_name_mode, "manual"))
  stop('sample_name_mode must be "parse" or "manual".')

# ── Fail fast: every file, before any generation ─────────────────────────────
.miss <- reference_las_paths[!file.exists(reference_las_paths)]
if (length(.miss))
  stop("These reference file(s) do not exist:\n  ", paste(.miss, collapse = "\n  "))

cat(sprintf("%d reference file(s):\n", length(reference_las_paths)))
for (.p in reference_las_paths)
  cat(sprintf("  %-36s -> %s\n", basename(.p),
              paste(unlist(.parse_ids(.p)), collapse = " | ")))   # stops here on a bad name
rm(.miss, .p)
```

## 0.4 Input Parameters

```{r}
#| label: input-parameters
#| code-summary: "Output directories, pivot mode, marker box, and the forward transform to apply"
#| code-fold: false

# ── Output directories, one per kind of artefact ─────────────────────────────
las_output_dir    <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/Samples"
csv_output_dir    <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/ReferenceCSV"
matrix_output_dir <- "C:/Users/ascsh/Documents/1_Artemis/11_PaqLab/115_TreeTimeAlign/TestSampleGen/ForwardBackward"

# ── Consolidated outputs, in csv_output_dir ──────────────────────────────────
# ONE file covering every reference and every tilt in the run. This is the file
# to hand the batch driver as truth_csv_path: it keys on sample_name and holds
# one row per sample, which is exactly the shape RunR splits back apart per
# pair. The per-sample CSVs below are a convenience, not the deliverable.
# The name is yours to set. It names a SWEEP, not a sample - every reference and
# every tilt in the run lands in this one file - so it is named after the
# reference set rather than after any row in it.
consolidated_csv_name  <- "SNBG_F1-F1006-F1125_a_2020_2024_TLS_TLS.csv"
consolidated_truth_csv <- file.path(csv_output_dir, consolidated_csv_name)

# One row per sample, written beside the consolidated file. OFF: the driver now
# resolves every path it needs from the consolidated table, so twenty-eight
# extra single-row files were twenty-eight more things to keep in step with it.
# Turn it on if you need to hand one sample to something on its own - Stage 1's
# "csv" pivot mode, for instance, wants a single-row file.
export_per_sample_csv <- FALSE

# Also write an untransformed copy of each reference beside its samples, so a
# pair can be handed to the pipeline as a self-contained unit. Written once per
# reference, not once per tilt.
export_reference_copy <- FALSE

# Sample names encode the tilt, so two runs producing the same name are
# producing the same sample. TRUE overwrites it; FALSE parks the second at _1,
# which across a sweep leaves a directory of near-duplicates whose names no
# longer identify them uniquely - and a consolidated CSV whose sample_name
# column no longer matches the files beside it.
overwrite_existing <- TRUE

# ── Pivot ────────────────────────────────────────────────────────────────────
# "auto"   -> the centre of the bounding box of the UNSHIFTED reference. This is
#             the same definition Stage 1 uses for pivot1 (the mid-point of the
#             cloud's EXTENT in all three axes, not the centroid of its points),
#             so the generated sample is turned about the same kind of point the
#             pipeline will later turn it back about. Note the Stage 1 caveat
#             applies here too: the extent is sensitive to a single stray return.
# "custom" -> pivot_custom below, in world coordinates. Used for EVERY
#             reference, which rarely makes sense with more than one.
pivot_mode   <- "auto"
pivot_custom <- c(395295.465375, 5309905.28875, 266.084)

# ── The forward transform: the damage to inflict ─────────────────────────────
# These are NOT what the pipeline has to solve for. They are the opposite of it.
# T.7 computes the targets.
#
# Shift, in metres, applied in world coordinates. Constant across the sweep: the
# point of the sweep is to vary the tilt and nothing else, so that any change in
# the pipeline's error is attributable to the tilt.
tg_shift_x <- +0.029
tg_shift_y <- +90.345
tg_shift_z <- +15.101

# Tilt (the "XY rotation"): how far off vertical, and which way.
#   magnitude - degrees away from vertical. A VECTOR: one sample per entry, per
#               reference file. A single value still works - it is a sweep of
#               length one.
#   direction - azimuth the +Z axis tips TOWARD, degrees CCW from +X, [0, 360)
tg_rot_xy_mag <- c(0, 45, 90, 135, 180)
tg_rot_xy_dir <- 20.85

# Yaw (the "Z rotation"): degrees about the vertical axis, CCW seen from above.
tg_rot_z <- 8.2400

# tg_rot_xy_mag also goes into the output filename, after Sample_Type_ID. Two
# decimals is enough to tell a swept series apart without making the name
# unreadable; the CSV carries the value at full precision either way, so nothing
# depends on this being exact.
tilt_tag_digits <- 2

# ── The pivot point written into the cloud ───────────────────────────────────
# TreeID is left empty on that point so no stem-wise step can ever pick it up as
# a stem. NA is the honest value; if the writer or a downstream reader rejects
# NA in that field, 0L is the fallback - it is the value the pipeline already
# treats as "not a tree".
pivot_treeid_value <- NA_integer_
pivot_field_name   <- "PivotPoint"

# ── The marker box written into the cloud ────────────────────────────────────
# Eight points at the vertices of a bounding box, written into the sample twice:
# once where they belong (TargetBox = 1, never transformed) and once put through
# the forward transform (TargetBox = 2). BoxCorner carries 1..8 on both copies.
#
# The box is built from the bounding box of the untransformed reference - the
# frame the pipeline is trying to get back to - once per reference, so every
# tilt off that reference is scored against the same eight points.
#
# The box is a bounding box, not a fit: a single stray return stretches it. That
# does not matter here. The corners are not measurements of anything, they are
# arbitrary points chosen to be far apart and unambiguously labelled - what is
# measured is how far each one moves, and any eight fixed points would serve.
# Spread is the only property that counts, because a rotation error shows up as
# displacement in proportion to the lever arm.
box_source_cloud  <- "reference"     # the only cloud there is
box_field_name    <- "TargetBox"     # 0 = ordinary point, 1 = truth, 2 = moved
corner_field_name <- "BoxCorner"     # 1..8 on the marker points, 0 elsewhere
box_treeid_value  <- NA_integer_     # as with the pivot: never a stem
```

## 0.4b Build the Job Table

Every sample this run will produce, named and checked, before any cloud is read.
The table is the cartesian product of the references and the tilts.

```{r}
#| label: build-job-table
#| code-summary: "reference x tilt, with the output name for each"

# The tag has to survive being put in a filename and read back, so: a period
# becomes "p" (a period in a basename invites tools to find an extension where
# there is none) and a minus becomes "m" (a leading dash is a command-line
# option on every platform that has a command line). A tilt magnitude should
# never be negative - .decompose_rotation returns it in [0, 180] - but the name
# is built before anything has validated that, so it is handled rather than
# assumed.
.tilt_tag <- function(mag) {
  s <- formatC(mag, format = "f", digits = tilt_tag_digits)
  gsub("-", "m", gsub(".", "p", s, fixed = TRUE), fixed = TRUE)
}

jobs <- do.call(rbind, lapply(seq_along(reference_las_paths), function(r) {
  ids <- .parse_ids(reference_las_paths[r])
  do.call(rbind, lapply(seq_along(tg_rot_xy_mag), function(k) {
    tag <- .tilt_tag(tg_rot_xy_mag[k])
    data.frame(
      ref_index   = r,
      tilt_index  = k,
      ref_path    = reference_las_paths[r],
      # I_<location>_<cluster>_<type>_<tilt>_<epoch1>_<epoch2>_<scanner1>_<scanner2>
      sample_name = paste(c(sample_prefix,
                            ids$Sample_location_ID, ids$Cluster_ID,
                            ids$Sample_Type_ID, tag,
                            ids$Epoch_1, ids$Epoch_2,
                            ids$Scanner_1, ids$Scanner_2),
                          collapse = "_"),
      tilt_mag    = tg_rot_xy_mag[k],
      tilt_tag    = tag,
      as.data.frame(ids, stringsAsFactors = FALSE),
      stringsAsFactors = FALSE
    )
  }))
}))

# ── Duplicate names ──────────────────────────────────────────────────────────
# Two jobs sharing a name would overwrite each other's .las and put two rows
# with the same sample_name into the consolidated CSV - the one column every
# downstream join keys on. Caught here, where the cause is visible, rather than
# as a mysteriously short output directory.
.dupes <- unique(jobs$sample_name[duplicated(jobs$sample_name)])
if (length(.dupes))
  stop(sprintf(paste0("%d job(s) share a name:\n  %s\n",
                      "Two tilts that round to the same %d-decimal tag collide. ",
                      "Raise tilt_tag_digits, or remove the duplicate tilt."),
               length(.dupes), paste(.dupes, collapse = "\n  "), tilt_tag_digits))

# Rounding that does not collide is still worth knowing about: the name will
# claim a tilt the sample does not exactly have.
.rounded <- tg_rot_xy_mag[
  abs(as.numeric(vapply(tg_rot_xy_mag,
                        function(m) formatC(m, format = "f", digits = tilt_tag_digits),
                        character(1))) - tg_rot_xy_mag) > 1e-9]

cat(sprintf("%d job(s): %d reference file(s) x %d tilt(s)\n",
            nrow(jobs), length(reference_las_paths), length(tg_rot_xy_mag)))
cat(sprintf("  tilts: %s deg\n",
            paste(formatC(tg_rot_xy_mag, format = "f", digits = 3), collapse = ", ")))
if (length(.rounded))
  cat(sprintf(paste0("  note: %s deg round to a shorter name tag (%d dp) - the name claims a\n",
                     "        tilt the sample does not exactly have. The CSV keeps full precision.\n"),
              paste(.rounded, collapse = ", "), tilt_tag_digits))
cat(sprintf("  shift (constant across the sweep): (%+.3f, %+.3f, %+.3f) m | %.3f m\n",
            tg_shift_x, tg_shift_y, tg_shift_z,
            sqrt(tg_shift_x^2 + tg_shift_y^2 + tg_shift_z^2)))
cat(sprintf("  yaw   (constant across the sweep): %.4f deg\n", tg_rot_z))
cat("\n")
print(jobs[, c("sample_name", "tilt_mag")], row.names = FALSE)

# ── Which job gets the full narration ────────────────────────────────────────
# Printing every check for twenty-eight samples buries the one that matters. One
# job is run verbosely, with all the per-step output and the 3D figure; the rest
# report one line each. Every check still RUNS on every job - only the printing
# is suppressed.
detail_job <- 1L
if (detail_job < 1L || detail_job > nrow(jobs))
  stop(sprintf("detail_job must be between 1 and %d.", nrow(jobs)))
cat(sprintf("\nFull narration for job %d: %s\n", detail_job, jobs$sample_name[detail_job]))

rm(.dupes, .rounded)
```

# Stage T - Build the Test Samples

Everything from the pivot to the truth row now lives in one function, called
once per job. There is one implementation, not one for the single-sample case
and another for the sweep, because two implementations of the same transform
drift apart and the drift shows up as a truth file that is quietly wrong.

**One reordering from the single-sample version:** the inverse transform and all
its checks now run *before* anything is written. In a sweep, a sample that fails
its round-trip check would otherwise already be on disk - a file that looks
ordinary with no way to know its truth values are wrong. Now nothing is written
until every check has passed, so the presence of a file is itself the guarantee.

The work that depends only on the reference cloud - loading it, the marker box,
the pivot - is done once per reference in the driver below. Only `R_fwd` and
what follows from it varies with the tilt.

## T.0 The Per-Sample Generator

```{r}
#| label: generate-sample-fn
#| code-summary: "T.1 - T.8 for one job: pivot, shift, rotate, markers, invert, check, write"

#' Generate one test sample.
#'
#' @param job one row of `jobs`
#' @param las_ref the reference cloud, already loaded
#' @param box_xyz_1 8x3 truth corners, built once per reference
#' @param pivot_base the unshifted pivot, resolved once per reference
#' @param verbose print every step's checks, as the single-sample document did
#' @param keep_coords return the before/after coordinates, for the figure. Off
#'   by default: twenty-eight clouds held at once is not a memory budget.
.generate_sample <- function(job, las_ref, box_xyz_1, pivot_base,
                             verbose = FALSE, keep_coords = FALSE) {

  .v <- function(...) if (isTRUE(verbose)) cat(...)
  box_corner_id <- seq_len(nrow(box_xyz_1))

  # ── T.1 Pivot ──────────────────────────────────────────────────────────────
  tg_shift <- c(tg_shift_x, tg_shift_y, tg_shift_z)

  # The pivot is a point in the cloud's frame, so when the cloud moves it moves.
  # Rotating the shifted cloud about the UNSHIFTED pivot would be a rotation
  # about a point that is no longer where it was in the geometry - the rotation
  # would then carry a translation of its own, that translation would fold
  # silently into the shift, and the "known" shift written to the CSV would no
  # longer be the shift the sample actually has.
  pivot_shifted <- pivot_base + tg_shift

  .v(sprintf("Pivot (base) : (%.4f, %.4f, %.4f)\n", pivot_base[1], pivot_base[2], pivot_base[3]))
  .v(sprintf("Shift applied: (%+.4f, %+.4f, %+.4f) m\n", tg_shift[1], tg_shift[2], tg_shift[3]))
  .v(sprintf("Pivot (moved): (%.4f, %.4f, %.4f)  <- the rotation origin\n",
             pivot_shifted[1], pivot_shifted[2], pivot_shifted[3]))

  # ── T.2 Apply the shift ────────────────────────────────────────────────────
  xyz_base    <- as.matrix(las_ref@data[, c("X", "Y", "Z")])
  xyz_shifted <- sweep(xyz_base, 2, tg_shift, `+`)

  # ── T.3 Compose the rotation ───────────────────────────────────────────────
  R_fwd <- .compose_rotation(job$tilt_mag, tg_rot_xy_dir, tg_rot_z)
  .validate_rotation(R_fwd, "R_fwd")

  # Read the finished matrix back through the decomposer and check it returns
  # the inputs. This is a real test, not a tautology: the composer and the
  # decomposer are written independently, and if the two ever fall out of step -
  # a sign flip, a changed order - this catches it here, rather than three
  # samples later when the errors all come out suspiciously large in one
  # direction.
  chk <- .decompose_rotation(R_fwd)
  .v("\n── Forward rotation ──────────────────────────────────────────────────────\n")
  .v(sprintf("  composition order  : tilt(mag, dir) %%*%% yaw(z)  (yaw first, then tilt)\n"))
  .v(sprintf("  XY rotation mag    : %8.4f deg   (entered %8.4f)\n", chk$xy_mag_deg, job$tilt_mag))
  .v(sprintf("  XY rotation dir    : %8.4f deg   (entered %8.4f)\n", chk$xy_dir_deg, tg_rot_xy_dir %% 360))
  .v(sprintf("  Z rotation (yaw)   : %8.4f deg   (entered %8.4f)\n", chk$z_deg, tg_rot_z))
  .v(sprintf("  full XYZ magnitude : %8.4f deg   (about the matrix's own axis)\n", chk$total_deg))
  .v(sprintf("  round-trip error   : %.2e\n", chk$rebuild_err))

  # At a tilt of exactly 0 there is no tilt to point anywhere, so the decomposer
  # returns NA for the direction rather than the 0 that atan2(0,0) happens to
  # give. The magnitude and yaw are still checked; the direction check is
  # skipped, because a value undefined by construction is not a disagreement.
  # A tilt of exactly 180 is the other degenerate case - the +Z axis lands on
  # -Z and every direction reaches it - and the same reasoning applies.
  tol <- 1e-8
  if (abs(chk$xy_mag_deg - job$tilt_mag) > tol ||
      (!is.na(chk$xy_dir_deg) &&
       abs(((chk$xy_dir_deg - tg_rot_xy_dir + 180) %% 360) - 180) > tol) ||
      abs(chk$z_deg - tg_rot_z) > tol)
    stop(sprintf(paste0("[T.3] The composer and the decomposer disagree on '%s'. One of them ",
                        "has been edited without the other; do not generate samples until ",
                        "they match."), job$sample_name), call. = FALSE)

  # ── T.4 Rotate about the shifted pivot ─────────────────────────────────────
  xyz_final <- sweep(t(R_fwd %*% t(sweep(xyz_shifted, 2, pivot_shifted))), 2, -pivot_shifted)

  # A rotation plus a translation cannot change any distance between two points.
  # .validate_rotation already ruled out a scale in the matrix; this catches an
  # error in how it was APPLIED - a transposed multiply, a pivot subtracted on
  # the wrong side. Sampled rather than exhaustive, which is enough: an
  # application error is systematic and shows up on any pair.
  set.seed(1L)
  .i  <- sample.int(nrow(xyz_base), min(2000L, nrow(xyz_base)))
  .j  <- sample.int(nrow(xyz_base), min(2000L, nrow(xyz_base)))
  rig <- max(abs(sqrt(rowSums((xyz_base[.i, ]  - xyz_base[.j, ])^2)) -
                 sqrt(rowSums((xyz_final[.i, ] - xyz_final[.j, ])^2))))
  if (rig > 1e-6)
    stop(sprintf("[T.4] The transform is not rigid on '%s' (max distance change %.3e m).",
                 job$sample_name, rig), call. = FALSE)
  .v(sprintf("\nRigidity check: max pairwise distance change %.3e m over %d sampled pairs\n",
             rig, length(.i)))

  disp <- sqrt(rowSums((xyz_final - xyz_base)^2))
  .v(sprintf("Net point displacement: min %.4f m | median %.4f m | max %.4f m\n",
             min(disp), median(disp), max(disp)))
  .v("  (the rotation contributes more the further a point sits from the pivot,\n")
  .v("   so the spread here is a lever-arm effect, not an inconsistency)\n")

  # ── T.4b The marker box ────────────────────────────────────────────────────
  # The SAME two operations, in the same order, about the same shifted pivot.
  # Written as one expression rather than by reusing xyz_shifted, so the cloud's
  # path and the box's path cannot drift apart if one of them is later edited.
  box_xyz_2 <- sweep(
    t(R_fwd %*% t(sweep(sweep(box_xyz_1, 2, tg_shift, `+`), 2, pivot_shifted))),
    2, -pivot_shifted)

  # Same rigidity argument on eight points instead of two thousand - cheap
  # enough here to be exhaustive: all 28 pairwise distances.
  erg <- max(abs(as.numeric(dist(box_xyz_1)) - as.numeric(dist(box_xyz_2))))
  if (erg > 1e-6)
    stop(sprintf("[T.4b] The marker box was not moved rigidly on '%s' (max edge change %.3e m).",
                 job$sample_name, erg), call. = FALSE)

  bd <- sqrt(rowSums((box_xyz_2 - box_xyz_1)^2))
  .v(sprintf("\nMarker box transformed | edge-length change %.3e m (must be ~0)\n", erg))
  .v(sprintf("Corner displacement: min %.4f m | mean %.4f m | max %.4f m\n",
             min(bd), mean(bd), max(bd)))
  .v("  (this is the displacement Stage 6 has to undo. Corners far from the pivot\n")
  .v("   move further - that lever arm is the whole point of putting them at the\n")
  .v("   extremes of the plot rather than near its centre)\n")
  if (isTRUE(verbose)) {
    cat(sprintf("\n  %-8s %12s %12s %12s %12s\n", "corner", "dx (m)", "dy (m)", "dz (m)", "|d| (m)"))
    for (i in box_corner_id)
      cat(sprintf("  %-8d %12.4f %12.4f %12.4f %12.4f\n", i,
                  box_xyz_2[i, 1] - box_xyz_1[i, 1],
                  box_xyz_2[i, 2] - box_xyz_1[i, 2],
                  box_xyz_2[i, 3] - box_xyz_1[i, 3], bd[i]))
  }

  # ── T.7 Invert, and check EVERYTHING before anything is written ────────────
  R_inv <- t(R_fwd)                    # for a rotation, the inverse is the transpose
  .validate_rotation(R_inv, "R_inv")
  inv <- .decompose_rotation(R_inv)

  # Straight negation - but only because the pivot was carried along in T.1.
  target_shift <- -tg_shift

  # Undoing "shift, then rotate about the moved pivot" means rotating back about
  # the moved pivot, then shifting back:
  #     x_base = (R_inv (x_sample - c') + c') - t
  T_inv <- diag(4)
  T_inv[1:3, 1:3] <- R_inv
  T_inv[1:3, 4]   <- as.numeric(pivot_shifted - R_inv %*% pivot_shifted - tg_shift)

  # The forward map in the same one-matrix form. Note the shift enters ROTATED -
  #     x_sample = R (x_base + t - c') + c'  =  R x_base + [R t + c' - R c']
  # so the translation column is R %*% t, not t. Only the INVERSE gets a plain
  # -t, because there the shift is undone after the rotation rather than before
  # it. Writing t here instead of R t is a millimetre-scale error at these
  # angles, which is exactly the size of error this document exists to measure.
  T_fwd <- diag(4)
  T_fwd[1:3, 1:3] <- R_fwd
  T_fwd[1:3, 4]   <- as.numeric(R_fwd %*% tg_shift + pivot_shifted - R_fwd %*% pivot_shifted)

  # Both matrices are checked against the coordinates actually produced, not
  # just against each other - a pair of matrices can be mutually consistent and
  # both wrong.
  fchk <- max(abs(t(T_fwd[1:3, 1:3] %*% t(xyz_base) + T_fwd[1:3, 4]) - xyz_final))
  if (fchk > 1e-6)
    stop(sprintf("[T.7] T_fwd does not reproduce the generated cloud for '%s' (max %.3e m).",
                 job$sample_name, fchk), call. = FALSE)

  # Push the generated sample back through the inverse and require the original
  # cloud. This is the one check that tests everything at once - composition
  # order, pivot handling, the sign of the shift, the transpose.
  rt <- max(abs(t(T_inv[1:3, 1:3] %*% t(xyz_final) + T_inv[1:3, 4]) - xyz_base))
  if (rt > 1e-6)
    stop(sprintf(paste0("[T.7] The inverse does not recover the reference for '%s' ",
                        "(max residual %.3e m). Nothing was written."),
                 job$sample_name, rt), call. = FALSE)

  # The same check on the marker box, which is what Stage 6 actually scores
  # against. If the inverse returns box 2 onto box 1 to machine precision here,
  # any gap Stage 6 reports is the pipeline's error, not a defect in how the box
  # was built or written.
  brt <- max(abs(t(T_inv[1:3, 1:3] %*% t(box_xyz_2) + T_inv[1:3, 4]) - box_xyz_1))
  if (brt > 1e-6)
    stop(sprintf(paste0("[T.7] The inverse does not return the marker box onto its truth ",
                        "copy for '%s' (max %.3e m). Nothing was written."),
                 job$sample_name, brt), call. = FALSE)

  .v(sprintf("\nForward 4x4 vs the applied transform       : max residual %.3e m\n", fchk))
  .v(sprintf("Round trip (sample -> inverse -> reference): max residual %.3e m\n", rt))
  .v(sprintf("Round trip (box 2 -> inverse -> box 1)     : max residual %.3e m\n", brt))

  if (isTRUE(verbose)) {
    cat("\n── Target values: what a perfect alignment run would report ──────────────\n")
    cat(sprintf("  XY rotation mag    : %8.4f deg\n", inv$xy_mag_deg))
    cat(sprintf("  XY rotation dir    : %8.4f deg\n", inv$xy_dir_deg))
    cat(sprintf("  Z rotation (yaw)   : %8.4f deg\n", inv$z_deg))
    cat(sprintf("  full XYZ magnitude : %8.4f deg\n", inv$total_deg))
    cat(sprintf("  X shift            : %+8.5f m\n", target_shift[1]))
    cat(sprintf("  Y shift            : %+8.5f m\n", target_shift[2]))
    cat(sprintf("  Z shift            : %+8.5f m\n", target_shift[3]))
    cat(sprintf("  XY shift magnitude : %8.5f m\n", sqrt(sum(target_shift[1:2]^2))))
    cat(sprintf("  XYZ shift magnitude: %8.5f m\n", sqrt(sum(target_shift^2))))
    cat(sprintf("  factorisation error: %.2e (angles rebuild R_inv exactly)\n", inv$rebuild_err))

    # Worth reading rather than skipping: the inverse tilt is the SAME magnitude
    # but points roughly the other way, and the yaw offsets it further. Anyone
    # expecting the target direction to be tg_rot_xy_dir with a minus sign will
    # mis-score every run.
    if (!is.na(inv$xy_dir_deg)) {
      cat(sprintf("\n  note: forward tilt dir %.4f deg -> target tilt dir %.4f deg (turned by %.4f deg,\n",
                  tg_rot_xy_dir %% 360, inv$xy_dir_deg,
                  (inv$xy_dir_deg - (tg_rot_xy_dir %% 360)) %% 360))
      cat("        which is ~180 deg from the forward direction, shifted by the yaw - not simply its negative)\n")
    }
  }

  # ── T.5 Assemble the cloud with its markers ────────────────────────────────
  dt <- data.table::as.data.table(las_ref@data)
  dt[, X := xyz_final[, 1]]
  dt[, Y := xyz_final[, 2]]
  dt[, Z := xyz_final[, 3]]

  # Every extra row is a COPY of an existing row with its coordinates
  # overwritten, not a row of NAs. A LAS point has required fields - Intensity,
  # ReturnNumber, Classification and so on - and a writer will refuse, or
  # silently write garbage, if they are missing. Copying a real point guarantees
  # every one of them is valid without this document having to know which fields
  # the file happens to carry.
  template <- data.table::copy(dt[1L])
  if ("TreeID" %in% names(template)) template[, TreeID := pivot_treeid_value]

  pivot_row <- data.table::copy(template)
  pivot_row[, `:=`(X = pivot_shifted[1], Y = pivot_shifted[2], Z = pivot_shifted[3])]

  box_rows <- function(M, box_value) {
    r <- data.table::copy(template[rep(1L, nrow(M))])
    r[, `:=`(X = M[, 1], Y = M[, 2], Z = M[, 3])]
    if ("TreeID" %in% names(r)) r[, TreeID := box_treeid_value]
    r[, (box_field_name)    := as.integer(box_value)]
    r[, (corner_field_name) := as.integer(box_corner_id)]
    r[, (pivot_field_name)  := 0L]
    r
  }

  # Defaults on the real points and on the pivot: not a marker, no corner. 0
  # rather than NA, because a LAS extra byte has no NA and a reader that meets
  # one will invent a value for it.
  dt[, (pivot_field_name)  := 0L]
  dt[, (box_field_name)    := 0L]
  dt[, (corner_field_name) := 0L]
  pivot_row[, (pivot_field_name)  := 1L]
  pivot_row[, (box_field_name)    := 0L]
  pivot_row[, (corner_field_name) := 0L]

  box_1_rows <- box_rows(box_xyz_1, 1L)
  box_2_rows <- box_rows(box_xyz_2, 2L)

  dt <- data.table::rbindlist(list(dt, pivot_row, box_1_rows, box_2_rows),
                              use.names = TRUE, fill = TRUE)

  las_out <- LAS(dt, header = las_ref@header)
  las_out <- lidR::las_update(las_out)   # refresh the header extent after moving every point

  # Register each flag as a LAS extra byte, so it survives the write and appears
  # as a named scalar field in CloudCompare rather than being dropped.
  reg <- function(las, nm, desc) {
    tryCatch(lidR::add_lasattribute(las, name = nm, desc = desc),
             error = function(e)
               lidR::add_lasattribute_manual(las, name = nm, desc = desc,
                                             type = "int", NA_value = NULL))
  }
  las_out <- reg(las_out, pivot_field_name,
                 "1 = rotation pivot used to generate this sample")
  las_out <- reg(las_out, box_field_name,
                 "1 = marker box at truth, 2 = marker box after the applied transform")
  las_out <- reg(las_out, corner_field_name,
                 "1-8 = which bounding-box vertex this marker is")

  .v(sprintf("\nCloud assembled: %s points (%s original + 1 pivot + %d marker)\n",
             format(nrow(las_out@data), big.mark = ","),
             format(nrow(xyz_final), big.mark = ","),
             nrow(box_1_rows) + nrow(box_2_rows)))

  # ── T.6 Write the .las ─────────────────────────────────────────────────────
  uniq <- function(p) if (isTRUE(overwrite_existing)) p else .unique_path(p)

  las_path <- uniq(file.path(las_output_dir, paste0(job$sample_name, ".las")))
  writeLAS(las_out, las_path)
  .v(sprintf("Saved %s points -> %s\n", format(nrow(las_out@data), big.mark = ","), las_path))

  # ── T.7b The forward and backward matrices, as matrices ────────────────────
  # T.8 also writes both, flattened into sixteen columns each so a batch can be
  # rbind'ed into one table. That layout is awkward for anything that just wants
  # "the matrix". These are the same two matrices again, each as its own
  # plain-text file: full double precision (17 significant digits, so the file
  # round-trips exactly), space-separated, no header - the same shape as the
  # TLS-registration matrix files elsewhere in the pipeline.
  #
  #   *_forward.txt  - T_fwd, reference -> sample. The damage applied. NOT what
  #                    the pipeline solves for.
  #   *_backward.txt - T_inv, sample -> reference. Score the pipeline's
  #                    recovered matrix against this one.
  fwd_path <- uniq(file.path(matrix_output_dir, paste0(job$sample_name, "_forward.txt")))
  bwd_path <- uniq(file.path(matrix_output_dir, paste0(job$sample_name, "_backward.txt")))
  .write_matrix_txt(T_fwd, fwd_path)
  .write_matrix_txt(T_inv, bwd_path)
  .v(sprintf("Saved forward  4x4 (reference -> sample) -> %s\n", fwd_path))
  .v(sprintf("Saved backward 4x4 (sample -> reference) -> %s\n", bwd_path))

  # ── T.8 The truth row ──────────────────────────────────────────────────────
  # ONE ROW, MANY COLUMNS. A long/tidy layout reads better on screen but this
  # exists to be stacked: one sample per row, joinable straight onto a table of
  # pipeline results. Matrices are flattened to R_fwd_11 ... R_fwd_33 in
  # row-major order so they read back with a single matrix(byrow = TRUE).
  truth <- c(
    list(
      # ── identity ────────────────────────────────────────────────────────
      sample_name        = job$sample_name,
      Sample_location_ID = job$Sample_location_ID,
      Cluster_ID         = job$Cluster_ID,
      Sample_Type_ID     = job$Sample_Type_ID,
      tilt_tag           = job$tilt_tag,
      tilt_tag_digits    = tilt_tag_digits,
      Epoch_1            = job$Epoch_1,
      Epoch_2            = job$Epoch_2,
      Scanner_1          = job$Scanner_1,
      Scanner_2          = job$Scanner_2,
      sample_prefix      = sample_prefix,
      sample_name_mode   = sample_name_mode,
      ref_index          = job$ref_index,
      tilt_index         = job$tilt_index,
      generated_utc      = format(Sys.time(), tz = "UTC", usetz = TRUE),
      # The R_ cloud this sample was damaged from. NOT the las_2 the pipeline
      # aligns against - that is a genuinely separate second epoch, and pairing
      # a sample against the cloud it was cut from would score the pipeline on
      # geometry it has already seen. Named reference_las_1 rather than
      # baseline_las_1 so the two cannot be confused at a glance in the CSV.
      reference_las_1    = job$ref_path,
      sample_las         = las_path,
      T_fwd_txt          = fwd_path,
      target_T_txt       = bwd_path,
      n_points_source    = nrow(xyz_final),
      n_points_written   = nrow(las_out@data),

      # ── marker box ──────────────────────────────────────────────────────
      box_source_cloud   = box_source_cloud,
      box_field_name     = box_field_name,
      corner_field_name  = corner_field_name,
      box_n_corners      = nrow(box_xyz_1),
      box_diagonal_m     = sqrt(sum((apply(box_xyz_1, 2, max) - apply(box_xyz_1, 2, min))^2)),
      box_corner_disp_max_m  = max(bd),
      box_corner_disp_mean_m = mean(bd),

      # ── pivot ───────────────────────────────────────────────────────────
      pivot_mode         = pivot_mode,
      pivot_base_x       = pivot_base[1],
      pivot_base_y       = pivot_base[2],
      pivot_base_z       = pivot_base[3],
      pivot_shifted_x    = pivot_shifted[1],
      pivot_shifted_y    = pivot_shifted[2],
      pivot_shifted_z    = pivot_shifted[3],

      # ── APPLIED: the damage done (NOT what the pipeline should report) ──
      applied_shift_x_m      = tg_shift[1],
      applied_shift_y_m      = tg_shift[2],
      applied_shift_z_m      = tg_shift[3],
      applied_shift_xy_m     = sqrt(sum(tg_shift[1:2]^2)),
      applied_shift_xyz_m    = sqrt(sum(tg_shift^2)),
      applied_rot_xy_mag_deg = job$tilt_mag,
      applied_rot_xy_dir_deg = tg_rot_xy_dir %% 360,
      applied_rot_z_deg      = tg_rot_z,
      applied_rot_total_deg  = acos(max(-1, min(1, (sum(diag(R_fwd)) - 1) / 2))) * 180 / pi,
      rotation_order         = "R = tilt(mag, dir) %*% yaw(z); shift applied before rotation",

      # ── TARGET: what a perfect run reports. SCORE AGAINST THESE. ────────
      target_shift_x_m      = target_shift[1],
      target_shift_y_m      = target_shift[2],
      target_shift_z_m      = target_shift[3],
      target_shift_xy_m     = sqrt(sum(target_shift[1:2]^2)),
      target_shift_xyz_m    = sqrt(sum(target_shift^2)),
      target_rot_xy_mag_deg = inv$xy_mag_deg,
      target_rot_xy_dir_deg = inv$xy_dir_deg,
      target_rot_z_deg      = inv$z_deg,
      target_rot_total_deg  = inv$total_deg,

      # ── provenance of the checks ────────────────────────────────────────
      factorisation_err         = inv$rebuild_err,
      check_fwd_max_m           = fchk,
      check_roundtrip_max_m     = rt,
      check_box_roundtrip_max_m = brt,
      check_rigidity_max_m      = rig
    ),
    .flat(R_fwd, "R_fwd"),
    .flat(R_inv, "target_R"),
    .flat(T_fwd, "T_fwd"),
    .flat(T_inv, "target_T"),
    # box1 = where the corners belong (the target), box2 = where they were moved to
    .flat_box(box_xyz_1, "box1"),
    .flat_box(box_xyz_2, "box2")
  )

  truth_df <- as.data.frame(truth, stringsAsFactors = FALSE)

  if (isTRUE(export_per_sample_csv)) {
    csv_path <- uniq(file.path(csv_output_dir, paste0(job$sample_name, ".csv")))
    truth_df$sample_csv <- csv_path
    utils::write.csv(truth_df, csv_path, row.names = FALSE)
    .v(sprintf("Saved %d truth column(s) -> %s\n", ncol(truth_df), csv_path))
  }

  list(truth    = truth_df,
       las_path = las_path,
       summary  = list(rt = rt, brt = brt, disp_max = max(disp), box_disp_max = max(bd)),
       coords   = if (isTRUE(keep_coords))
                    list(xyz_base = xyz_base, xyz_final = xyz_final,
                         box_xyz_1 = box_xyz_1, box_xyz_2 = box_xyz_2,
                         pivot_base = pivot_base, pivot_shifted = pivot_shifted,
                         rot_total = truth_df$applied_rot_total_deg)
                  else NULL)
}
```

## T.9 Run the Sweep

The outer loop is over reference files, the inner over tilts. That order is not
cosmetic: reading a cloud and computing its bounding box are the expensive parts
and they depend only on the reference, so each file is read once and its marker
box and pivot are built once, then reused across every tilt. Reading four files
twenty-eight times would be seven times the I/O for identical results - and,
worse, would rebuild the truth box each time, so the samples off one reference
would no longer all be scored against the same eight points.

A job that fails its checks is recorded and the sweep continues. The others are
independent of it, and nothing was written for it.

```{r}
#| label: run-sweep
#| code-summary: "Load each reference once, then generate every tilt from it"

for (.d in c(las_output_dir, csv_output_dir, matrix_output_dir))
  dir.create(.d, recursive = TRUE, showWarnings = FALSE)

results  <- vector("list", nrow(jobs))
failures <- list()
detail   <- NULL
t_start  <- Sys.time()

for (r in unique(jobs$ref_index)) {

  rows <- which(jobs$ref_index == r)
  cat(sprintf("\n══ Reference %d/%d: %s ══════════════════════════════════\n",
              r, length(unique(jobs$ref_index)), basename(jobs$ref_path[rows[1]])))

  las_ref <- readLAS(jobs$ref_path[rows[1]])
  if (is.empty(las_ref)) stop("Reference read as empty: ", jobs$ref_path[rows[1]])
  cat(sprintf("  loaded %s points\n", format(nrow(las_ref@data), big.mark = ",")))

  # ── The marker box, once per reference ─────────────────────────────────────
  bx <- range(las_ref@data$X); by <- range(las_ref@data$Y); bz <- range(las_ref@data$Z)

  # expand.grid varies its FIRST factor fastest, so the eight rows come out as
  #   1 (xmin,ymin,zmin) 2 (xmax,ymin,zmin) 3 (xmin,ymax,zmin) 4 (xmax,ymax,zmin)
  #   5 (xmin,ymin,zmax) 6 (xmax,ymin,zmax) 7 (xmin,ymax,zmax) 8 (xmax,ymax,zmax)
  # i.e. the bottom face first, then the top, each in the same x-then-y order.
  # Nothing downstream matches corners by proximity; corner 3 is compared with
  # corner 3 because both files agree on which vertex that is.
  box_xyz_1 <- as.matrix(expand.grid(X = bx, Y = by, Z = bz,
                                     KEEP.OUT.ATTRS = FALSE)[, c("X", "Y", "Z")])
  dimnames(box_xyz_1) <- NULL

  cat(sprintf("  marker box: %.1f x %.1f x %.1f m | diagonal %.2f m (the longest lever\n",
              diff(bx), diff(by), diff(bz),
              sqrt(diff(bx)^2 + diff(by)^2 + diff(bz)^2)))
  cat("              arm a rotation error has to act on)\n")

  # ── The pivot, once per reference ──────────────────────────────────────────
  pivot_base <- switch(pivot_mode,
                       auto = c(mean(bx), mean(by), mean(bz)),
                       custom = as.numeric(pivot_custom),
                       stop('pivot_mode must be "auto" or "custom".'))
  if (length(pivot_base) != 3L || anyNA(pivot_base))
    stop("The pivot did not resolve to three finite coordinates.")
  cat(sprintf("  pivot (%s): (%.3f, %.3f, %.3f)\n", pivot_mode,
              pivot_base[1], pivot_base[2], pivot_base[3]))

  if (isTRUE(export_reference_copy)) {
    ref_out <- file.path(las_output_dir, basename(jobs$ref_path[rows[1]]))
    if (!isTRUE(overwrite_existing)) ref_out <- .unique_path(ref_out)
    writeLAS(las_ref, ref_out)
    cat(sprintf("  reference copy -> %s\n", basename(ref_out)))
    rm(ref_out)
  }

  # ── The tilts ──────────────────────────────────────────────────────────────
  for (k in rows) {
    job     <- jobs[k, ]
    verbose <- identical(k, detail_job)

    if (verbose)
      cat(sprintf("\n── job %d/%d  %s  (full narration) ───────────────────\n",
                  k, nrow(jobs), job$sample_name))

    out <- tryCatch(
      .generate_sample(job, las_ref, box_xyz_1, pivot_base,
                       verbose = verbose, keep_coords = verbose),
      error = function(e) {
        # Recorded, not raised. The remaining jobs are independent of this one,
        # and nothing was written for it - every check runs before every write.
        failures[[length(failures) + 1L]] <<-
          data.frame(sample_name = job$sample_name, tilt_mag = job$tilt_mag,
                     error = conditionMessage(e), stringsAsFactors = FALSE)
        NULL
      })

    if (is.null(out)) {
      cat(sprintf("  [%2d/%d] %-46s FAILED - see the table below\n",
                  k, nrow(jobs), job$sample_name))
      next
    }

    results[[k]] <- out$truth
    if (verbose) detail <- out$coords

    cat(sprintf("  [%2d/%d] %-46s tilt %7.3f deg | max disp %7.3f m | round trip %.1e m\n",
                k, nrow(jobs), job$sample_name, job$tilt_mag,
                out$summary$disp_max, out$summary$rt))
  }

  rm(las_ref, box_xyz_1, pivot_base, bx, by, bz)
  invisible(gc(verbose = FALSE))
}

results <- Filter(Negate(is.null), results)
cat(sprintf("\n%d of %d job(s) generated in %.1f min\n",
            length(results), nrow(jobs),
            as.numeric(difftime(Sys.time(), t_start, units = "mins"))))
rm(.d)
```

## T.10 The Consolidated Reference CSV

One row per sample, every reference file and every tilt in the run, in a single
file. This is the one to hand the batch driver as `truth_csv_path`.

```{r}
#| label: export-consolidated
#| code-summary: "The stacked truth table, and a ready-made las_pairs for the driver"

if (!length(results))
  stop("No samples were generated - there is nothing to consolidate. See the failure table.")

# bind_rows rather than rbind: it fills a missing column with NA instead of
# refusing, which matters the day a column is added and a half-finished output
# directory holds rows from both versions.
truth_all <- dplyr::bind_rows(results)

# Written with .unique_path OFF regardless of overwrite_existing: a consolidated
# file that parked itself at _1 would leave the previous run's file sitting
# there under the name everything else points at, which is the one failure mode
# worth ruling out by construction.
utils::write.csv(truth_all, consolidated_truth_csv, row.names = FALSE)
cat(sprintf("Reference CSV: %d row(s) x %d column(s) -> %s\n",
            nrow(truth_all), ncol(truth_all), consolidated_truth_csv))

# No separate las_pairs file. Everything the driver needs to build one is
# already in the table above - sample_las is las_1, and the ID fields name
# las_2 - so a second file would be a copy that can fall out of step with it.

# Every sample_name must appear exactly once, because that column is what every
# downstream join keys on. The job table was already checked for duplicates, so
# reaching here means something changed between then and now.
if (anyDuplicated(truth_all$sample_name))
  stop("The consolidated table has duplicate sample_name values: ",
       paste(unique(truth_all$sample_name[duplicated(truth_all$sample_name)]), collapse = ", "))

cat(sprintf("  covering %d reference file(s) x %d tilt(s), %d sample(s) in total\n",
            length(unique(truth_all$reference_las_1)),
            length(unique(truth_all$applied_rot_xy_mag_deg)),
            nrow(truth_all)))

if (length(failures)) {
  fail_df <- dplyr::bind_rows(failures)
  cat(sprintf("\n%d job(s) FAILED and were not written:\n", nrow(fail_df)))
  for (i in seq_len(nrow(fail_df)))
    cat(sprintf("  %-46s %s\n", fail_df$sample_name[i], fail_df$error[i]))
} else {
  cat("\nNo failures.\n")
}

# ── What the sweep looks like ────────────────────────────────────────────────
# The shift is constant, so target_shift_* is identical down every row - that is
# the control. What varies is the tilt, and with it how far the marker box
# corners had to travel. A corner displacement that does not grow with the tilt
# would mean the tilt is not reaching the geometry.
cat("\n── The sweep ─────────────────────────────────────────────────────────────\n")
print(truth_all[, c("sample_name", "applied_rot_xy_mag_deg", "applied_rot_total_deg",
                    "target_rot_xy_mag_deg", "box_corner_disp_max_m",
                    "check_roundtrip_max_m")],
      row.names = FALSE, digits = 5)
```

### Vis - reference, generated sample, and the markers

Drawn for `detail_job` only. Twenty-eight interactive 3D plots in one HTML
document is a file nobody can open.

```{r}
#| label: fig-generated-sample
#| fig-cap: "The reference cloud and one generated sample, with the rotation pivot and both marker boxes."

if (params$make_plots && !is.null(detail)) {

  vis_max_pts <- 20000L
  job_d <- jobs[detail_job, ]

  .thin <- function(m, lab) {
    n <- nrow(m)
    i <- if (n > vis_max_pts) sample.int(n, vis_max_pts) else seq_len(n)
    data.frame(X = m[i, 1], Y = m[i, 2], Z = m[i, 3], cloud = lab)
  }

  set.seed(7L)
  .df <- rbind(
    .thin(detail$xyz_base,  "reference (aligned)"),
    .thin(detail$xyz_final, "generated sample (misaligned)")
  )

  # Plotted about the pivot, so the axes read as metres of displacement rather
  # than as UTM coordinates where a 20 cm shift is invisible in the last digits.
  .df$Xl <- .df$X - detail$pivot_base[1]
  .df$Yl <- .df$Y - detail$pivot_base[2]

  plot_ly() |>
    add_markers(data = .df, x = ~Xl, y = ~Yl, z = ~Z, color = ~cloud,
                colors = c("reference (aligned)"           = COL_1,
                           "generated sample (misaligned)" = COL_P),
                marker = list(size = 1.5, opacity = 0.5),
                hoverinfo = "text", text = ~cloud) |>
    add_markers(x = detail$pivot_shifted[1] - detail$pivot_base[1],
                y = detail$pivot_shifted[2] - detail$pivot_base[2],
                z = detail$pivot_shifted[3],
                marker = list(size = 8, color = "black", symbol = "x"),
                name = "pivot (shifted)") |>
    add_markers(x = detail$box_xyz_1[, 1] - detail$pivot_base[1],
                y = detail$box_xyz_1[, 2] - detail$pivot_base[2],
                z = detail$box_xyz_1[, 3],
                marker = list(size = 6, color = COL_2, symbol = "diamond"),
                text = paste0("TargetBox 1 (truth), corner ", 1:8),
                hoverinfo = "text", name = "marker box 1 (truth)") |>
    add_markers(x = detail$box_xyz_2[, 1] - detail$pivot_base[1],
                y = detail$box_xyz_2[, 2] - detail$pivot_base[2],
                z = detail$box_xyz_2[, 3],
                marker = list(size = 6, color = COL_1, symbol = "diamond-open"),
                text = paste0("TargetBox 2 (moved), corner ", 1:8),
                hoverinfo = "text", name = "marker box 2 (moved)") |>
    layout(title = sprintf("%s - shift %.3f m, tilt %.3f deg, full rotation %.4f deg",
                           job_d$sample_name,
                           sqrt(tg_shift_x^2 + tg_shift_y^2 + tg_shift_z^2),
                           job_d$tilt_mag, detail$rot_total),
           scene = list(xaxis = list(title = "X - pivot (m)"),
                        yaxis = list(title = "Y - pivot (m)"),
                        zaxis = list(title = "Z (m)"),
                        aspectmode = "data"))
}
```

Generation complete

Each `.las` in `las_output_dir` is now an ordinary-looking unaligned sample.
Point the batch driver's `truth_csv_path` at the consolidated CSV and list the
sample names you want in `sample_names` - the driver resolves `las_1_path` and
`las_2_path` from that table itself - then run the pipeline through
to Stage 6 exactly as normal - nothing downstream is told these are tests. Stage
6 Step 6.6 finds each sample's row by `sample_name`, applies the transform it
solved to the `TargetBox = 2` corners, and reports the gap to their
`TargetBox = 1` twins as the primary shift residual.

Because the shift and the yaw are held constant across the sweep and only the
tilt varies, a plot of Stage 6's `box_res_xyz_m` against
`applied_rot_xy_mag_deg` reads directly as how far the tilt can be pushed before
the pipeline stops recovering it.

Stage 1 from `Stage1_2026-08-19_1.qmd` onward drops the sixteen `TargetBox`
points at load, so the marker box no longer reaches `pivot1_auto` or the Stage 1A
sideways-forest test. On an older Stage 1, use `pivot1_mode = "las"` and be aware
that the box-2 corners - which at a 180 degree tilt sit a long way outside the
forest - are still inflating the cloud extent.
